---
title: Ultimate Environment Construction Guide
subtitle: Complete setup for comparative genomics analysis
jupyter: python3
---

## Info Summary

This is the comprehensive environment setup guide for the comparative genomics analysis project. It provides step-by-step instructions for setting up all required bioinformatics tools and dependencies.

### Project Information
- **Docker Image**: `huang/comparative-genomics:ultimate` (after complete setup)
- **Jupyter Access**: `http://localhost:38888/?token=[your-token]`

### Available Conda Environments
After completing this setup, you will have the following environments:
- `base` - Base conda environment
- `env_biopython` - Python environment with BioPython and analysis libraries
- `env_orthofinder` - OrthoFinder for orthology analysis
- `bioconductor-clusterprofiler_env` - R environment for functional enrichment analysis
- `eggnog_env` - EggNOG mapper for functional annotation
- `pastml_env` - PastML for ancestral state reconstruction
- `gff2bed_env` - BEDOPS utilities for format conversion
- `genome_annotation_env` - RepeatModeler, RepeatMasker, gffread, BUSCO for genome annotation
- `braker3_env` - BRAKER3 and AUGUSTUS for automated genome annotation

### Available Jupyter Kernels
- `env_biopython` - Python (Biopython) kernel
- `ir` - Base R kernel
- `julia-1.7` - Julia kernel
- `python3` - Base Python kernel
- `ir_clusterprofiler_481_env` - clusterprofiler R kernel

## Environment List Check

First, let's check the current environment setup:

```{python}
!hostname
!mamba env list
!jupyter kernelspec list
```

```{r}
R.home()
```

## 0. Install basic modules
PyYAML is required for configuration loading. Install with:

```{python}
!pip install pyyaml
```

## 1. Install OrthoFinder (3.0.1b1)

OrthoFinder is used for phylogenetic orthology inference and comparative genomics analysis.

```{python}
!mamba create -n env_orthofinder -y bioconda::orthofinder
```

Verify the installation:

```{python}
!conda run -n env_orthofinder orthofinder -h
```

## 2. Install Biopython Environment

Set up the main Python environment for bioinformatics analysis.

### 2.1 Create Base Environment (1.85)

```{python}
%%bash
mamba create -n env_biopython -y conda-forge::biopython pip ipykernel
conda run -n env_biopython python -m ipykernel install --user --name env_biopython --display-name "Python (Biopython)"
```

### 2.2 Install GFF Utilities (0.13)

```{python}
!mamba install -n env_biopython -y bioconda::gffutils
```

### 2.3 Install Data Analysis Libraries

Install essential Python libraries for data manipulation and visualization:

```{python}
!mamba install -y -n env_biopython matplotlib
```

> 3.10.0

```{python}
!mamba install -y -n env_biopython pandas
```

> 2.2.3

```{python}
!mamba install -y -n env_biopython seaborn
```

> 0.13.2

```{python}
!mamba install -y -n env_biopython scikit-learn
```

### 2.4 Install Specialized Plotting Libraries (0.9.0)

```{python}
!mamba install -y -n env_biopython conda-forge::upsetplot
```

## 3. Install clusterProfiler (R Environment)

clusterProfiler is used for functional enrichment analysis of gene clusters.

### 3.1 Search Available Versions

```{python}
!mamba search bioconda::bioconductor-clusterprofiler 
```

### 3.2 Create clusterProfiler Environment

```{python}
!mamba create --name bioconductor-clusterprofiler_env -y bioconda::bioconductor-clusterprofiler=4.8.1
```

### 3.3 Install R Jupyter Kernel (1.3.2)

```{python}
!/opt/conda/envs/bioconductor-clusterprofiler_env/bin/Rscript -e "install.packages('IRkernel', repos = 'https://ftp.yz.yamagata-u.ac.jp/pub/cran/')"
```

```{python}
%%bash
/opt/conda/envs/bioconductor-clusterprofiler_env/bin/Rscript - <<EOF
library(IRkernel)
packageVersion("IRkernel")
EOF
```

### 3.4 Register Kernel and Verify

```{python}
!/opt/conda/envs/bioconductor-clusterprofiler_env/bin/Rscript -e "IRkernel::installspec(name = 'ir_clusterprofiler_481_env', displayname = 'clusterprofiler R', user = FALSE)"
!jupyter kernelspec list
```

### 3.5 Install tools for plotting
```{r}
install.packages("phytools",repos="https://cloud.r-project.org",quiet=TRUE)
install.packages('deeptime',repos="https://cloud.r-project.org",quiet=TRUE)
```

> also installing the dependencies 'iterators', 'quadprog', 'maps', 'clusterGeneration', 'coda', 'combinat', 'DEoptim', 'doParallel', 'expm', 'foreach', 'mnormt', 'numDeriv', 'optimParallel', 'phangorn', 'scatterplot3d'

### 3.6 Install yaml package for configuration loading

The yaml package is required for loading the centralized path configuration from `paths.yaml`:

```{r}
install.packages("yaml", repos="https://cloud.r-project.org", quiet=TRUE)
```

## 4. Install EggNOG Mapper

EggNOG mapper provides functional annotation of genes and proteins.

### 4.1 Create Environment (2.1.12)

```{python}
!mamba create -n eggnog_env -y -c bioconda -c conda-forge eggnog-mapper
```

### 4.2 Download EggNOG Database (5.0.2)

```{python}
!mkdir -p "./DB/Eggnog"
%conda run -n eggnog_env download_eggnog_data.py -y --data_dir "./DB/Eggnog"
```

## 5. Install PastML

PastML is used for ancestral state reconstruction on phylogenetic trees.

### 5.1 Create Python Environment

```{python}
print("Installing PastML...")
!mamba create -n pastml_env -y python=3.10
```

```{python}
!conda run -n pastml_env python3 --version
```

### 5.2 Install PastML via pip (1.9.50)

```{python}
!conda run -n pastml_env python3 -m pip install pastml
```

### 5.3 Verify Installation

```{python}
!conda run -n pastml_env which pastml
```

### 5.4 Clone PastML Repository for Examples

```{python}
!ls Apps
```

```{python}
!git clone https://github.com/evolbioinfo/pastml.git Apps/PastML
```

### 5.5 Test PastML with Example Data

```{python}
!conda run -n pastml_env pastml --tree ./Apps/PastML/examples/Albania/data/Albanian.tree.152tax.tre --data ./Apps/PastML/examples/Albania/data/data.txt --columns Country --html_compressed ./Test/Albanian_map.html --data_sep , 
```

## 6. Install MCScanX

MCScanX is used for synteny and collinearity analysis.

### 6.1 Install Java Dependency

```{python}
!mamba install -y conda-forge::openjdk
```

### 6.2 Download MCScanX Source

```{bash}
%%bash
ls ./Apps
mkdir -p ./Apps/MCScanX
wget https://github.com/wyp1125/MCScanX/archive/refs/tags/v1.0.0.tar.gz -O ./Apps/MCScanX/v1.0.0.tar.gz
```

### 6.3 Compile MCScanX

```{bash}
%%bash
mkdir ./Apps/MCScanX/V1
tar zxvf ./Apps/MCScanX/v1.0.0.tar.gz -C ./Apps/MCScanX/V1
cd ./Apps/MCScanX/V1/MCScanX-1.0.0; make
```

### 6.4 Verify Installation

```{bash}
%%bash
ls -lth ./Apps/MCScanX/V1/MCScanX-1.0.0
ls -lth ./Apps/MCScanX/V1/MCScanX-1.0.0/downstream_analyses
```

### 6.5 Test MCScanX Tools

```{python}
!./Apps/MCScanX/V1/MCScanX-1.0.0/duplicate_gene_classifier -h
```

## 7. Install BLAST+

BLAST+ is used for sequence similarity searches.

### 7.1 Download and Install BLAST+

```{python}
!mkdir -p ./Apps/Blast
!wget https://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/2.16.0/ncbi-blast-2.16.0+-x64-linux.tar.gz -O ./Apps/Blast/blast-2.16.0.tar.gz
!tar zxvf ./Apps/Blast/blast-2.16.0.tar.gz -C ./Apps/Blast
!ls ./Apps/Blast/ncbi-blast-2.16.0+/bin/
```

## 8. Install BEDOPS (gff2bed)

BEDOPS provides utilities for genomic interval operations and format conversion.

### 8.1 Create BEDOPS Environment

```{python}
!mamba create -n gff2bed_env -y bioconda::bedops
```

### 8.2 Verify Installation

```{python}
!conda run -n gff2bed_env which gff2bed
```

## 9. Install Genome Annotation Tools

This section installs essential tools for genome annotation workflows including repeat identification, masking, and quality assessment.

### 9.1 Create Genome Annotation Environment

Create a dedicated environment with RepeatModeler, RepeatMasker, gffread, and BUSCO:

```{python}
%%bash
mamba create -n genome_annotation_env -y -c conda-forge -c bioconda \
    repeatmodeler repeatmasker gffread busco=6.0.0 sepp=4.5.5
```

> **Note**: BUSCO 6.0.0 should be installed with SEPP 4.5.5 for auto-lineage compatibility. This installation may take several minutes due to the many dependencies these tools require.

### 9.2 Verify RepeatModeler Installation

RepeatModeler is used for de novo repeat family identification and modeling.

```{python}
!conda run -n genome_annotation_env RepeatModeler -version
```

### 9.3 Verify RepeatMasker Installation

RepeatMasker screens DNA sequences for interspersed repeats and low complexity regions.

```{python}
!conda run -n genome_annotation_env RepeatMasker -version
```

### 9.4 Verify gffread Installation

gffread is a utility for parsing and converting GFF/GTF files.

```{python}
!conda run -n genome_annotation_env gffread --version
```

### 9.5 Verify BUSCO Installation

BUSCO (Benchmarking Universal Single-Copy Orthologs) assesses genome assembly and annotation completeness.

```{python}
!conda run -n genome_annotation_env busco --version
```

### 9.6 Troubleshooting Notes

If you encounter issues during installation:

- If mamba/conda hangs during environment solving, try creating a fresh environment rather than installing into an existing one
- For RepeatModeler, version 2.0.x includes the LTR structural discovery pipeline (`-LTRStruct` option)
- If you get perl-related errors with RepeatModeler, install `perl-File-Which`:

```{python}
# Only run if needed for perl-related errors
# !mamba install -n genome_annotation_env -y perl-file-which
```

## 10. Install BRAKER3

BRAKER3 is a fully automated pipeline for eukaryotic genome annotation that combines GeneMark-ETP with AUGUSTUS. It uses protein and/or RNA-Seq evidence for training gene finders.

> **Note**: The bioconda BRAKER package may lag behind the GitHub development version, but is sufficient for standard annotation workflows. For bleeding-edge features, consider manual installation from GitHub.

### 10.1 Create BRAKER Environment

Create a dedicated environment for BRAKER3 with all dependencies:

```{python}
%%bash
mamba create -n braker3_env -y -c bioconda -c conda-forge \
    braker3 augustus
```

> **Note**: This installation may take 10-20 minutes due to the many Perl dependencies and compiled tools required.

### 10.2 Verify BRAKER Installation

```{python}
!conda run -n braker3_env braker.pl --version
```

### 10.3 Verify AUGUSTUS Installation

AUGUSTUS is the core gene prediction tool used by BRAKER:

```{python}
!conda run -n braker3_env augustus --version
```

### 10.4 Check AUGUSTUS Config Path

BRAKER needs a writable AUGUSTUS config directory:

```{python}
!conda run -n braker3_env bash -c 'echo $AUGUSTUS_CONFIG_PATH'
```

### 10.5 Download GeneMark-ETP (Manual Step Required)

GeneMark-ETP requires a license agreement and must be downloaded manually:

1. Visit: http://exon.gatech.edu/GeneMark/license_download.cgi
2. Select "GeneMark-ETP" and agree to the license
3. Download the appropriate version for Linux 64-bit
4. Extract to `./Apps/GeneMark-ETP/`

```{python}
# After downloading, extract GeneMark-ETP:
# !mkdir -p ./Apps/GeneMark-ETP
# !tar -xzf gmetp_linux_64.tar.gz -C ./Apps/GeneMark-ETP
# !gzip -d ./Apps/gm_key.gz
# !cp ./Apps/gm_key ~/.gm_key  # Copy license key
```

### 10.6 Set GeneMark Path (After Manual Download)

```{python}
# After GeneMark-ETP is installed, set the path:
# import os
# os.environ['GENEMARK_PATH'] = './Apps/GeneMark-ETP/gmes_linux_64_4'
```

### 10.7 Test BRAKER (Quick Validation)

Test that BRAKER can find its dependencies:

```{python}
!conda run -n braker3_env braker.pl --help | head -50
```

### 10.8 BRAKER Usage Example

Example command for protein-only mode (EP mode / BRAKER2-style):

```bash
# Run from the project directory with braker3_env activated
conda run -n braker3_env braker.pl \
    --genome=./path/to/genome.fna \
    --prot_seq=./DB/Metazoa.fa \
    --species=species_name \
    --threads=16 \
    --gff3 \
    --workingdir=./Outputs/BRAKER/species_name
```

### 10.9 Troubleshooting BRAKER

Common issues and solutions:

- **"AUGUSTUS_CONFIG_PATH not writable"**: Copy the config directory to a writable location:
  ```bash
  cp -r $CONDA_PREFIX/config ./Augustus_config
  export AUGUSTUS_CONFIG_PATH=$(pwd)/Augustus_config
  ```

- **"GeneMark not found"**: Ensure GeneMark-ETP is downloaded and `GENEMARK_PATH` is set

- **Perl module errors**: Install missing modules:
  ```bash
  mamba install -n braker3_env -y perl-yaml perl-hash-merge perl-parallel-forkmanager
  ```

## 11. Install State-of-the-Art Visualization Tools

This section installs modern visualization tools identified in the bioinformatics literature (2023-2025) for publication-quality figures. These tools provide significant improvements over legacy approaches.

### 11.1 Python Visualization Tools (env_biopython)

#### ETE Toolkit v4 - Advanced Tree Visualization

ETE4 provides the `ete-evol` module for CodeML integration and a web-based "smartview" that eliminates Qt/Xvfb dependencies for HPC headless rendering.

```{python}
!conda run -n env_biopython pip install ete4
```

#### pyGenomeViz - Modern Synteny Visualization

pyGenomeViz (v1.6.1+) provides CLI workflows (`pgv-blast`, `pgv-mummer`) that automate alignment-to-visualization pipelines, replacing legacy Java-based MCScanX visualization tools.

```{python}
!conda run -n env_biopython pip install pygenomeviz
```

#### pyCirclize - Circos-style Circular Plots

pyCirclize enables Circos-style circular genome visualizations with chord diagrams for synteny blocks and genomic rearrangements.

```{python}
!conda run -n env_biopython pip install pycirclize
```

### 11.2 Verify Python Visualization Tools

```{python}
%%bash
conda run -n env_biopython python -c "
import ete4; print(f'ETE4: {ete4.__version__}')
import pygenomeviz; print(f'pyGenomeViz: {pygenomeviz.__version__}')
import pycirclize; print(f'pyCirclize: {pycirclize.__version__}')
"
```

### 11.3 R Visualization Tools (bioconductor-clusterprofiler_env)
```{python}
!mamba install -n bioconductor-clusterprofiler_env -y conda-forge::mpfr conda-forge::gmp
```

#### ggstatsplot - Statistical Annotations

ggstatsplot provides APA-formatted statistical annotations for scatter plots, including correlation coefficients, p-values, and confidence intervals.

```{r}
install.packages("Rmpfr", repos="https://cloud.r-project.org")
install.packages("ggstatsplot", repos="https://cloud.r-project.org")
```

#### ggpubr - Publication-Ready Plots

ggpubr provides `stat_cor()` and `stat_compare_means()` for easy statistical annotation, plus `ggarrange()` for multi-panel figures.

```{r}
install.packages("ggpubr", repos="https://cloud.r-project.org", quiet=TRUE)
```

#### tidygraph + ggraph - Network Visualization

These packages provide grammar-of-graphics network visualization for GO term relationships and orthogroup networks.

```{r}
install.packages(c("tidygraph", "ggraph"), repos="https://cloud.r-project.org", quiet=TRUE)
```

### 11.4 Verify R Visualization Tools

```{r}
cat("Checking R visualization packages...\n")
packages <- c("ggstatsplot", "ggpubr", "tidygraph", "ggraph")
for (pkg in packages) {
    if (requireNamespace(pkg, quietly = TRUE)) {
        cat(paste0(pkg, ": ", packageVersion(pkg), "\n"))
    } else {
        cat(paste0(pkg, ": NOT INSTALLED\n"))
    }
}
```

### 11.5 State-of-the-Art Tools Reference

| Tool | Language | Purpose | Key Feature |
|------|----------|---------|-------------|
| ETE4 | Python | Tree visualization | smartview (no Xvfb), `ete-evol` for CodeML |
| pyGenomeViz | Python | Synteny plots | CLI workflows, modern design |
| pyCirclize | Python | Circular plots | Circos-style, chord diagrams |
| ggstatsplot | R | Statistical plots | APA-formatted annotations |
| ggpubr | R | Publication plots | Easy stat annotation |
| ggraph | R | Networks | Grammar-of-graphics networks |

> **Reference**: See `References/state-of-the-art-tools.md` for detailed tool evaluation and literature citations.

## Installation Complete

Your comparative genomics analysis environment is now fully configured with all necessary tools:

### Core Analysis Tools
- **OrthoFinder**: Orthology inference and comparative genomics
- **MCScanX**: Synteny and collinearity analysis  
- **BLAST+**: Sequence similarity searches
- **PastML**: Ancestral state reconstruction

### Functional Annotation
- **EggNOG Mapper**: Gene/protein functional annotation
- **clusterProfiler**: Functional enrichment analysis (R)

### Genome Annotation & Quality Assessment
- **BRAKER3**: Fully automated eukaryotic genome annotation pipeline
- **AUGUSTUS**: Ab initio gene prediction with hints
- **RepeatModeler**: De novo repeat family identification and modeling
- **RepeatMasker**: Repeat element screening and masking
- **gffread**: GFF/GTF file parsing and conversion
- **BUSCO**: Genome assembly and annotation completeness assessment

### Data Processing & Visualization
- **BioPython**: Python bioinformatics libraries
- **Pandas, NumPy**: Data manipulation
- **Matplotlib, Seaborn**: Visualization
- **UpSetPlot**: Set intersection visualization
- **BEDOPS**: Genomic interval operations

### State-of-the-Art Visualization (2023-2025)
- **ETE4**: Advanced tree visualization with CodeML integration
- **pyGenomeViz**: Modern synteny visualization (replaces MCScanX Java tools)
- **pyCirclize**: Circos-style circular genome plots
- **ggstatsplot**: APA-formatted statistical annotations (R)
- **ggraph**: Grammar-of-graphics network visualization (R)

### Development Environment
- **Jupyter Lab/Notebook**: Interactive analysis
- **Multiple kernels**: Python, R, Julia support
- **Conda environments**: Isolated tool dependencies

You can now proceed with your comparative genomics analysis workflows using these properly configured environments.
