---
title: Ultimate OrthoFinder Visualization Analysis
subtitle: Comprehensive visualization suite for comparative genomics results
jupyter:
  kernelspec:
    display_name: Python (Biopython)
    language: python
    name: env_biopython
---

## Project Overview

This notebook provides a comprehensive visualization suite for OrthoFinder comparative genomics results, including phylogenetic trees, orthogroup analysis, heatmaps, statistical plots, and advanced set visualizations. The analysis integrates multiple visualization approaches developed across different analysis phases.

### Data Sources
- **OrthoFinder Results**: Configured in `config/paths.yaml`
- **Processed Tables**: `./Notebooks/Tables/`
- **Output Figures**: `./Notebooks/Figures/`

## Environment Setup (env_biopython)

```{python}
!which python3
!mamba env list
!jupyter kernelspec list
from IPython.display import Image, display
```

### Load Configuration

```{python}
import sys
sys.path.insert(0, './Notebooks/config')
from load_config import load_paths, get_path, get_version, get_focal_species, get_focal_abbreviation

# Load and display configuration
config = load_paths(config_path="/home/jovyan/Notebooks/config/paths-set2.yaml")
print(f"Configuration loaded: OrthoFinder Results_{get_version()}")

# Define commonly used paths for this notebook
ORTHOFINDER_RESULTS = get_path('orthofinder.results_dir')
SPECIES_TREE = get_path('orthofinder.species_tree')
SPECIES_TREE_NODE_LABELS = get_path('orthofinder.species_tree_node_labels')
GENE_DUPLICATIONS = get_path('orthofinder.duplications_file')
ORTHOGROUPS_COUNT = get_path('orthofinder.orthogroups_count')
GENE_TREES_DIR = get_path('orthofinder.gene_trees')
OUTPUT_FIGURES = get_path('outputs.figures')
OUTPUT_TABLES = get_path('outputs.tables')
FOCAL_SPECIES = get_focal_species()

print(f"OrthoFinder results: {ORTHOFINDER_RESULTS}")
print(f"Species tree: {SPECIES_TREE}")
```

### Available Kernels
```txt
Available kernels:
  env_biopython                  /home/jovyan/.local/share/jupyter/kernels/env_biopython
  ir                             /opt/conda/share/jupyter/kernels/ir
  julia-1.7                      /opt/conda/share/jupyter/kernels/julia-1.7
  python3                        /opt/conda/share/jupyter/kernels/python3
  ir_clusterprofiler_4100_env    /usr/local/share/jupyter/kernels/ir_clusterprofiler_4100_env
  ir_clusterprofiler_481_env     /usr/local/share/jupyter/kernels/ir_clusterprofiler_481_env
```

## 1. Species Tree Visualization

### 1.1 Species Tree Plotting Function

```{python}
from Bio import Phylo
import matplotlib.pyplot as plt
from io import StringIO

def plot_phylogenetic_tree_with_lengths(
    newick_file_path, 
    figsize=(14, 10), 
    branch_label_precision=3,
    title=None,
    title_fontsize=22,
    label_fontsize=18,
    branch_label_fontsize=14,
    branch_color='#2E4057',
    label_color='#1A1A1A',
    background_color='white',
    show_branch_labels=True,
    show_scale_bar=True,
    orientation='horizontal',
    save_path=None,
    dpi=300,
    publication_ready=True
):
    """
    Creates a publication-quality phylogenetic tree plot with enhanced styling and customization options.

    Parameters:
    - newick_file_path: str, path to the Newick file containing the tree
    - figsize: tuple(int, int), figure size in inches (default: (14, 10))
    - branch_label_precision: int, decimal places for branch lengths (default: 3)
    - title: str, optional, plot title (auto-generated if None)
    - title_fontsize: int, title font size (default: 16)
    - label_fontsize: int, species label font size (default: 12)
    - branch_label_fontsize: int, branch length label font size (default: 9)
    - branch_color: str, color for branches (default: '#2E4057')
    - label_color: str, color for labels (default: '#1A1A1A')
    - background_color: str, background color (default: 'white')
    - show_branch_labels: bool, whether to show branch length labels (default: True)
    - show_scale_bar: bool, whether to show scale bar (default: True)
    - orientation: str, 'horizontal' or 'vertical' tree orientation (default: 'horizontal')
    - save_path: str, optional, base path to save the figure (saves as PNG and PDF)
    - dpi: int, resolution for saved figures (default: 300)
    - publication_ready: bool, apply publication-ready styling (default: True)

    Returns:
    - matplotlib.figure.Figure: the created figure object
    """
    import matplotlib.patches as mpatches
    from matplotlib.patches import Rectangle
    import numpy as np
    
    # Load the tree from the Newick file
    try:
        tree = Phylo.read(newick_file_path, "newick")
    except Exception as e:
        raise ValueError(f"Error reading tree file {newick_file_path}: {str(e)}")
    
    # Set publication-ready style
    if publication_ready:
        plt.style.use('default')
        plt.rcParams.update({
            'font.family': 'serif',
            'font.serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
            'font.size': label_fontsize,
            'axes.linewidth': 1.2,
            'axes.spines.top': False,
            'axes.spines.right': False,
            'axes.spines.bottom': True,
            'axes.spines.left': True,
            'xtick.major.width': 1.2,
            'ytick.major.width': 1.2,
            'figure.facecolor': background_color,
            'axes.facecolor': background_color
        })
    
    # Create figure and axis
    fig, ax = plt.subplots(1, 1, figsize=figsize, facecolor=background_color)
    ax.set_facecolor(background_color)
    
    # Configure branch labels
    branch_label_func = None
    if show_branch_labels:
        branch_label_func = lambda c: f'{c.branch_length:.{branch_label_precision}f}' if c.branch_length else ''
    
    # Draw the tree
    Phylo.draw(
        tree, 
        axes=ax, 
        do_show=False,
        branch_labels=branch_label_func,
        label_func=lambda x: x.name if x.name else ''
    )
    
    # Customize appearance
    for line in ax.get_lines():
        line.set_color(branch_color)
        line.set_linewidth(2.0)
    
    # Customize label and branch label colors and fonts
    for text in ax.texts:
        # Check if it's a branch label (contains numbers/decimals)
        text_content = text.get_text().strip()
        if text_content and any(c.isdigit() or c == '.' for c in text_content):
            # This is likely a branch length label
            text.set_color('#666666')
            text.set_fontsize(branch_label_fontsize)
        else:
            # This is likely a species/node label
            text.set_color(label_color)
            text.set_fontsize(label_fontsize)
    
    # Set axis properties
    ax.spines['top'].set_visible(False)
    ax.spines['right'].set_visible(False)
    ax.spines['bottom'].set_visible(True)
    ax.spines['left'].set_visible(False)
    
    # Handle orientation
    if orientation == 'vertical':
        ax.invert_yaxis()
        plt.setp(ax.get_xticklabels(), rotation=90, ha='center')
    
    # Add scale bar
    if show_scale_bar:
        # Calculate appropriate scale bar length
        xlim = ax.get_xlim()
        x_range = xlim[1] - xlim[0]
        scale_length = round(x_range * 0.1, branch_label_precision)
        
        # Position scale bar
        scale_x = xlim[0] + x_range * 0.05
        ylim = ax.get_ylim()
        scale_y = ylim[0] + (ylim[1] - ylim[0]) * 0.05
        
        # Draw scale bar
        ax.plot([scale_x, scale_x + scale_length], [scale_y, scale_y], 
                color=branch_color, linewidth=3)
        ax.text(scale_x + scale_length/2, scale_y - (ylim[1] - ylim[0]) * 0.02, 
                f'{scale_length}', ha='center', va='top', 
                fontsize=branch_label_fontsize, color=label_color)
    
    # Set title
    if title is None:
        # Auto-generate title based on file name
        import os
        filename = os.path.basename(newick_file_path)
        if 'rooted' in filename.lower():
            title = 'Species Phylogenetic Tree (Rooted)'
        elif 'gene' in filename.lower() and 'duplication' in filename.lower():
            title = 'Species Tree with Gene Duplication Events'
        elif 'species' in filename.lower():
            title = 'Species Phylogenetic Tree'
        else:
            title = 'Phylogenetic Tree'
    
    if title:
        ax.set_title(title, fontsize=title_fontsize, fontweight='bold', 
                    color=label_color, pad=20)
    
    # Add subtle grid if publication ready
    if publication_ready:
        ax.grid(True, alpha=0.2, linestyle='--', linewidth=0.5)
    
    # Adjust layout and spacing
    plt.tight_layout()
    
    # Add frame around the plot for publication
    if publication_ready:
        for spine in ax.spines.values():
            spine.set_edgecolor(label_color)
            spine.set_alpha(0.8)
    
    # Save the plot with high quality settings
    if save_path:
        # Save PNG with high DPI
        plt.savefig(f"{save_path}.png", dpi=dpi, bbox_inches='tight', 
                   facecolor=background_color, edgecolor='none', 
                   transparent=False, format='png')
        
        # Save PDF for vector graphics
        plt.savefig(f"{save_path}.pdf", bbox_inches='tight', 
                   facecolor=background_color, edgecolor='none', 
                   transparent=False, format='pdf')
        
        print(f"Publication-quality tree saved as {save_path}.png and {save_path}.pdf")
    
    # Display the plot
    plt.show()
    
    return fig
```

### 1.2 Species Tree Visualizations

#### Rooted Species Tree
- [Tree File](.SPECIES_TREE)

```{python}
# Publication-ready rooted species tree with enhanced styling
plot_phylogenetic_tree_with_lengths(
    newick_file_path=SPECIES_TREE,
    figsize=(16, 12),
    branch_label_precision=4,
    title="Species Phylogenetic Tree (Rooted)",
    title_fontsize=24,
    label_fontsize=20,
    branch_label_fontsize=16,
    branch_color='#2E4057',
    label_color='#1A1A1A',
    show_scale_bar=True,
    save_path=f"{OUTPUT_FIGURES}/SpeciesTree_rooted",
    dpi=300,
    publication_ready=True
)
```

#### Species Tree with Node Labels
- [Tree File](.SPECIES_TREE_NODE_LABELS)

```{python}
# Publication-ready species tree with node labels
plot_phylogenetic_tree_with_lengths(
    newick_file_path=SPECIES_TREE_NODE_LABELS,
    figsize=(16, 12),
    branch_label_precision=3,
    title="Species Phylogenetic Tree with Node Labels",
    title_fontsize=24,
    label_fontsize=20,
    branch_label_fontsize=16,
    branch_color='#2E4057',
    label_color='#1A1A1A',
    show_scale_bar=True,
    save_path=f"{OUTPUT_FIGURES}/SpeciesTree_rooted_node_labels",
    dpi=300,
    publication_ready=True
)
```

#### Gene Duplications Tree
- [Tree File](.GENE_DUPLICATIONS)

```{python}
# Publication-ready gene duplication tree with enhanced visualization
plot_phylogenetic_tree_with_lengths(
    newick_file_path=GENE_DUPLICATIONS,
    figsize=(16, 12),
    branch_label_precision=3,
    title="Species Tree with Gene Duplication Events (≥50% Support)",
    title_fontsize=24,
    label_fontsize=20,
    branch_label_fontsize=16,
    branch_color='#8B4513',  # Brown color to distinguish duplication events
    label_color='#1A1A1A',
    show_scale_bar=True,
    save_path=f"{OUTPUT_FIGURES}/SpeciesTree_Gene_Duplications_0.5_Support",
    dpi=300,
    publication_ready=True
)
```

## 1.5. Advanced Species Tree Visualization with R ggtree

### 1.5.1 R ggtree Functions for Colorful Phylogenetic Trees

```{r}
# Load required libraries
library(ggtree)
library(ape)
library(dplyr)
library(RColorBrewer)
library(viridis)
library(ggplot2)
library(treeio)
library(phytools)

# Function to assign colors to clades based on clustering
assign_clade_colors <- function(tree, n_clusters = 5, color_palette = "Set1", method = "height") {
    # Automatically assign colors to tree clades based on clustering algorithms.
    # 
    # Parameters:
    # - tree: phylo object, the phylogenetic tree
    # - n_clusters: int, number of color clusters to create (default: 5)
    # - color_palette: str, color palette name ('Set1', 'Set2', 'viridis', 'plasma', 'inferno')
    # - method: str, clustering method ('height', 'depth', 'support')
    # 
    # Returns:
    # - list: color assignment for each clade/tip
    
    # Get tree information
    n_tips <- length(tree$tip.label)
    n_nodes <- tree$Nnode
    
    # Generate color palette
    if (color_palette %in% c("viridis", "plasma", "inferno", "magma", "cividis")) {
        # Use appropriate viridis function based on palette name
        if (color_palette == "viridis") {
            colors <- viridis::viridis(n = n_clusters)
        } else if (color_palette == "plasma") {
            colors <- viridis::plasma(n = n_clusters)
        } else if (color_palette == "inferno") {
            colors <- viridis::inferno(n = n_clusters)
        } else if (color_palette == "magma") {
            colors <- viridis::magma(n = n_clusters)
        } else if (color_palette == "cividis") {
            colors <- viridis::cividis(n = n_clusters)
        }
    } else if (color_palette %in% c("Set1", "Set2", "Set3", "Dark2", "Paired")) {
        colors <- RColorBrewer::brewer.pal(min(n_clusters, 8), color_palette)
        if (n_clusters > 8) {
            # Extend palette if needed
            colors <- colorRampPalette(colors)(n_clusters)
        }
    } else {
        # Custom palette
        colors <- rainbow(n_clusters)
    }
    
    # Cluster assignment based on method
    if (method == "height") {
        # Cluster based on node heights
        node_heights <- node.depth.edgelength(tree)
        height_threshold <- quantile(node_heights, probs = seq(0, 1, length.out = n_clusters + 1))
        
        tip_colors <- rep(colors[1], n_tips)
        for (i in 1:n_tips) {
            tip_height <- node_heights[i]
            cluster_idx <- findInterval(tip_height, height_threshold, rightmost.closed = TRUE)
            cluster_idx <- max(1, min(cluster_idx, n_clusters))
            tip_colors[i] <- colors[cluster_idx]
        }
    } else if (method == "depth") {
        # Cluster based on tree depth (distance from root)
        distances <- cophenetic(tree)
        # Use hierarchical clustering on distances
        hclust_result <- hclust(as.dist(distances))
        clusters <- cutree(hclust_result, k = n_clusters)
        tip_colors <- colors[clusters]
    } else {
        # Default: simple alternating colors
        tip_colors <- rep_len(colors, n_tips)
    }
    
    names(tip_colors) <- tree$tip.label
    return(tip_colors)
}

# Function to create publication-quality ggtree plot
plot_phylogenetic_tree_ggtree <- function(
    newick_file_path,
    # Layout and structure options
    layout = "rectangular",  # rectangular, circular, fan, radial, unrooted
    branch_length = TRUE,
    ladderize = TRUE,
    
    # Coloring and styling options
    color_scheme = "clade_auto",  # clade_auto, clade_manual, branch_length, uniform, gradient
    color_palette = "Set1",
    n_clusters = 5,
    clade_method = "depth",  # height, depth, support
    
    # Branch and tip styling
    branch_color = "#2E4057",
    branch_size = 1.2,
    tip_size = 4,
    tip_shape = 16,
    
    # Label options
    show_tip_labels = TRUE,
    tip_label_size = 18,
    tip_label_color = "#1A1A1A",
    tip_label_offset = 0.1,
    tip_label_angle = 0,
    
    # Branch length and support options
    show_branch_length = FALSE,
    branch_length_size = 14,
    branch_length_color = "#666666",
    branch_length_digits = 3,
    show_node_support = FALSE,
    node_support_size = 14,
    
    # Scale and axes
    show_scale_bar = TRUE,
    scale_bar_position = "bottomleft",
    show_axis = FALSE,
    
    # Title and theme options
    title = NULL,
    title_size = 22,
    plot_width = 14,
    plot_height = 10,
    
    # Publication options
    theme_style = "publication",  # publication, minimal, classic, modern
    background_color = "white",
    dpi = 300,
    output_formats = c("png", "pdf"), # output_formats = c("png", "pdf", "svg"),
    save_path = NULL,
    
    # Advanced visualization options
    highlight_clades = NULL,  # vector of clade names to highlight
    highlight_color = "#FF6B6B",
    add_heatmap = FALSE,
    heatmap_data = NULL,
    add_barplot = FALSE,
    barplot_data = NULL,
    add_node_labels = FALSE,
    node_label_size = 8
) {
    # Create publication-quality phylogenetic trees with advanced colorful clustering using ggtree.
    # 
    # This function provides comprehensive customization options for phylogenetic tree visualization,
    # including automatic clade coloring, publication-ready themes, and multiple output formats.
    # 
    # Parameters:
    # - newick_file_path: str, path to the Newick format tree file
    # - layout: str, tree layout type (rectangular, circular, fan, radial, unrooted)
    # - branch_length: logical, whether to use branch lengths for plotting
    # - ladderize: logical, whether to ladderize the tree for better visualization
    # - color_scheme: str, coloring strategy for the tree
    # - color_palette: str, color palette for automatic coloring
    # - n_clusters: int, number of color clusters for clade coloring
    # - save_path: str, optional base path to save the figure (without extension)
    # 
    # Returns:
    # - ggtree object: the created tree plot
    
    # Load the tree
    tryCatch({
        tree <- read.tree(newick_file_path)
    }, error = function(e) {
        stop(paste("Error reading tree file", newick_file_path, ":", e$message))
    })
    
    # Ladderize tree for better visualization
    if (ladderize) {
        tree <- ladderize(tree)
    }
    
    # Create base ggtree plot
    p <- ggtree(tree, layout = layout, branch.length = if(branch_length) "branch.length" else "none")
    
    # Apply coloring scheme
    if (color_scheme == "clade_auto") {
        # Automatic clade coloring
        tip_colors <- assign_clade_colors(tree, n_clusters, color_palette, clade_method)
        
        # Create a data frame for tip colors
        tip_data <- data.frame(
            label = names(tip_colors),
            color = tip_colors,
            stringsAsFactors = FALSE
        )
        
        # Add tip colors to the plot
        p <- p %<+% tip_data +
            geom_tippoint(aes(color = color), size = tip_size, shape = tip_shape) +
            scale_color_identity()
        
        # Color branches by clade if requested
        if (color_scheme == "clade_auto") {
            # Create grouping for branches
            p <- p + geom_tree(aes(color = I(branch_color)), size = branch_size)
        }
    } else if (color_scheme == "gradient") {
        # Gradient coloring based on branch lengths or node depth
        p <- p + geom_tree(aes(color = branch.length), size = branch_size) +
            scale_color_viridis_c(option = "plasma", name = "Branch Length")
    } else if (color_scheme == "uniform") {
        # Uniform coloring
        p <- p + geom_tree(color = branch_color, size = branch_size) +
            geom_tippoint(color = branch_color, size = tip_size, shape = tip_shape)
    } else if (color_scheme == "branch_length") {
        # Color by branch length
        p <- p + geom_tree(aes(color = branch.length), size = branch_size) +
            scale_color_viridis_c(name = "Branch Length") +
            geom_tippoint(color = "#333333", size = tip_size, shape = tip_shape)
    }
    
    # Add tip labels if requested
    if (show_tip_labels) {
        p <- p + geom_tiplab(
            size = tip_label_size / 3.5,  # Convert from points to ggplot units
            color = tip_label_color,
            offset = tip_label_offset,
            angle = tip_label_angle,
            fontface = "italic",  # Italic style for taxonomic nomenclature
            hjust = if(layout == "rectangular") 1 else if(tip_label_angle == 0) 0 else 0.5
        )
    }
    
    # Add branch lengths if requested
    if (show_branch_length) {
        p <- p + geom_text2(
            aes(label = sprintf(paste0("%.", branch_length_digits, "f"), branch.length),
                subset = branch.length > 0),
            size = branch_length_size / 3.5,
            color = branch_length_color,
            nudge_x = 0.01,
            nudge_y = 0.01
        )
    }
    
    # Add node support values if requested
    if (show_node_support && !is.null(tree$node.label)) {
        p <- p + geom_nodelab(
            size = node_support_size / 3.5,
            color = branch_length_color,
            nudge_x = -0.02,
            nudge_y = 0.02
        )
    }
    
    # Add scale bar
    if (show_scale_bar && branch_length) {
        # Calculate scale bar length (10% of tree width)
        tree_width <- max(node.depth.edgelength(tree))
        scale_length <- tree_width * 0.1
        
        p <- p + geom_treescale(
            x = if(scale_bar_position %in% c("bottomleft", "topleft")) 0 else tree_width * 0.8,
            y = if(scale_bar_position %in% c("bottomleft", "bottomright")) 1 else length(tree$tip.label),
            width = scale_length,
            color = branch_color,
            fontsize = 6  # Increased from 3 to 6 for better readability
        )
    }
    
    # Highlight specific clades if requested
    if (!is.null(highlight_clades)) {
        for (clade_name in highlight_clades) {
            if (clade_name %in% tree$tip.label) {
                clade_node <- which(tree$tip.label == clade_name)
                p <- p + geom_highlight(node = clade_node, fill = highlight_color, alpha = 0.3)
            }
        }
    }
    
    # Add title
    if (is.null(title)) {
        # Auto-generate title based on filename
        filename <- basename(newick_file_path)
        if (grepl("rooted", filename, ignore.case = TRUE)) {
            title <- "Species Phylogenetic Tree (Rooted) - ggtree"
        } else if (grepl("gene.*duplication", filename, ignore.case = TRUE)) {
            title <- "Species Tree with Gene Duplication Events - ggtree"
        } else if (grepl("species", filename, ignore.case = TRUE)) {
            title <- "Species Phylogenetic Tree - ggtree"
        } else {
            title <- "Phylogenetic Tree - ggtree"
        }
    }
    
    # Apply theme
    if (theme_style == "publication") {
        p <- p + theme_tree2() +
            theme(
                plot.title = element_text(size = title_size, face = "bold", hjust = 0.5, family = "serif"),
                plot.background = element_rect(fill = background_color, color = NA),
                panel.background = element_rect(fill = background_color, color = NA),
                legend.background = element_rect(fill = background_color, color = NA),
                text = element_text(family = "serif"),
                axis.title = element_text(size = 18),  # Increased from 12 to 18
                axis.text = element_text(size = 16),   # Increased from 10 to 16
                legend.title = element_text(size = 18, face = "bold"),  # Added explicit legend title size
                legend.text = element_text(size = 16),   # Added explicit legend text size
                legend.key.size = unit(1.2, "lines")    # Increased legend key size for better visibility
            ) +
            labs(title = title)
    } else if (theme_style == "minimal") {
        p <- p + theme_tree() +
            theme_minimal() +
            theme(
                plot.title = element_text(size = title_size, face = "bold", hjust = 0.5),
                plot.background = element_rect(fill = background_color, color = NA),
                legend.title = element_text(size = 16, face = "bold"),  # Added explicit legend title size
                legend.text = element_text(size = 14),   # Added explicit legend text size
                legend.key.size = unit(1.1, "lines")    # Increased legend key size
            ) +
            labs(title = title)
    } else {
        p <- p + theme_tree() +
            theme(
                legend.title = element_text(size = 14, face = "bold"),  # Added explicit legend title size
                legend.text = element_text(size = 12),   # Added explicit legend text size
                legend.key.size = unit(1.0, "lines")    # Added legend key size
            ) +
            labs(title = title)
    }
    
    # Remove axes if not requested
    if (!show_axis) {
        p <- p + theme(
            axis.text = element_blank(),
            axis.ticks = element_blank(),
            axis.line = element_blank()
        )
    }
    
    # Save the plot if requested
    if (!is.null(save_path)) {
        for (format in output_formats) {
            filename <- paste0(save_path, "_ggtree.", format)
            
            ggsave(
                filename = filename,
                plot = p,
                width = plot_width,
                height = plot_height,
                dpi = dpi,
                device = format,
                units = "in",
                bg = background_color
            )
            
            cat("Saved:", filename, "\\n")
        }
    }
    
    return(p)
}

# Function to create circular tree with enhanced clade coloring
plot_circular_tree_ggtree <- function(newick_file_path, tip_label_angle = 0, tip_label_offset = 0.02, ...) {
    # Create a circular phylogenetic tree with advanced clade coloring.
    plot_phylogenetic_tree_ggtree(
        newick_file_path = newick_file_path,
        layout = "circular",
        tip_label_angle = tip_label_angle,
        tip_label_offset = tip_label_offset,
        ...
    )
}

# Function to create fan tree with gradient coloring
plot_fan_tree_ggtree <- function(newick_file_path, color_scheme = "gradient", ...) {
    # Create a fan-shaped phylogenetic tree with gradient branch coloring.
    plot_phylogenetic_tree_ggtree(
        newick_file_path = newick_file_path,
        layout = "fan",
        color_scheme = color_scheme,
        ...
    )
}

# Function to create comparison plots with multiple color schemes
create_tree_comparison_plots <- function(newick_file_path, save_base_path = NULL) {
    # Create a comparison of the same tree with different color schemes and layouts.
    
    schemes <- list(
        list(name = "Auto_Clade", scheme = "clade_auto", palette = "Set1"),
        list(name = "Viridis_Gradient", scheme = "gradient", palette = "viridis"),
        list(name = "Plasma_Clade", scheme = "clade_auto", palette = "plasma"),
        list(name = "Branch_Length", scheme = "branch_length", palette = "viridis")
    )
    
    plots <- list()
    
    for (i in seq_along(schemes)) {
        scheme_info <- schemes[[i]]
        
        p <- plot_phylogenetic_tree_ggtree(
            newick_file_path = newick_file_path,
            color_scheme = scheme_info$scheme,
            color_palette = scheme_info$palette,
            tip_label_size = 20,
            title = paste("Tree Visualization:", scheme_info$name),
            save_path = if(!is.null(save_base_path)) paste0(save_base_path, "_", scheme_info$name) else NULL
        )
        
        plots[[scheme_info$name]] <- p
    }
    
    return(plots)
}
```

### 1.5.2 Advanced Species Tree Visualizations with R ggtree

#### Publication-Quality Rooted Species Tree with Clade Coloring
- [Tree File](.SPECIES_TREE)

```{r}
# Publication-quality rooted species tree with automatic clade coloring
p_rooted_colorful <- plot_phylogenetic_tree_ggtree(
    newick_file_path = SPECIES_TREE,
    layout = "rectangular",
    color_scheme = "clade_auto",
    color_palette = "Set1",
    n_clusters = 6,
    clade_method = "depth",
    branch_size = 1.5,
    tip_size = 5,
    show_tip_labels = TRUE,
    tip_label_size = 20,
    tip_label_offset = 0.01,
    show_scale_bar = TRUE,
    title = "Species Phylogenetic Tree (Rooted) - Colorful Clades",
    title_size = 24,
    plot_width = 16,
    plot_height = 12,
    theme_style = "publication",
    save_path = f"{OUTPUT_FIGURES}/SpeciesTree_rooted_ggtree_colorful",
    # output_formats = c("png", "pdf", "svg")
    output_formats = c("png", "pdf")
)

print(p_rooted_colorful)
```

#### Circular Tree with Plasma Color Palette
```{r}
# Beautiful circular tree with plasma color scheme
p_circular <- plot_circular_tree_ggtree(
    newick_file_path = SPECIES_TREE,
    color_scheme = "clade_auto",
    color_palette = "plasma",
    n_clusters = 5,
    branch_size = 1.2,
    tip_size = 6,
    show_tip_labels = TRUE,
    tip_label_size = 18,
    title = "Species Phylogenetic Tree - Circular Layout (Plasma Colors)",
    title_size = 22,
    plot_width = 14,
    plot_height = 14,
    save_path = f"{OUTPUT_FIGURES}/SpeciesTree_circular_plasma",
    output_formats = c("png", "pdf")
)

print(p_circular)
```

#### Fan Tree with Gradient Branch Coloring
```{r}
# Fan-shaped tree with gradient coloring based on branch lengths
p_fan_gradient <- plot_fan_tree_ggtree(
    newick_file_path = SPECIES_TREE,
    color_scheme = "gradient",
    color_palette = "viridis",
    branch_size = 1.0,
    tip_size = 4,
    show_tip_labels = TRUE,
    tip_label_size = 16,
    tip_label_angle = 0,
    show_branch_length = TRUE,
    branch_length_digits = 3,
    title = "Species Tree - Fan Layout with Branch Length Gradient",
    title_size = 22,
    plot_width = 14,
    plot_height = 14,
    save_path = f"{OUTPUT_FIGURES}/SpeciesTree_fan_gradient",
    output_formats = c("png", "pdf")
)

print(p_fan_gradient)
```

#### Gene Duplications Tree with Enhanced Styling
- [Tree File](.GENE_DUPLICATIONS)

```{r}
# Gene duplication tree with special highlighting and node support
p_gene_dup <- plot_phylogenetic_tree_ggtree(
    newick_file_path = GENE_DUPLICATIONS,
    layout = "rectangular",
    color_scheme = "clade_auto",
    color_palette = "Dark2",
    n_clusters = 7,
    clade_method = "height",
    branch_size = 1.8,
    tip_size = 5,
    show_tip_labels = TRUE,
    tip_label_size = 20,
    show_node_support = TRUE,
    node_support_size = 16,
    show_branch_length = TRUE,
    branch_length_digits = 3,
    title = "Species Tree with Gene Duplication Events - Enhanced ggtree",
    title_size = 24,
    plot_width = 18,
    plot_height = 12,
    theme_style = "publication",
    save_path = f"{OUTPUT_FIGURES}/SpeciesTree_GeneDup_ggtree_enhanced",
    # output_formats = c("png", "pdf", "svg")
    output_formats = c("png", "pdf")
)

print(p_gene_dup)
```

#### Multi-Layout Comparison
```{r}
# Create comparison plots with different visualization schemes
tree_comparisons <- create_tree_comparison_plots(
    newick_file_path = SPECIES_TREE,
    save_base_path = f"{OUTPUT_FIGURES}/SpeciesTree_comparison"
)

# Display all comparison plots
for (plot_name in names(tree_comparisons)) {
    cat("\\n=== Tree Visualization:", plot_name, "===\\n")
    print(tree_comparisons[[plot_name]])
}
```

#### Interactive Tree Exploration
```{r}
# Advanced tree with multiple annotation layers
p_advanced <- plot_phylogenetic_tree_ggtree(
    newick_file_path = SPECIES_TREE_NODE_LABELS,
    layout = "rectangular",
    color_scheme = "clade_auto",
    color_palette = "Spectral",
    n_clusters = 8,
    clade_method = "depth",
    branch_size = 2.0,
    tip_size = 6,
    show_tip_labels = TRUE,
    tip_label_size = 22,
    tip_label_offset = 0.005,
    show_node_support = TRUE,
    node_support_size = 18,
    show_branch_length = TRUE,
    branch_length_digits = 4,
    show_scale_bar = TRUE,
    title = "Advanced Species Tree - Multi-Annotated Visualization",
    title_size = 26,
    plot_width = 20,
    plot_height = 14,
    theme_style = "publication",
    save_path = f"{OUTPUT_FIGURES}/SpeciesTree_advanced_multi_annotated",
    output_formats = c("png", "pdf"), # output_formats = c("png", "pdf", "svg"),
    dpi = 600  # Ultra-high resolution for publication
)

print(p_advanced)
```

## 1.6 Time-Calibrated Species Tree (Chronogram with MYA)

This section converts the OrthoFinder species tree from substitution-based branch lengths to a **time-calibrated chronogram** with divergence times in **millions of years ago (MYA)**. This is accomplished using:

1. **Penalized likelihood rate-smoothing** via `ape::chronos()` (Sanderson 2002)
2. **Calibration points** from the TimeTree database and published literature
3. **Visualization** with `ggtree` + `revts()` for reversed time axis

### Calibration Sources

| Node | Age (Ma) | Range (Ma) | Source |
|------|----------|------------|--------|
| Root: Nemacheilidae vs Cyprinidae | ~100 | 80–120 | Hirt et al. 2017; Cypriniformes phylogenomics |
| *Danio rerio* vs Cyprininae | ~49 | 40–60 | Xu et al. 2019 (Nature Comm); TimeTree 5 |
| *Carassius auratus* vs *Cyprinus carpio* | ~23 | 10–35 | Xu et al. 2019; TimeTree 5 |

### References
- Hirt MV et al. (2017) Phylogenetic relationships of the cypriniform family Xenocyprididae. *Biological Journal of the Linnean Society*.
- Xu P et al. (2019) The allotetraploid origin and asymmetrical genome evolution of the common carp *Cyprinus carpio*. *Nature Communications* 10:4625.
- Kumar S et al. (2022) TimeTree 5: An Expanded Resource for Species Divergence Times. *Molecular Biology and Evolution* 39(8):msac174.
- Sanderson MJ (2002) Estimating absolute rates of molecular evolution and divergence times: a penalized likelihood approach. *Molecular Biology and Evolution* 19(1):101-109.

### 1.6.1 Time Calibration with R ape::chronos()

```{r}
# =============================================================================
# Time-Calibrated Chronogram: Convert OrthoFinder species tree to MYA
# =============================================================================
# This section uses penalized likelihood (ape::chronos) to convert
# substitution-based branch lengths to divergence times in MYA.
# Calibration points are derived from TimeTree 5 and published literature.
# =============================================================================

library(ape)
library(ggtree)
library(treeio)
library(ggplot2)

# Ensure path variables are available in R
# (Quarto cross-language sharing should provide SPECIES_TREE and OUTPUT_FIGURES
#  from the Python cells above, but we add a fallback using R config loader)
if (!exists("SPECIES_TREE") || !exists("OUTPUT_FIGURES")) {
    cat("Loading R configuration...\n")
    source("./Notebooks/config/load_config.R")
    paths <- load_paths()
    SPECIES_TREE <- get_path("orthofinder.species_tree")
    OUTPUT_FIGURES <- get_path("outputs.figures")
    cat("SPECIES_TREE:", SPECIES_TREE, "\n")
    cat("OUTPUT_FIGURES:", OUTPUT_FIGURES, "\n")
}

# Check if deeptime is available for geological timescale bars
deeptime_available <- requireNamespace("deeptime", quietly = TRUE)
if (deeptime_available) {
    library(deeptime)
    cat("deeptime package loaded — geological timescale bars available\n")
} else {
    cat("deeptime package not installed — geological timescale bars disabled\n")
    cat("Install with: install.packages('deeptime')\n")
}

# ─────────────────────────────────────────────────────────────────────
# Step 1: Load the OrthoFinder species tree
# ─────────────────────────────────────────────────────────────────────
tree <- read.tree(SPECIES_TREE)
cat("Species tree loaded with", Ntip(tree), "tips\n")
cat("Tips:", paste(tree$tip.label, collapse = ", "), "\n\n")

# Ladderize for better visualization
tree <- ladderize(tree)

# ─────────────────────────────────────────────────────────────────────
# Step 2: Define calibration points from literature
# ─────────────────────────────────────────────────────────────────────
# We dynamically detect which species pairs are present in the tree
# and apply the corresponding calibration points.

# Known calibration data: list of species pairs with divergence times
# Each entry: list(sp1, sp2, age_min_Ma, age_max_Ma, source)
calibration_database <- list(
    # Root calibration: Nemacheilidae (Triplophysa) vs Cyprinidae
    # Cypriniformes divergence — Hirt et al. 2017, Zhu et al. 2017
    list(
        sp1 = c("Triplophysa_tibetana", "Triplophysa_bombifrons",
                 "Triplophysa_dalaica", "Triplophysa_pappenheimi",
                 "Triplophysa_yaopeizhii"),
        sp2 = c("Danio_rerio", "Cyprinus_carpio", "Carassius_auratus"),
        age_min = 80, age_max = 120,
        source = "Root: Nemacheilidae vs Cyprinidae (Hirt et al. 2017)"
    ),
    # Danio rerio vs Cyprininae
    # D. rerio–C. carpio ~60 Ma (Xu et al. 2019), D. rerio–C. idella ~44 Ma
    # TimeTree consensus ~49 Ma
    list(
        sp1 = c("Danio_rerio"),
        sp2 = c("Cyprinus_carpio", "Carassius_auratus", "Labeo_rohita",
                 "Sinocyclocheilus_grahami", "Aspiorhynchus_laticeps",
                 "Schizothorax_macropogon", "Onychostoma_macrolepis",
                 "Anabarilius_grahami", "Diptychus_maculatus",
                 "Oxygymnocypris_stewartii", "Gymnocypris_eckloni",
                 "Schizopygopsis_younghusbandi"),
        age_min = 40, age_max = 60,
        source = "Danio vs Cyprininae (Xu et al. 2019; TimeTree 5)"
    ),
    # Carassius auratus vs Cyprinus carpio
    # Progenitor divergence ~23 Ma (Xu et al. 2019, Nature Comm)
    # TimeTree range: 10.1–61.0 Ma; conservative: 10–35 Ma
    list(
        sp1 = c("Carassius_auratus"),
        sp2 = c("Cyprinus_carpio"),
        age_min = 10, age_max = 35,
        source = "Carassius vs Cyprinus (Xu et al. 2019; TimeTree 5)"
    )
)

# Find applicable calibration points based on species in the tree
tip_labels <- tree$tip.label
calibration_nodes <- c()
calibration_age_min <- c()
calibration_age_max <- c()
calibration_sources <- c()

for (cal in calibration_database) {
    # Find which species from each group are in the tree
    sp1_in_tree <- intersect(cal$sp1, tip_labels)
    sp2_in_tree <- intersect(cal$sp2, tip_labels)

    if (length(sp1_in_tree) > 0 && length(sp2_in_tree) > 0) {
        # Find the MRCA node for the first available species from each group
        all_species_for_mrca <- c(sp1_in_tree[1], sp2_in_tree[1])
        node_num <- getMRCA(tree, all_species_for_mrca)

        if (!is.null(node_num) && !is.na(node_num)) {
            calibration_nodes <- c(calibration_nodes, node_num)
            calibration_age_min <- c(calibration_age_min, cal$age_min)
            calibration_age_max <- c(calibration_age_max, cal$age_max)
            calibration_sources <- c(calibration_sources, cal$source)
            cat("✓ Calibration applied:", cal$source, "\n")
            cat("  Node:", node_num,
                "| Species:", all_species_for_mrca[1], "&", all_species_for_mrca[2],
                "| Age:", cal$age_min, "-", cal$age_max, "Ma\n")
        }
    }
}

cat("\nTotal calibration points:", length(calibration_nodes), "\n")

# ─────────────────────────────────────────────────────────────────────
# Step 3: Create calibration data frame and run chronos()
# ─────────────────────────────────────────────────────────────────────
if (length(calibration_nodes) == 0) {
    stop("No calibration points could be matched to species in the tree!")
}

calib_df <- data.frame(
    node = calibration_nodes,
    age.min = calibration_age_min,
    age.max = calibration_age_max,
    soft.bounds = rep(FALSE, length(calibration_nodes))
)

cat("\nCalibration data frame:\n")
print(calib_df)

# Run penalized likelihood dating with correlated rates model
# lambda = 1 is a reasonable default; model = "correlated" allows rate variation
cat("\nRunning ape::chronos() with correlated rates model...\n")
chrono_tree <- chronos(
    tree,
    calibration = calib_df,
    lambda = 1,
    model = "correlated"
)

# Verify ultrametric property
cat("\nIs ultrametric:", is.ultrametric(chrono_tree), "\n")

# Get root age and node depths
node_depths <- node.depth.edgelength(chrono_tree)
root_age <- max(node_depths)
cat("Estimated root age:", round(root_age, 2), "MYA\n")

# Print all tip-to-root distances (should be equal for ultrametric tree)
tip_distances <- node_depths[1:Ntip(chrono_tree)]
cat("Tip-to-root distances (range):",
    round(min(tip_distances), 4), "-", round(max(tip_distances), 4), "\n")

# Print estimated divergence times for all internal nodes
cat("\nEstimated divergence times for internal nodes:\n")
for (i in (Ntip(chrono_tree) + 1):(Ntip(chrono_tree) + chrono_tree$Nnode)) {
    # Use ape::extract.clade to get tip labels (no phytools dependency)
    subclade <- extract.clade(chrono_tree, i)
    desc_names <- subclade$tip.label
    node_age <- root_age - node_depths[i]
    if (length(desc_names) <= 4) {
        cat(sprintf("  Node %d (%.1f Ma): %s\n", i, node_age,
                    paste(desc_names, collapse = " + ")))
    } else {
        cat(sprintf("  Node %d (%.1f Ma): %s + ... (%d spp)\n", i, node_age,
                    paste(desc_names[1:2], collapse = " + "), length(desc_names)))
    }
}

# Save the calibrated tree for downstream use
chrono_tree_path <- paste0(dirname(SPECIES_TREE), "/SpeciesTree_chronogram_MYA.tre")
write.tree(chrono_tree, chrono_tree_path)
cat("\n✓ Chronogram saved to:", chrono_tree_path, "\n")
```

### 1.6.2 Publication-Quality Chronogram Visualization (ggtree)

```{r}
# =============================================================================
# Time-Calibrated Tree Visualization with MYA Axis
# =============================================================================

# Helper function to format species names (replace _ with space for display)
format_species_name <- function(name) {
    gsub("_", " ", name)
}

# Calculate axis parameters from root age
root_age_rounded <- ceiling(root_age / 10) * 10  # Round up to nearest 10
axis_breaks <- seq(0, root_age_rounded, by = 10)
if (root_age_rounded > 80) {
    axis_breaks <- seq(0, root_age_rounded, by = 20)
}

# ─────────────────────────────────────────────────────────────────────
# Plot 1: Rectangular chronogram with MYA axis (publication-ready)
# ─────────────────────────────────────────────────────────────────────

# Create base ggtree plot with divergence time labels at nodes
p_chrono <- ggtree(chrono_tree, size = 2.0, color = "#2E4057") +
    geom_tiplab(
        aes(label = gsub("_", " ", label)),
        fontface = "italic",
        size = 7,            # ~24.5pt — publication-ready
        offset = root_age * 0.02,
        color = "#1A1A1A"
    ) +
    geom_nodepoint(
        color = "#4A90D9",
        size = 4,
        alpha = 0.8,
        shape = 16
    ) +
    # Add divergence time (MYA) labels directly on the tree at each node
    geom_text2(
        aes(
            label = ifelse(!isTip, sprintf("%.1f", root_age - x), ""),
            subset = !isTip
        ),
        size = 5.5,          # ~19pt — clearly readable
        color = "#CC4444",
        fontface = "bold",
        nudge_y = 0.35,
        nudge_x = -root_age * 0.005
    ) +
    theme_tree2()

# Reverse time scale: past on left, present (0) on right
p_chrono_mya <- revts(p_chrono) +
    scale_x_continuous(
        name = "Divergence Time (Million Years Ago)",
        breaks = -rev(axis_breaks),
        labels = rev(axis_breaks),
        expand = expansion(mult = c(0.02, 0.25))
    ) +
    theme(
        axis.title.x = element_text(size = 22, face = "bold", color = "#1A1A1A",
                                     family = "serif"),
        axis.text.x = element_text(size = 20, color = "#333333", family = "serif"),
        plot.title = element_text(size = 26, face = "bold", hjust = 0.5,
                                   color = "#1A1A1A", family = "serif"),
        plot.subtitle = element_text(size = 18, hjust = 0.5, color = "#666666",
                                      family = "serif"),
        plot.margin = margin(10, 40, 10, 10)
    ) +
    labs(
        title = "Time-Calibrated Species Phylogeny (Chronogram)",
        subtitle = paste0("Penalized likelihood dating (chronos, λ=1) | Root age: ",
                          round(root_age, 1), " MYA")
    )

print(p_chrono_mya)

# Save rectangular chronogram
ggsave(
    filename = paste0(OUTPUT_FIGURES, "/SpeciesTree_chronogram_MYA_ggtree.png"),
    plot = p_chrono_mya,
    width = 18, height = 12, dpi = 300, bg = "white"
)
ggsave(
    filename = paste0(OUTPUT_FIGURES, "/SpeciesTree_chronogram_MYA_ggtree.pdf"),
    plot = p_chrono_mya,
    width = 18, height = 12, bg = "white"
)
cat("✓ Rectangular chronogram saved\n")
```

### 1.6.3 Chronogram with Node Age Labels

```{r}
# =============================================================================
# Chronogram with divergence time labels at each internal node
# =============================================================================

# Calculate node ages for annotation
n_tips <- Ntip(chrono_tree)
n_internal <- chrono_tree$Nnode
node_ages <- data.frame(
    node = (n_tips + 1):(n_tips + n_internal),
    age = round(root_age - node_depths[(n_tips + 1):(n_tips + n_internal)], 1)
)

# Create chronogram with node age annotations
p_chrono_annotated <- ggtree(chrono_tree, size = 2.0, color = "#2E4057") +
    geom_tiplab(
        aes(label = gsub("_", " ", label)),
        fontface = "italic",
        size = 7,            # ~24.5pt — publication-ready
        offset = root_age * 0.02,
        color = "#1A1A1A"
    ) +
    geom_nodepoint(
        color = "#4A90D9",
        size = 4,
        alpha = 0.8,
        shape = 16
    ) +
    # Add divergence time labels at nodes
    geom_text2(
        aes(
            label = ifelse(!isTip, sprintf("%.1f", root_age - x), ""),
            subset = !isTip
        ),
        size = 5.5,          # ~19pt — clearly readable
        color = "#CC4444",
        fontface = "bold",
        nudge_y = 0.35,
        nudge_x = -root_age * 0.005
    ) +
    theme_tree2()

# Apply reversed time scale
p_chrono_annotated <- revts(p_chrono_annotated) +
    scale_x_continuous(
        name = "Divergence Time (Million Years Ago)",
        breaks = -rev(axis_breaks),
        labels = rev(axis_breaks),
        expand = expansion(mult = c(0.02, 0.25))
    ) +
    theme(
        axis.title.x = element_text(size = 22, face = "bold", color = "#1A1A1A",
                                     family = "serif"),
        axis.text.x = element_text(size = 20, color = "#333333", family = "serif"),
        plot.title = element_text(size = 26, face = "bold", hjust = 0.5,
                                   color = "#1A1A1A", family = "serif"),
        plot.subtitle = element_text(size = 18, hjust = 0.5, color = "#666666",
                                      family = "serif"),
        plot.margin = margin(10, 40, 10, 10)
    ) +
    labs(
        title = "Time-Calibrated Species Phylogeny with Divergence Times",
        subtitle = paste0("Node labels show estimated divergence times (MYA) | ",
                          "Calibrated with ", length(calibration_nodes),
                          " TimeTree/literature calibration points")
    )

print(p_chrono_annotated)

# Save annotated chronogram
ggsave(
    filename = paste0(OUTPUT_FIGURES, "/SpeciesTree_chronogram_MYA_annotated_ggtree.png"),
    plot = p_chrono_annotated,
    width = 18, height = 12, dpi = 300, bg = "white"
)
ggsave(
    filename = paste0(OUTPUT_FIGURES, "/SpeciesTree_chronogram_MYA_annotated_ggtree.pdf"),
    plot = p_chrono_annotated,
    width = 18, height = 12, bg = "white"
)
cat("✓ Annotated chronogram saved\n")
```

### 1.6.4 Chronogram with Geological Timescale (deeptime)

```{r}
# =============================================================================
# Chronogram with Geological Epoch/Period Bars (requires 'deeptime' package)
# =============================================================================
# Install deeptime if needed: install.packages("deeptime")
# The deeptime package adds geological timescale bars (epochs, periods)
# to phylogenetic trees, providing geological context for divergence events.
# =============================================================================

if (deeptime_available) {
    cat("Creating chronogram with geological timescale bars...\n")

    # Build the base plot
    p_geo_base <- ggtree(chrono_tree, size = 2.0, color = "#2E4057") +
        geom_tiplab(
            aes(label = gsub("_", " ", label)),
            fontface = "italic",
            size = 7,            # ~24.5pt — publication-ready
            offset = root_age * 0.02,
            color = "#1A1A1A"
        ) +
        geom_nodepoint(
            color = "#4A90D9",
            size = 4,
            alpha = 0.8,
            shape = 16
        ) +
        # Add divergence time labels directly on the tree
        geom_text2(
            aes(
                label = ifelse(!isTip, sprintf("%.1f", root_age - x), ""),
                subset = !isTip
            ),
            size = 5.5,          # ~19pt — clearly readable
            color = "#CC4444",
            fontface = "bold",
            nudge_y = 0.35
        ) +
        theme_tree2()

    # Add geological timescale using coord_geo()
    # NOTE: xlim right bound must extend past 0 to leave room for tip labels
    # (species names), otherwise coord_geo clips them at x=0 (present).
    p_geo <- revts(p_geo_base) +
        coord_geo(
            xlim = c(-(root_age_rounded + 10), root_age * 0.30),
            ylim = c(-2, Ntip(chrono_tree) + 1),
            neg = TRUE,
            abbrv = list(TRUE, FALSE),   # Abbreviate narrow epochs, keep periods full
            pos = list("bottom", "bottom"),
            dat = list("epochs", "periods"),
            size = list(2.5, 4.5),       # Smaller epoch text to avoid overlap
            center_end_labels = TRUE
        ) +
        scale_x_continuous(
            breaks = -rev(axis_breaks),
            labels = rev(axis_breaks)
        ) +
        theme(
            axis.title.x = element_blank(),
            axis.text.x = element_text(size = 20, color = "#333333"),
            plot.title = element_text(size = 26, face = "bold", hjust = 0.5,
                                       family = "serif"),
            plot.subtitle = element_text(size = 18, hjust = 0.5, color = "#666666",
                                          family = "serif"),
            plot.margin = margin(10, 40, 10, 10)
        ) +
        labs(
            title = "Time-Calibrated Species Phylogeny with Geological Context",
            subtitle = paste0("Epochs and periods shown below | Root age: ",
                              round(root_age, 1), " MYA")
        )

    print(p_geo)

    # Save geological context chronogram
    ggsave(
        filename = paste0(OUTPUT_FIGURES, "/SpeciesTree_chronogram_geological_ggtree.png"),
        plot = p_geo,
        width = 20, height = 14, dpi = 300, bg = "white"
    )
    ggsave(
        filename = paste0(OUTPUT_FIGURES, "/SpeciesTree_chronogram_geological_ggtree.pdf"),
        plot = p_geo,
        width = 20, height = 14, bg = "white"
    )
    cat("✓ Geological context chronogram saved\n")

} else {
    cat("Skipping geological timescale visualization (deeptime not installed)\n")
    cat("To enable, install deeptime: install.packages('deeptime')\n")
}
```

### 1.6.5 Time-Calibrated Tree Visualization (Python - Bio.Phylo)

```{python}
# =============================================================================
# Python Visualization of the Time-Calibrated Chronogram
# =============================================================================
# This cell reads the R-calibrated chronogram and creates a matplotlib plot
# with MYA axis labels. If the R-calibrated tree is not yet available,
# it falls back to a simple linear rescaling approach.
# =============================================================================

from Bio import Phylo
import matplotlib.pyplot as plt
import numpy as np
from pathlib import Path

# Try to load the R-calibrated chronogram first
chrono_tree_path = Path(SPECIES_TREE).parent / "SpeciesTree_chronogram_MYA.tre"

if chrono_tree_path.exists():
    tree = Phylo.read(str(chrono_tree_path), "newick")
    source_label = "ape::chronos() penalized likelihood"
    print(f"✓ Loaded R-calibrated chronogram from {chrono_tree_path}")
else:
    # Fallback: simple linear rescaling
    # Scale the OrthoFinder tree so root depth = estimated root age
    tree = Phylo.read(SPECIES_TREE, "newick")
    
    # Known approximate root age from literature (Nemacheilidae vs Cyprinidae)
    ESTIMATED_ROOT_AGE_MYA = 100.0
    
    # Calculate current max root-to-tip distance
    depths = tree.depths()
    max_depth = max(depths.values())
    scale_factor = ESTIMATED_ROOT_AGE_MYA / max_depth if max_depth > 0 else 1.0
    
    # Rescale all branch lengths
    for clade in tree.find_clades():
        if clade.branch_length is not None:
            clade.branch_length *= scale_factor
    
    source_label = f"Linear rescaling (root = {ESTIMATED_ROOT_AGE_MYA} MYA)"
    print(f"⚠ R-calibrated tree not found. Using linear rescaling.")
    print(f"  Root age set to {ESTIMATED_ROOT_AGE_MYA} MYA")

# Calculate tree depth for axis configuration
depths = tree.depths()
max_depth = max(depths.values())
root_age_mya = max_depth

print(f"Root age: {root_age_mya:.1f} MYA")
print(f"Source: {source_label}")

# ─────────────────────────────────────────────────────────────────────
# Create publication-quality plot
# ─────────────────────────────────────────────────────────────────────
plt.style.use('default')
plt.rcParams.update({
    'font.family': 'serif',
    'font.serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
    'font.size': 14,
    'axes.linewidth': 1.2,
    'figure.facecolor': 'white',
    'axes.facecolor': 'white'
})

fig, ax = plt.subplots(1, 1, figsize=(16, 12), facecolor='white')

# Draw the tree
Phylo.draw(
    tree,
    axes=ax,
    do_show=False,
    branch_labels=None,
    label_func=lambda x: x.name.replace('_', ' ') if x.name else ''
)

# Customize branch appearance
for line in ax.get_lines():
    line.set_color('#2E4057')
    line.set_linewidth(2.0)

# Customize text labels
for text in ax.texts:
    text_content = text.get_text().strip()
    if text_content:
        text.set_color('#1A1A1A')
        text.set_fontsize(16)
        text.set_fontstyle('italic')

# Configure MYA axis (reverse direction: root age on left, 0 on right)
ax.set_xlim(-root_age_mya * 0.02, root_age_mya * 1.15)

# Create MYA labels: reverse so max is on left, 0 on right
tick_interval = 20 if root_age_mya > 80 else 10
tick_positions = np.arange(0, root_age_mya + tick_interval, tick_interval)

# The x-axis in Bio.Phylo goes from 0 (root) to max (tips)
# We want to label it as MYA decreasing from root_age to 0
ax.set_xticks(tick_positions)
ax.set_xticklabels([f'{root_age_mya - t:.0f}' for t in tick_positions], fontsize=14)
ax.set_xlabel('Divergence Time (Million Years Ago)', fontsize=16, fontweight='bold')

# Remove unnecessary spines
ax.spines['top'].set_visible(False)
ax.spines['right'].set_visible(False)
ax.spines['left'].set_visible(False)
ax.spines['bottom'].set_visible(True)
ax.spines['bottom'].set_color('#333333')

# Add subtle grid lines at major time points
for t in tick_positions[1:]:
    ax.axvline(x=t, color='#DDDDDD', linewidth=0.5, linestyle='--', zorder=0)

# Add title
ax.set_title(
    'Time-Calibrated Species Phylogeny (Chronogram)',
    fontsize=20, fontweight='bold', color='#1A1A1A', pad=20
)

# Add source annotation
ax.text(
    0.02, 0.02,
    f'Calibration: {source_label}',
    transform=ax.transAxes,
    fontsize=10, color='#888888',
    verticalalignment='bottom'
)

plt.tight_layout()

# Save
plt.savefig(f"{OUTPUT_FIGURES}/SpeciesTree_chronogram_MYA_python.png",
            dpi=300, bbox_inches='tight', facecolor='white')
plt.savefig(f"{OUTPUT_FIGURES}/SpeciesTree_chronogram_MYA_python.pdf",
            bbox_inches='tight', facecolor='white')
print(f"✓ Python chronogram saved to {OUTPUT_FIGURES}/SpeciesTree_chronogram_MYA_python.png|pdf")

plt.show()
```

## 2. Individual Orthogroup Analysis

### 2.1 Orthogroup Bar Chart Function

```{python}
import pandas as pd
import matplotlib.pyplot as plt

def plot_orthogroup_bar_chart(tsv_file_path, orthogroup_id):
    """
    Create bar chart showing gene counts per species for a specific orthogroup.
    """
    # Load the TSV file into a pandas DataFrame
    df = pd.read_csv(tsv_file_path, sep='\t')
    
    # Find the row corresponding to the given orthogroup_id
    row = df[df['Orthogroup'] == orthogroup_id]
    
    if row.empty:
        print(f"No data found for Orthogroup ID: {orthogroup_id}")
        return

    # Extract the data, excluding the 'Orthogroup' and 'Total' columns
    data = row.iloc[0].drop(labels=['Orthogroup', 'Total']).to_dict()

    # Create a bar plot for the specified orthogroup_id
    plt.figure(figsize=(10, 6))
    plt.bar(data.keys(), data.values())
    plt.title(f'Gene Counts for {orthogroup_id}')
    plt.xlabel('Species')
    plt.ylabel('Gene Count')
    plt.xticks(rotation=45, ha='right')
    
    # Show the plot
    plt.tight_layout()
    plt.savefig(f"{OUTPUT_FIGURES}/OG_bars/" + orthogroup_id + "_OG_genecount_barplot.png")
    plt.savefig(f"{OUTPUT_FIGURES}/OG_bars/" + orthogroup_id + "_OG_genecount_barplot.pdf")
    plt.show()
```

### 2.2 Example Individual Orthogroups

#### OG0000000
```{python}
plot_orthogroup_bar_chart(ORTHOGROUPS_COUNT, 'OG0000000')
```

#### OG0000001
```{python}
plot_orthogroup_bar_chart(ORTHOGROUPS_COUNT, 'OG0000001')
```

### 2.3 Batch Generation of All Orthogroup Bar Charts

```{python}
# Generate bar charts for all orthogroups
df = pd.read_csv(ORTHOGROUPS_COUNT, sep='\t')
orthogroup_list = df['Orthogroup'].tolist()

print(f"Generating bar charts for {len(orthogroup_list)} orthogroups...")
for i, orthogroup in enumerate(orthogroup_list):
    if i % 1000 == 0:
        print(f"Progress: {i}/{len(orthogroup_list)} orthogroups processed")
    plot_orthogroup_bar_chart(ORTHOGROUPS_COUNT, orthogroup)
print("All orthogroup bar charts generated!")
```

## 3. Gene Tree Visualization

### 3.1 Gene Tree Analysis Functions

```{python}
from Bio import Phylo
import matplotlib.pyplot as plt

def select_representatives(tree, max_depth=2):
    """
    Selects representative clade labels based on tree depth.

    Parameters:
    - tree: Phylogenetic tree object loaded using Bio.Phylo.
    - max_depth: int, the maximum depth level to consider clades from.

    Returns:
    - representative_labels: list of str, containing clade names considered representative.
    """
    representative_labels = []
    
    def visit_clade(clade, current_depth):
        if current_depth == max_depth:
            # Leaf node at the required depth
            if clade.name:
                representative_labels.append(clade.name)
        else:
            for child in clade.clades:
                visit_clade(child, current_depth + 1)
    
    visit_clade(tree.root, 0)
    return representative_labels

def prune_and_plot_tree(newick_file_path, max_depth=2, figsize=(13, 10), orientation='horizontal', save_path=None):
    """
    Load a phylogenetic tree from a Newick file, select representatives based on depth, and plot.

    Parameters:
    - newick_file_path: str, path to the Newick file containing the tree.
    - max_depth: int, the maximum depth level to keep as representatives.
    - figsize: tuple(int, int), optional, size of the plot figure.
    - orientation: str, 'horizontal' or 'vertical' to orient the tree.
    - save_path: str, optional, base path to save the figure (saves as PNG and PDF)

    Returns:
    - None, but displays the pruned phylogenetic tree plot.
    """
    # Load the tree from the Newick file
    tree = Phylo.read(newick_file_path, "newick")
    
    # Automatically select representative clades
    representative_labels = select_representatives(tree, max_depth)

    # Prune the tree
    for clade in tree.get_terminals():
        if clade.name not in representative_labels:
            # Remove unwanted clades
            tree.prune(target=clade)

    # Configure Matplotlib figure
    fig = plt.figure(figsize=figsize)
    axes = fig.add_subplot(1, 1, 1)
    
    # Draw the pruned tree
    Phylo.draw(tree, axes=axes, do_show=False)

    # Adjust tree orientation if needed
    if orientation == 'vertical':
        axes.set_aspect('equal')
        plt.xticks(rotation=90)

    # Save the plot if path provided
    if save_path:
        plt.savefig(f"{save_path}.png")
        plt.savefig(f"{save_path}.pdf")
    
    # Display the plot
    plt.tight_layout()
    plt.show()
```

### 3.2 Example Gene Trees

#### OG0000000 Gene Tree
- Tree File: `{orthofinder.gene_trees}/OG0000000_tree.txt` (from config)

```{python}
prune_and_plot_tree(f'{GENE_TREES_DIR}/OG0000000_tree.txt', max_depth=8, save_path=f"{OUTPUT_FIGURES}/OG0000000_gene_tree")
```

#### OG0000001 Gene Tree
- Tree File: `{orthofinder.gene_trees}/OG0000001_tree.txt` (from config)

```{python}
prune_and_plot_tree(f'{GENE_TREES_DIR}/OG0000001_tree.txt', max_depth=8, save_path=f"{OUTPUT_FIGURES}/OG0000001_gene_tree")
```

## 4. Comprehensive Heatmap Analysis

### 4.1 Heatmap Analysis Functions

```{python}
import pandas as pd
import seaborn as sns
import matplotlib.pyplot as plt
from typing import Optional, Tuple, List

def load_and_process_orthogroup_data(
    file_path: str, 
    drop_total: bool = True
) -> pd.DataFrame:
    """
    Load and preprocess orthogroup gene count data.
    
    Parameters:
    - file_path: str, path to the TSV file containing orthogroup gene counts
    - drop_total: bool, whether to drop the 'Total' column (default: True)
    
    Returns:
    - pd.DataFrame: processed data with Orthogroup as index
    """
    try:
        data = pd.read_csv(file_path, sep='\t')
        data.set_index('Orthogroup', inplace=True)
        
        if drop_total and 'Total' in data.columns:
            data = data.drop(columns='Total')
        
        return data
    except Exception as e:
        raise ValueError(f"Error loading data from {file_path}: {str(e)}")

def create_publication_heatmap(
    data: pd.DataFrame,
    title: str,
    save_path: str,
    figsize: Tuple[int, int] = (12, 10),
    cmap: str = 'viridis',
    annot: bool = False,
    xlabel: str = 'Species',
    ylabel: str = 'Orthogroups',
    rotation: int = 45,
    dpi: int = 300,
    publication_ready: bool = True,
    show_dendrograms: bool = False,
    cluster_rows: bool = False,
    cluster_cols: bool = False,
    center_cmap: bool = False,
    robust_color_scale: bool = True,
    add_colorbar_label: str = 'Gene Count',
    font_scale: float = 1.2,
    cell_aspect: str = 'auto',
    border_color: str = 'white',
    border_width: float = 0.5,
    # Advanced publication options
    data_transform: str = 'none',
    show_statistics: bool = False,
    stat_position: str = 'bottom_right',
    colorbar_extend: str = 'neither',
    annotation_threshold: Optional[float] = None,
    custom_annotation_func: Optional[callable] = None,
    subplot_margins: dict = None,
    journal_style: str = 'nature',
    add_grid: bool = False,
    grid_style: dict = None,
    mask_zeros: bool = False,
    highlight_diagonal: bool = False,
    add_subplot_labels: bool = False,
    subplot_label_style: dict = None
) -> None:
    """
    Create an advanced publication-quality heatmap with comprehensive styling, statistical overlays, 
    and professional formatting optimized for scientific journals.
    
    Parameters:
    - data: pd.DataFrame, data to plot
    - title: str, plot title
    - save_path: str, base path to save the figure (saves as PNG and PDF)
    - figsize: tuple, figure size in inches
    - cmap: str, colormap ('viridis', 'plasma', 'publication_blue', 'publication_warm', 
             'nature_blue', 'science_red', 'cell_green', 'plos_orange')
    - annot: bool, whether to annotate cells with values
    - xlabel: str, x-axis label
    - ylabel: str, y-axis label  
    - rotation: int, x-axis label rotation angle
    - dpi: int, resolution for saved figures (300+ recommended for publication)
    - publication_ready: bool, apply publication-ready styling
    - show_dendrograms: bool, whether to show clustering dendrograms
    - cluster_rows: bool, whether to cluster rows hierarchically
    - cluster_cols: bool, whether to cluster columns hierarchically
    - center_cmap: bool, whether to center colormap at zero
    - robust_color_scale: bool, use robust percentiles for color scaling
    - add_colorbar_label: str, label for the colorbar
    - font_scale: float, scaling factor for fonts
    - cell_aspect: str, aspect ratio of cells ('auto', 'equal', or float)
    - border_color: str, color of cell borders
    - border_width: float, width of cell borders
    
    Advanced Options:
    - data_transform: str, data transformation ('none', 'log10', 'log2', 'zscore', 'minmax')
    - show_statistics: bool, overlay statistical information on the plot
    - stat_position: str, position for statistics ('bottom_right', 'top_left', 'bottom_left', 'top_right')
    - colorbar_extend: str, colorbar extension ('neither', 'both', 'min', 'max')
    - annotation_threshold: float, optional threshold for conditional cell annotations
    - custom_annotation_func: callable, custom function for cell annotations
    - subplot_margins: dict, custom margins ({'left': 0.1, 'right': 0.9, 'top': 0.9, 'bottom': 0.1})
    - journal_style: str, journal-specific styling ('nature', 'science', 'cell', 'plos', 'generic')
    - add_grid: bool, whether to add a subtle grid overlay
    - grid_style: dict, grid styling options
    - mask_zeros: bool, whether to mask zero values
    - highlight_diagonal: bool, whether to highlight diagonal elements (for symmetric matrices)
    - add_subplot_labels: bool, whether to add subplot labels (A, B, C, etc.)
    - subplot_label_style: dict, styling for subplot labels
    """
    import seaborn as sns
    import numpy as np
    from matplotlib.colors import LinearSegmentedColormap, LogNorm, PowerNorm
    from matplotlib.patches import Rectangle
    from sklearn.preprocessing import StandardScaler, MinMaxScaler
    from scipy import stats
    import matplotlib.patches as mpatches
    
    # Validate input data
    if data.empty:
        raise ValueError("Input data is empty")
    
    # Apply data transformations
    data_processed = data.copy()

    # Format species names for display (replace underscores with spaces)
    data_processed.columns = data_processed.columns.str.replace('_', ' ')
    data_processed.index = data_processed.index.str.replace('_', ' ')

    transform_label = ""

    if data_transform == 'log10':
        data_processed = np.log10(data_processed + 1)  # Add 1 to handle zeros
        transform_label = " (log₁₀ transformed)"
        add_colorbar_label += " (log₁₀)"
    elif data_transform == 'log2':
        data_processed = np.log2(data_processed + 1)
        transform_label = " (log₂ transformed)"
        add_colorbar_label += " (log₂)"
    elif data_transform == 'zscore':
        scaler = StandardScaler()
        data_processed = pd.DataFrame(
            scaler.fit_transform(data_processed),
            index=data.index,
            columns=data.columns
        )
        transform_label = " (Z-score normalized)"
        add_colorbar_label = "Z-score"
    elif data_transform == 'minmax':
        scaler = MinMaxScaler()
        data_processed = pd.DataFrame(
            scaler.fit_transform(data_processed),
            index=data.index,
            columns=data.columns
        )
        transform_label = " (Min-Max normalized)"
        add_colorbar_label = "Normalized Value"
    
    # Journal-specific styling configurations
    journal_configs = {
        'nature': {
            'font_family': 'sans-serif',
            'font_sans': ['Arial', 'Helvetica', 'DejaVu Sans'],
            'title_size': int(20 * font_scale),
            'label_size': int(18 * font_scale),
            'tick_size': int(16 * font_scale),
            'cmap_default': 'viridis'
        },
        'science': {
            'font_family': 'serif',
            'font_serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
            'title_size': int(22 * font_scale),
            'label_size': int(20 * font_scale),
            'tick_size': int(18 * font_scale),
            'cmap_default': 'plasma'
        },
        'cell': {
            'font_family': 'sans-serif',
            'font_sans': ['Arial', 'Helvetica', 'DejaVu Sans'],
            'title_size': int(21 * font_scale),
            'label_size': int(19 * font_scale),
            'tick_size': int(17 * font_scale),
            'cmap_default': 'cell_green'
        },
        'plos': {
            'font_family': 'serif',
            'font_serif': ['Liberation Serif', 'Times New Roman', 'DejaVu Serif'],
            'title_size': int(20 * font_scale),
            'label_size': int(18 * font_scale),
            'tick_size': int(16 * font_scale),
            'cmap_default': 'plos_orange'
        },
        'generic': {
            'font_family': 'serif',
            'font_serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
            'title_size': int(22 * font_scale),
            'label_size': int(20 * font_scale),
            'tick_size': int(18 * font_scale),
            'cmap_default': 'viridis'
        }
    }
    
    # Set publication-ready style with journal-specific settings
    config = journal_configs.get(journal_style, journal_configs['generic'])
    
    if publication_ready:
        plt.style.use('default')
        rcParams_config = {
            'font.family': config['font_family'],
            'font.size': config['tick_size'],
            'axes.linewidth': 1.8,
            'axes.spines.top': False,
            'axes.spines.right': False,
            'figure.facecolor': 'white',
            'axes.facecolor': 'white',
            'savefig.facecolor': 'white',
            'savefig.edgecolor': 'none',
            'axes.edgecolor': '#333333',
            'axes.labelcolor': '#333333',
            'text.color': '#333333',
            'xtick.color': '#333333',
            'ytick.color': '#333333',
            'patch.linewidth': 0.5,
            'patch.facecolor': '#348ABD',
            'patch.edgecolor': '#EEEEEE',
            'patch.antialiased': True
        }
        
        # Add font-specific settings
        if config['font_family'] == 'serif':
            rcParams_config['font.serif'] = config['font_serif']
        else:
            rcParams_config['font.sans-serif'] = config['font_sans']
            
        plt.rcParams.update(rcParams_config)
    
    # Advanced colormap options with journal-specific schemes
    if cmap == 'publication_blue':
        colors = ['#f7fbff', '#deebf7', '#c6dbef', '#9ecae1', '#6baed6', '#4292c6', '#2171b5', '#084594']
        cmap = LinearSegmentedColormap.from_list('publication_blue', colors)
    elif cmap == 'publication_warm':
        colors = ['#fff7ec', '#fee8c8', '#fdd49e', '#fdbb84', '#fc8d59', '#ef6548', '#d7301f', '#990000']
        cmap = LinearSegmentedColormap.from_list('publication_warm', colors)
    elif cmap == 'nature_blue':
        colors = ['#ffffff', '#e6f3ff', '#cce7ff', '#99d6ff', '#66c2ff', '#3399ff', '#0066cc', '#004499']
        cmap = LinearSegmentedColormap.from_list('nature_blue', colors)
    elif cmap == 'science_red':
        colors = ['#ffffff', '#ffe6e6', '#ffcccc', '#ff9999', '#ff6666', '#ff3333', '#cc0000', '#990000']
        cmap = LinearSegmentedColormap.from_list('science_red', colors)
    elif cmap == 'cell_green':
        colors = ['#ffffff', '#e6ffe6', '#ccffcc', '#99ff99', '#66ff66', '#33cc33', '#009900', '#006600']
        cmap = LinearSegmentedColormap.from_list('cell_green', colors)
    elif cmap == 'plos_orange':
        colors = ['#ffffff', '#fff2e6', '#ffe6cc', '#ffd699', '#ffc266', '#ffad33', '#ff9900', '#cc7700']
        cmap = LinearSegmentedColormap.from_list('plos_orange', colors)
    
    # Configure color scaling with advanced options
    vmin, vmax = None, None
    if robust_color_scale:
        vmin, vmax = np.percentile(data_processed.values.flatten(), [2, 98])
    
    if center_cmap:
        # Center colormap at zero for diverging data
        max_abs = max(abs(data_processed.min().min()), abs(data_processed.max().max()))
        vmin, vmax = -max_abs, max_abs
        if cmap in ['viridis', 'plasma', 'nature_blue']:  # Switch to diverging colormap
            cmap = 'RdBu_r'
    
    # Create mask for zero values if requested
    mask = None
    if mask_zeros:
        mask = data_processed == 0
    
    # Create figure with enhanced layout control
    if subplot_margins is None:
        subplot_margins = {'left': 0.12, 'right': 0.88, 'top': 0.92, 'bottom': 0.15}
    
    fig, ax = plt.subplots(figsize=figsize, facecolor='white')
    fig.subplots_adjust(**subplot_margins)
    
    # Enhanced annotation logic
    annot_data = None
    if annot:
        if custom_annotation_func:
            annot_data = data_processed.applymap(custom_annotation_func)
        elif annotation_threshold is not None:
            # Conditional annotations - only show values above threshold
            annot_data = data_processed.where(data_processed.abs() >= annotation_threshold, '')
        else:
            annot_data = data_processed
    
    # Enhanced colorbar configuration
    colorbar_kws = {
        'shrink': 0.8,
        'aspect': 25,
        'pad': 0.03,
        'label': add_colorbar_label,
        'extend': colorbar_extend,
        'orientation': 'vertical',
        'format': '%.1f' if data_transform in ['zscore', 'minmax'] else '%.0f'
    }
    
    # Create the heatmap with enhanced options
    heatmap_kwargs = {
        'cmap': cmap,
        'annot': annot_data if annot else False,
        'fmt': '' if annot_data is not None else '.1f',
        'annot_kws': {'fontsize': int(8 * font_scale), 'color': '#333333'},
        'linewidths': border_width,
        'linecolor': border_color,
        'cbar_kws': colorbar_kws,
        'square': cell_aspect == 'equal',
        'xticklabels': True,
        'yticklabels': True,
        'mask': mask
    }
    
    # Add clustering if requested
    if show_dendrograms or cluster_rows or cluster_cols:
        # Use clustermap for hierarchical clustering
        g = sns.clustermap(
            data_processed,
            row_cluster=cluster_rows,
            col_cluster=cluster_cols,
            figsize=figsize,
            dendrogram_ratio=0.15 if show_dendrograms else 0.0,
            colors_ratio=0.03,
            cmap=cmap,
            **{k: v for k, v in heatmap_kwargs.items() if k not in ['square', 'cmap']}
        )
        
        # Enhanced clustermap styling
        g.fig.suptitle(f'{title}{transform_label}', 
                      fontsize=config['title_size'], 
                      fontweight='bold', 
                      y=0.98)
        g.ax_heatmap.set_xlabel(xlabel, fontsize=config['label_size'], labelpad=12)
        g.ax_heatmap.set_ylabel(ylabel, fontsize=config['label_size'], labelpad=12)
        
        # Enhanced tick styling
        plt.setp(g.ax_heatmap.get_xticklabels(), 
                rotation=rotation, ha='right', fontsize=config['tick_size'])
        plt.setp(g.ax_heatmap.get_yticklabels(), 
                rotation=0, fontsize=config['tick_size'])
        
        # Add statistics if requested
        if show_statistics:
            stats_text = _create_statistics_text(data_processed)
            _add_statistics_annotation(g.fig, stats_text, stat_position, font_scale)
        
        # Save the clustermap with enhanced quality
        g.savefig(f"{save_path}.png", dpi=dpi, bbox_inches='tight', 
                 facecolor='white', edgecolor='none')
        g.savefig(f"{save_path}.pdf", bbox_inches='tight', 
                 facecolor='white', edgecolor='none')
        
    else:
        # Standard heatmap with advanced features
        if vmin is not None and vmax is not None:
            heatmap_kwargs.update({'vmin': vmin, 'vmax': vmax})
        
        # Create the main heatmap
        im = sns.heatmap(data_processed, ax=ax, **heatmap_kwargs)
        
        # Highlight diagonal if requested (useful for correlation matrices)
        if highlight_diagonal and data_processed.shape[0] == data_processed.shape[1]:
            for i in range(min(data_processed.shape)):
                rect = Rectangle((i, i), 1, 1, fill=False, edgecolor='black', 
                               linewidth=2, alpha=0.8)
                ax.add_patch(rect)
        
        # Add grid overlay if requested
        if add_grid:
            grid_defaults = {'alpha': 0.2, 'linewidth': 0.5, 'color': '#666666'}
            if grid_style:
                grid_defaults.update(grid_style)
            ax.grid(True, **grid_defaults)
            ax.set_axisbelow(True)
        
        # Enhanced title with transformation info
        full_title = f'{title}{transform_label}'
        ax.set_title(full_title, fontsize=config['title_size'], 
                    fontweight='bold', pad=25, color='#333333')
        
        # Enhanced axis styling
        ax.set_xlabel(xlabel, fontsize=config['label_size'], labelpad=12, color='#333333')
        ax.set_ylabel(ylabel, fontsize=config['label_size'], labelpad=12, color='#333333')
        
        # Advanced tick customization
        ax.tick_params(
            axis='x', 
            rotation=rotation, 
            labelsize=config['tick_size'],
            pad=8,
            colors='#333333',
            which='major',
            length=6,
            width=1.2
        )
        ax.tick_params(
            axis='y', 
            rotation=0, 
            labelsize=config['tick_size'],
            pad=8,
            colors='#333333',
            which='major',
            length=6,
            width=1.2
        )
        
        # Enhanced label alignment
        plt.setp(ax.get_xticklabels(), ha='right', va='center_baseline', fontstyle='italic')
        plt.setp(ax.get_yticklabels(), ha='right', va='center')
        
        # Add subplot labels if requested
        if add_subplot_labels:
            label_defaults = {
                'fontsize': int(16 * font_scale),
                'fontweight': 'bold',
                'color': '#333333'
            }
            if subplot_label_style:
                label_defaults.update(subplot_label_style)
            
            ax.text(-0.15, 1.05, 'A', transform=ax.transAxes, **label_defaults)
        
        # Add statistical overlay if requested
        if show_statistics:
            stats_text = _create_statistics_text(data_processed)
            _add_statistics_annotation(fig, stats_text, stat_position, font_scale)
        
        # Enhanced layout with better spacing
        plt.tight_layout()
        
        # Save figures with publication-quality settings
        save_kwargs = {
            'dpi': dpi,
            'bbox_inches': 'tight',
            'facecolor': 'white',
            'edgecolor': 'none',
            'metadata': {
                'Title': full_title,
                'Author': 'Comparative Genomics Analysis',
                'Subject': 'Heatmap Visualization',
                'Creator': 'Python Matplotlib/Seaborn'
            }
        }
        
        plt.savefig(f"{save_path}.png", format='png', **save_kwargs)
        plt.savefig(f"{save_path}.pdf", format='pdf', **save_kwargs)
        
        # Save high-resolution version for print
        plt.savefig(f"{save_path}_print.png", format='png', dpi=600, 
                   bbox_inches='tight', facecolor='white', edgecolor='none')
    
    plt.show()

def _create_statistics_text(data: pd.DataFrame) -> str:
    """Helper function to create statistical summary text for heatmap overlay."""
    stats_info = []
    
    # Basic statistics
    stats_info.append(f"Shape: {data.shape[0]}×{data.shape[1]}")
    stats_info.append(f"Range: {data.min().min():.2f}–{data.max().max():.2f}")
    stats_info.append(f"Mean: {data.mean().mean():.2f}")
    
    # Calculate correlation if data is square (suitable for species comparison)
    if data.shape[0] == data.shape[1]:
        # Calculate mean off-diagonal correlation for symmetric matrices
        np.fill_diagonal(data.values, np.nan)  # Exclude diagonal
        mean_corr = np.nanmean(data.values)
        stats_info.append(f"Mean off-diag: {mean_corr:.2f}")
    
    return "\\n".join(stats_info)

def _add_statistics_annotation(fig, stats_text: str, position: str, font_scale: float) -> None:
    """Helper function to add statistical annotations to the figure."""
    position_coords = {
        'bottom_right': (0.98, 0.02),
        'top_right': (0.98, 0.98),
        'bottom_left': (0.02, 0.02),
        'top_left': (0.02, 0.98)
    }
    
    position_va = {
        'bottom_right': 'bottom',
        'top_right': 'top',
        'bottom_left': 'bottom',
        'top_left': 'top'
    }
    
    x, y = position_coords.get(position, (0.98, 0.02))
    va = position_va.get(position, 'bottom')
    
    # Add statistics text box
    fig.text(
        x, y, stats_text,
        transform=fig.transFigure,
        fontsize=int(9 * font_scale),
        bbox=dict(
            boxstyle="round,pad=0.5",
            facecolor='white',
            edgecolor='#CCCCCC',
            alpha=0.9
        ),
        ha='right' if 'right' in position else 'left',
        va=va,
        color='#333333'
    )

def filter_orthogroups_by_variance(
    data: pd.DataFrame, 
    top_n: int = 100
) -> pd.DataFrame:
    """
    Filter orthogroups by variance across species.
    
    Parameters:
    - data: pd.DataFrame, orthogroup gene count data
    - top_n: int, number of top variance orthogroups to return
    
    Returns:
    - pd.DataFrame: filtered data with highest variance orthogroups
    """
    data_with_variance = data.copy()
    data_with_variance['Variance'] = data_with_variance.var(axis=1)
    
    top_orthogroups = data_with_variance.sort_values(
        by='Variance', ascending=False
    ).head(top_n)
    
    return top_orthogroups.drop(columns='Variance')

def filter_orthogroups_by_species_extreme(
    data: pd.DataFrame,
    target_species: str,
    extreme_type: str = 'max',
    apply_variance_filter: bool = False,
    variance_top_n: int = 50
) -> Tuple[pd.DataFrame, List[str]]:
    """
    Filter orthogroups where target species has extreme (max/min) gene counts.
    
    Parameters:
    - data: pd.DataFrame, orthogroup gene count data
    - target_species: str, species name to analyze
    - extreme_type: str, 'max' or 'min' for maximum or minimum counts
    - apply_variance_filter: bool, whether to apply additional variance filtering
    - variance_top_n: int, number of top variance orthogroups if filtering
    
    Returns:
    - tuple: (filtered_data, orthogroup_ids)
    """
    if target_species not in data.columns:
        raise ValueError(f"Species '{target_species}' not found in data columns")
    
    if extreme_type == 'max':
        data_extreme = data.max(axis=1)
        mask = data[target_species] == data_extreme
    elif extreme_type == 'min':
        data_extreme = data.min(axis=1)
        mask = data[target_species] == data_extreme
    else:
        raise ValueError("extreme_type must be 'max' or 'min'")
    
    extreme_genes_OG = data.index[mask].tolist()
    filtered_data = data.loc[extreme_genes_OG]
    
    if apply_variance_filter:
        filtered_data = filter_orthogroups_by_variance(filtered_data, variance_top_n)
    
    return filtered_data, extreme_genes_OG

def filter_species_specific_orthogroups(
    data: pd.DataFrame,
    target_species: str,
    apply_variance_filter: bool = False,
    variance_top_n: int = 50
) -> Tuple[pd.DataFrame, List[str]]:
    """
    Filter orthogroups where only the target species has non-zero gene counts.
    
    Parameters:
    - data: pd.DataFrame, orthogroup gene count data
    - target_species: str, species name to analyze
    - apply_variance_filter: bool, whether to apply additional variance filtering
    - variance_top_n: int, number of top variance orthogroups if filtering
    
    Returns:
    - tuple: (filtered_data, orthogroup_ids)
    """
    if target_species not in data.columns:
        raise ValueError(f"Species '{target_species}' not found in data columns")
    
    other_species = data.columns.drop(target_species)
    condition1 = data[target_species] > 0
    condition2 = (data[other_species] == 0).all(axis=1)
    final_mask = condition1 & condition2
    
    unique_nonzero_genes_OG = data.index[final_mask].tolist()
    filtered_data = data.loc[unique_nonzero_genes_OG]
    
    if apply_variance_filter:
        filtered_data = filter_orthogroups_by_variance(filtered_data, variance_top_n)
    
    return filtered_data, unique_nonzero_genes_OG
```

### 4.2 Data Overview

```{python}
import subprocess
subprocess.run(['head', '-n', '10', ORTHOGROUPS_COUNT])
```

### 4.3 Top Variance Heatmap

```{python}
# Load and process the data
file_path = ORTHOGROUPS_COUNT
data = load_and_process_orthogroup_data(file_path)

# Filter by variance and create heatmap
top_variance_data = filter_orthogroups_by_variance(data, top_n=100)

create_publication_heatmap(
    data=top_variance_data,
    title='Gene Counts Heatmap for Top 100 Orthogroups with Highest Variance',
    cmap='coolwarm',
    save_path=f"{OUTPUT_FIGURES}/heatmap_top_variance",
    figsize=(14, 12),
    publication_ready=True
)
```

### 4.4 Focal Species-Focused Analysis

#### 4.4.1 Increased Gene Count Heatmaps

##### Maximum Count Analysis (With Variance Sorting)
```{python}
# Same analysis but with variance filtering
focal = get_focal_species()
focal_display = focal.replace('_', ' ')

max_count_variance_data, _ = filter_orthogroups_by_species_extreme(
    data=data,
    target_species=FOCAL_SPECIES,
    extreme_type='max',
    apply_variance_filter=True,
    variance_top_n=50
)

create_publication_heatmap(
    data=max_count_variance_data,
    title=f"Gene Counts: '{focal_display}' Maximum Count (Top 50 by Variance)",
    cmap='coolwarm',
    save_path=f"{OUTPUT_FIGURES}/heatmap_Aspiorhynchus_max_variance",
    figsize=(16, 16),
    publication_ready=True
)
```

#### 4.4.2 Decreased Gene Count Heatmaps

##### Minimum Count Analysis (With Variance Sorting)
```{python}
# Same analysis but with variance filtering
focal = get_focal_species()
focal_display = focal.replace('_', ' ')

min_count_variance_data, _ = filter_orthogroups_by_species_extreme(
    data=data,
    target_species=FOCAL_SPECIES,
    extreme_type='min',
    apply_variance_filter=True,
    variance_top_n=50
)

create_publication_heatmap(
    data=min_count_variance_data,
    title=f"Gene Counts: '{focal_display}' Minimum Count (Top 50 by Variance)",
    cmap='coolwarm',
    save_path=f"{OUTPUT_FIGURES}/heatmap_Aspiorhynchus_min_variance",
    figsize=(16, 16),
    publication_ready=True
)
```

#### 4.4.3 Species-Specific Heatmaps

##### Species-Specific Analysis (With Variance Sorting)
```{python}
# Same analysis but with variance filtering
focal = get_focal_species()
focal_display = focal.replace('_', ' ')

species_specific_variance_data, _ = filter_species_specific_orthogroups(
    data=data,
    target_species=FOCAL_SPECIES,
    apply_variance_filter=True,
    variance_top_n=50
)

create_publication_heatmap(
    data=species_specific_variance_data,
    title=f"Gene Counts: '{focal_display}' Species-Specific (Top 50 by Variance)",
    cmap='coolwarm',
    save_path=f"{OUTPUT_FIGURES}/heatmap_unique_nonzero_Aspiorhynchus_variance",
    figsize=(16, 16),
    publication_ready=True
)
```

## 5. Species Overlap Analysis

### 5.1 Species Overlap Analysis Functions

```{python}
import pandas as pd
import seaborn as sns
import matplotlib.pyplot as plt
from sklearn.preprocessing import MinMaxScaler
from typing import Optional

def create_species_overlap_heatmap(
    file_path: str,
    save_path: str,
    normalize: bool = False,
    figsize: Tuple[int, int] = (12, 10),
    cmap: str = 'viridis',
    annot: bool = False,
    linewidths: float = 0.8,
    dpi: int = 300,
    publication_ready: bool = True,
    # Advanced styling options
    title: Optional[str] = None,
    xlabel: str = 'Species',
    ylabel: str = 'Species',
    rotation: int = 45,
    font_scale: float = 1.2,
    journal_style: str = 'nature',
    data_transform: str = 'none',
    show_statistics: bool = False,
    stat_position: str = 'bottom_right',
    # Matrix-specific options
    highlight_diagonal: bool = True,
    mask_diagonal: bool = False,
    cluster_rows: bool = False,
    cluster_cols: bool = False,
    show_dendrograms: bool = False,
    # Advanced annotation options
    annotation_threshold: Optional[float] = None,
    custom_annotation_func: Optional[callable] = None,
    # Color and layout options
    colorbar_extend: str = 'neither',
    add_colorbar_label: str = 'Shared Orthogroups',
    robust_color_scale: bool = True,
    add_grid: bool = False,
    grid_style: dict = None,
    subplot_margins: dict = None,
    add_subplot_labels: bool = False,
    # Clustering and correlation options
    show_correlation_values: bool = False,
    correlation_method: str = 'pearson',
    add_significance_stars: bool = False
) -> pd.DataFrame:
    """
    Create an advanced publication-quality heatmap for species overlap analysis with comprehensive
    styling options, statistical overlays, and professional formatting optimized for scientific journals.
    
    Parameters:
    - file_path: str, path to the TSV file containing species overlap data
    - save_path: str, base path to save the figure (saves as PNG and PDF)
    - normalize: bool, whether to normalize the data using MinMaxScaler (default: False)
    - figsize: tuple, figure size in inches
    - cmap: str, colormap ('viridis', 'plasma', 'Blues', 'Oranges', 'RdBu_r', 'publication_blue', 
             'nature_overlap', 'science_matrix', 'cell_symmetric', 'plos_overlap')
    - annot: bool, whether to annotate cells
    - linewidths: float, width of lines separating cells
    - dpi: int, resolution for saved figures (300+ recommended for publication)
    - publication_ready: bool, apply publication-ready styling
    
    Advanced Styling Options:
    - title: str, optional custom title
    - xlabel: str, x-axis label
    - ylabel: str, y-axis label
    - rotation: int, x-axis label rotation angle
    - font_scale: float, scaling factor for fonts
    - journal_style: str, journal-specific styling ('nature', 'science', 'cell', 'plos', 'generic')
    - data_transform: str, data transformation ('none', 'log10', 'sqrt', 'zscore', 'minmax')
    - show_statistics: bool, overlay statistical information on the plot
    - stat_position: str, position for statistics overlay
    
    Matrix-Specific Options:
    - highlight_diagonal: bool, whether to highlight diagonal elements (species vs itself)
    - mask_diagonal: bool, whether to mask/hide diagonal elements
    - cluster_rows: bool, whether to cluster rows hierarchically
    - cluster_cols: bool, whether to cluster columns hierarchically
    - show_dendrograms: bool, whether to show clustering dendrograms
    
    Advanced Annotation Options:
    - annotation_threshold: float, optional threshold for conditional cell annotations
    - custom_annotation_func: callable, custom function for cell annotations
    
    Color and Layout Options:
    - colorbar_extend: str, colorbar extension ('neither', 'both', 'min', 'max')
    - add_colorbar_label: str, label for the colorbar
    - robust_color_scale: bool, use robust percentiles for color scaling
    - add_grid: bool, whether to add a subtle grid overlay
    - grid_style: dict, grid styling options
    - subplot_margins: dict, custom margins
    - add_subplot_labels: bool, whether to add subplot labels (A, B, C, etc.)
    
    Clustering and Correlation Options:
    - show_correlation_values: bool, display correlation coefficients
    - correlation_method: str, correlation method ('pearson', 'spearman', 'kendall')
    - add_significance_stars: bool, add significance indicators
    
    Returns:
    - pd.DataFrame: the loaded (and optionally normalized/transformed) data
    """
    import seaborn as sns
    import numpy as np
    from matplotlib.colors import LinearSegmentedColormap
    from matplotlib.patches import Rectangle
    from sklearn.preprocessing import StandardScaler, MinMaxScaler
    from scipy import stats
    from scipy.cluster.hierarchy import linkage, dendrogram
    import matplotlib.patches as mpatches
    
    try:
        # Load the species overlaps data
        data = pd.read_csv(file_path, sep='\t', index_col=0)
        
        # Apply data transformations
        data_processed = data.copy()

        # Format species names for display (replace underscores with spaces)
        data_processed.columns = data_processed.columns.str.replace('_', ' ')
        data_processed.index = data_processed.index.str.replace('_', ' ')

        transform_label = ""

        if data_transform == 'log10':
            data_processed = np.log10(data_processed + 1)  # Add 1 to handle zeros
            transform_label = " (log₁₀ transformed)"
            add_colorbar_label += " (log₁₀)"
        elif data_transform == 'sqrt':
            data_processed = np.sqrt(data_processed)
            transform_label = " (√ transformed)"
            add_colorbar_label += " (√)"
        elif data_transform == 'zscore':
            scaler = StandardScaler()
            # For symmetric matrices, normalize across all values (flatten)
            values_flat = data_processed.values.flatten()
            values_scaled = scaler.fit_transform(values_flat.reshape(-1, 1)).flatten()
            data_processed = pd.DataFrame(
                values_scaled.reshape(data_processed.shape),
                index=data_processed.index,
                columns=data_processed.columns
            )
            transform_label = " (Z-score normalized)"
            add_colorbar_label = "Z-score"
        elif data_transform == 'minmax':
            if normalize:  # Use existing normalize logic
                scaler = MinMaxScaler()
                data_processed = pd.DataFrame(
                    scaler.fit_transform(data_processed),
                    index=data_processed.index,
                    columns=data_processed.columns
                )
            else:
                # Apply MinMaxScaler to entire matrix for symmetric consistency
                scaler = MinMaxScaler()
                values_flat = data_processed.values.flatten()
                values_scaled = scaler.fit_transform(values_flat.reshape(-1, 1)).flatten()
                data_processed = pd.DataFrame(
                    values_scaled.reshape(data_processed.shape),
                    index=data_processed.index,
                    columns=data_processed.columns
                )
            transform_label = " (Min-Max normalized)"
            add_colorbar_label = "Normalized Value"
        elif normalize and data_transform == 'none':
            # Legacy normalization for backward compatibility
            scaler = MinMaxScaler()
            data_processed = pd.DataFrame(
                scaler.fit_transform(data_processed),
                index=data_processed.index,
                columns=data_processed.columns
            )
            transform_label = " (Normalized)"
        
        # Journal-specific styling configurations
        journal_configs = {
            'nature': {
                'font_family': 'sans-serif',
                'font_sans': ['Arial', 'Helvetica', 'DejaVu Sans'],
                'title_size': int(14 * font_scale),
                'label_size': int(12 * font_scale),
                'tick_size': int(10 * font_scale),
                'cmap_default': 'Blues'
            },
            'science': {
                'font_family': 'serif',
                'font_serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
                'title_size': int(16 * font_scale),
                'label_size': int(14 * font_scale),
                'tick_size': int(12 * font_scale),
                'cmap_default': 'plasma'
            },
            'cell': {
                'font_family': 'sans-serif',
                'font_sans': ['Arial', 'Helvetica', 'DejaVu Sans'],
                'title_size': int(15 * font_scale),
                'label_size': int(13 * font_scale),
                'tick_size': int(11 * font_scale),
                'cmap_default': 'Oranges'
            },
            'plos': {
                'font_family': 'serif',
                'font_serif': ['Liberation Serif', 'Times New Roman', 'DejaVu Serif'],
                'title_size': int(14 * font_scale),
                'label_size': int(12 * font_scale),
                'tick_size': int(10 * font_scale),
                'cmap_default': 'viridis'
            },
            'generic': {
                'font_family': 'serif',
                'font_serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
                'title_size': int(16 * font_scale),
                'label_size': int(14 * font_scale),
                'tick_size': int(12 * font_scale),
                'cmap_default': 'viridis'
            }
        }
        
        # Set publication-ready style with journal-specific settings
        config = journal_configs.get(journal_style, journal_configs['generic'])
        
        if publication_ready:
            plt.style.use('default')
            rcParams_config = {
                'font.family': config['font_family'],
                'font.size': config['tick_size'],
                'axes.linewidth': 2.0,
                'axes.spines.top': False,
                'axes.spines.right': False,
                'figure.facecolor': 'white',
                'axes.facecolor': 'white',
                'savefig.facecolor': 'white',
                'savefig.edgecolor': 'none',
                'axes.edgecolor': '#333333',
                'axes.labelcolor': '#333333',
                'text.color': '#333333',
                'xtick.color': '#333333',
                'ytick.color': '#333333',
                'patch.linewidth': 0.8,
                'patch.facecolor': '#348ABD',
                'patch.edgecolor': '#FFFFFF',
                'patch.antialiased': True
            }
            
            # Add font-specific settings
            if config['font_family'] == 'serif':
                rcParams_config['font.serif'] = config['font_serif']
            else:
                rcParams_config['font.sans-serif'] = config['font_sans']
                
            plt.rcParams.update(rcParams_config)
        
        # Advanced colormap options with overlap-specific schemes
        if cmap == 'publication_blue':
            colors = ['#ffffff', '#deebf7', '#c6dbef', '#9ecae1', '#6baed6', '#4292c6', '#2171b5', '#084594']
            cmap = LinearSegmentedColormap.from_list('publication_blue', colors)
        elif cmap == 'nature_overlap':
            colors = ['#ffffff', '#f0f8ff', '#e1f4ff', '#c2e9ff', '#85d4ff', '#47bfff', '#0099ff', '#0066cc']
            cmap = LinearSegmentedColormap.from_list('nature_overlap', colors)
        elif cmap == 'science_matrix':
            colors = ['#ffffff', '#fff5f0', '#fed4c4', '#fd9483', '#fc4d4d', '#dc143c', '#b22222', '#800000']
            cmap = LinearSegmentedColormap.from_list('science_matrix', colors)
        elif cmap == 'cell_symmetric':
            colors = ['#ffffff', '#f0fff0', '#d4edda', '#b7e5b4', '#85c285', '#52a352', '#2d5a2d', '#1e3d1e']
            cmap = LinearSegmentedColormap.from_list('cell_symmetric', colors)
        elif cmap == 'plos_overlap':
            colors = ['#ffffff', '#fff8dc', '#ffe4b5', '#ffa500', '#ff8c00', '#ff6347', '#dc143c', '#b22222']
            cmap = LinearSegmentedColormap.from_list('plos_overlap', colors)
        
        # Create mask for diagonal if requested
        mask = None
        if mask_diagonal:
            mask = np.eye(len(data_processed), dtype=bool)
        
        # Configure color scaling with advanced options
        vmin, vmax = None, None
        if robust_color_scale:
            if mask_diagonal:
                # Exclude diagonal for color scaling
                non_diag_values = data_processed.values[~np.eye(len(data_processed), dtype=bool)]
                vmin, vmax = np.percentile(non_diag_values, [2, 98])
            else:
                vmin, vmax = np.percentile(data_processed.values.flatten(), [2, 98])
        
        # Create figure with enhanced layout control
        if subplot_margins is None:
            subplot_margins = {'left': 0.15, 'right': 0.88, 'top': 0.92, 'bottom': 0.18}
        
        # Enhanced annotation logic
        annot_data = None
        if annot:
            if custom_annotation_func:
                annot_data = data_processed.applymap(custom_annotation_func)
            elif annotation_threshold is not None:
                # Conditional annotations - only show values above threshold
                annot_data = data_processed.where(data_processed.abs() >= annotation_threshold, '')
            else:
                annot_data = data_processed
        
        # Enhanced colorbar configuration
        colorbar_kws = {
            'shrink': 0.85,
            'aspect': 30,
            'pad': 0.04,
            'label': add_colorbar_label,
            'extend': colorbar_extend,
            'orientation': 'vertical',
            'format': '%.2f' if data_transform in ['zscore', 'minmax'] or normalize else '%.0f'
        }
        
        # Create the heatmap with enhanced options
        heatmap_kwargs = {
            'cmap': cmap,
            'annot': annot_data if annot else False,
            'fmt': '' if annot_data is not None else ('.2f' if data_transform in ['zscore', 'minmax'] or normalize else '.0f'),
            'annot_kws': {'fontsize': int(9 * font_scale), 'color': '#333333', 'weight': 'bold'},
            'linewidths': linewidths,
            'linecolor': 'white',
            'cbar_kws': colorbar_kws,
            'square': True,  # Essential for symmetric overlap matrices
            'xticklabels': True,
            'yticklabels': True,
            'mask': mask
        }
        
        # Add clustering if requested
        if show_dendrograms or cluster_rows or cluster_cols:
            # Use clustermap for hierarchical clustering
            g = sns.clustermap(
                data_processed,
                row_cluster=cluster_rows,
                col_cluster=cluster_cols,
                figsize=figsize,
                dendrogram_ratio=0.15 if show_dendrograms else 0.0,
                colors_ratio=0.03,
                cmap=cmap,
                **{k: v for k, v in heatmap_kwargs.items() if k not in ['square', 'cmap']}
            )
            
            # Enhanced clustermap styling
            plot_title = title if title else f'Species Overlap Matrix{transform_label}'
            g.fig.suptitle(plot_title, 
                          fontsize=config['title_size'], 
                          fontweight='bold', 
                          y=0.98)
            g.ax_heatmap.set_xlabel(xlabel, fontsize=config['label_size'], labelpad=15)
            g.ax_heatmap.set_ylabel(ylabel, fontsize=config['label_size'], labelpad=15)
            
            # Enhanced tick styling
            plt.setp(g.ax_heatmap.get_xticklabels(), 
                    rotation=rotation, ha='right', fontsize=config['tick_size'], fontstyle='italic')
            plt.setp(g.ax_heatmap.get_yticklabels(), 
                    rotation=0, fontsize=config['tick_size'], fontstyle='italic')
            
            # Add statistics if requested
            if show_statistics:
                stats_text = _create_overlap_statistics_text(data_processed, mask_diagonal)
                _add_statistics_annotation(g.fig, stats_text, stat_position, font_scale)
            
            # Save the clustermap with enhanced quality
            g.savefig(f"{save_path}.png", dpi=dpi, bbox_inches='tight', 
                     facecolor='white', edgecolor='none')
            g.savefig(f"{save_path}.pdf", bbox_inches='tight', 
                     facecolor='white', edgecolor='none')
        
        else:
            # Standard heatmap with advanced features
            fig, ax = plt.subplots(figsize=figsize, facecolor='white')
            fig.subplots_adjust(**subplot_margins)
            
            if vmin is not None and vmax is not None:
                heatmap_kwargs.update({'vmin': vmin, 'vmax': vmax})
            
            # Create the main heatmap
            im = sns.heatmap(data_processed, ax=ax, **heatmap_kwargs)
            
            # Highlight diagonal if requested and not masked
            if highlight_diagonal and not mask_diagonal:
                for i in range(min(data_processed.shape)):
                    rect = Rectangle((i, i), 1, 1, fill=False, edgecolor='#FF4444', 
                                   linewidth=3, alpha=0.9)
                    ax.add_patch(rect)
            
            # Add grid overlay if requested
            if add_grid:
                grid_defaults = {'alpha': 0.3, 'linewidth': 0.8, 'color': '#CCCCCC'}
                if grid_style:
                    grid_defaults.update(grid_style)
                ax.grid(True, **grid_defaults)
                ax.set_axisbelow(True)
            
            # Enhanced title with transformation info
            plot_title = title if title else f'Species Overlap Matrix{transform_label}'
            ax.set_title(plot_title, fontsize=config['title_size'], 
                        fontweight='bold', pad=30, color='#333333')
            
            # Enhanced axis styling
            ax.set_xlabel(xlabel, fontsize=config['label_size'], labelpad=15, color='#333333')
            ax.set_ylabel(ylabel, fontsize=config['label_size'], labelpad=15, color='#333333')
            
            # Advanced tick customization
            ax.tick_params(
                axis='x', 
                rotation=rotation, 
                labelsize=config['tick_size'],
                pad=10,
                colors='#333333',
                which='major',
                length=8,
                width=1.5
            )
            ax.tick_params(
                axis='y', 
                rotation=0, 
                labelsize=config['tick_size'],
                pad=10,
                colors='#333333',
                which='major',
                length=8,
                width=1.5
            )
            
            # Enhanced label alignment with italic species names
            plt.setp(ax.get_xticklabels(), ha='right', va='center_baseline', fontstyle='italic')
            plt.setp(ax.get_yticklabels(), ha='right', va='center', fontstyle='italic')
            
            # Add subplot labels if requested
            if add_subplot_labels:
                label_defaults = {
                    'fontsize': int(18 * font_scale),
                    'fontweight': 'bold',
                    'color': '#333333'
                }
                ax.text(-0.20, 1.08, 'A', transform=ax.transAxes, **label_defaults)
            
            # Add statistical overlay if requested
            if show_statistics:
                stats_text = _create_overlap_statistics_text(data_processed, mask_diagonal)
                _add_statistics_annotation(fig, stats_text, stat_position, font_scale)
            
            # Enhanced layout with better spacing
            plt.tight_layout()
            
            # Save figures with publication-quality settings
            save_kwargs = {
                'dpi': dpi,
                'bbox_inches': 'tight',
                'facecolor': 'white',
                'edgecolor': 'none',
                'metadata': {
                    'Title': plot_title,
                    'Author': 'Comparative Genomics Analysis',
                    'Subject': 'Species Overlap Matrix',
                    'Creator': 'Python Matplotlib/Seaborn'
                }
            }
            
            plt.savefig(f"{save_path}.png", format='png', **save_kwargs)
            plt.savefig(f"{save_path}.pdf", format='pdf', **save_kwargs)
            
            # Save high-resolution version for print
            plt.savefig(f"{save_path}_print.png", format='png', dpi=600, 
                       bbox_inches='tight', facecolor='white', edgecolor='none')
        
        plt.show()
        
        return data
        
    except Exception as e:
        raise ValueError(f"Error processing species overlap data from {file_path}: {str(e)}")

def _create_overlap_statistics_text(data: pd.DataFrame, mask_diagonal: bool = False) -> str:
    """Helper function to create statistical summary text for overlap matrix overlay."""
    stats_info = []
    
    # Basic statistics
    stats_info.append(f"Species: {data.shape[0]}")
    
    if mask_diagonal:
        # Exclude diagonal elements for statistics
        mask = ~np.eye(len(data), dtype=bool)
        values = data.values[mask]
        stats_info.append(f"Off-diagonal range: {values.min():.0f}–{values.max():.0f}")
        stats_info.append(f"Mean overlap: {values.mean():.1f}")
        stats_info.append(f"Median overlap: {np.median(values):.1f}")
    else:
        stats_info.append(f"Range: {data.min().min():.0f}–{data.max().max():.0f}")
        stats_info.append(f"Mean: {data.mean().mean():.1f}")
        
        # Calculate mean off-diagonal if symmetric
        if data.shape[0] == data.shape[1]:
            off_diag_mask = ~np.eye(len(data), dtype=bool)
            off_diag_mean = data.values[off_diag_mask].mean()
            stats_info.append(f"Off-diag mean: {off_diag_mean:.1f}")
    
    return "\\n".join(stats_info)

def analyze_species_overlap_statistics(data: pd.DataFrame) -> dict:
    """
    Calculate summary statistics for species overlap data.
    
    Parameters:
    - data: pd.DataFrame, species overlap matrix
    
    Returns:
    - dict: summary statistics including mean, median, min, max overlaps
    """
    # Remove diagonal (self-overlaps) for analysis
    mask = ~pd.DataFrame(index=data.index, columns=data.columns).eq(
        pd.DataFrame(index=data.index, columns=data.columns).T
    ).values
    
    off_diagonal_values = data.values[mask]
    
    stats = {
        'mean_overlap': off_diagonal_values.mean(),
        'median_overlap': pd.Series(off_diagonal_values).median(),
        'min_overlap': off_diagonal_values.min(),
        'max_overlap': off_diagonal_values.max(),
        'std_overlap': off_diagonal_values.std(),
        'total_species': len(data.columns)
    }
    
    return stats

def compute_species_overlap_from_genecount(genecount_path: str) -> pd.DataFrame:
    """
    Compute species overlap matrix from OrthoFinder's Orthogroups.GeneCount.tsv file.

    The species overlap matrix counts how many orthogroups are shared between each
    pair of species (i.e., both species have at least one gene in that orthogroup).

    This is useful when OrthoFinder doesn't generate Orthogroups_SpeciesOverlaps.tsv
    directly (which happens in some run configurations).

    Parameters:
    - genecount_path: str, path to Orthogroups.GeneCount.tsv file

    Returns:
    - pd.DataFrame: symmetric matrix of shared orthogroup counts between species pairs
    """
    # Load gene count data
    genecount = pd.read_csv(genecount_path, sep='\t', index_col=0)

    # Remove the 'Total' column if present
    if 'Total' in genecount.columns:
        genecount = genecount.drop(columns=['Total'])

    # Get species list
    species = genecount.columns.tolist()
    n_species = len(species)

    # Convert to presence/absence matrix (1 if gene_count > 0, else 0)
    presence = (genecount > 0).astype(int)

    # Compute overlap matrix: dot product of presence matrix with its transpose
    # Each element [i,j] = number of orthogroups where both species i and j have genes
    overlap_matrix = presence.T.dot(presence)

    # Create DataFrame with species names as index and columns
    overlap_df = pd.DataFrame(
        overlap_matrix.values,
        index=species,
        columns=species
    )

    print(f"Computed species overlap matrix from {len(genecount)} orthogroups")
    print(f"Species: {n_species}")

    return overlap_df
```

### 5.2 Species Overlap Heatmap (Non-normalized)

```{python}
# Compute species overlap from GeneCount (since SpeciesOverlaps.tsv may not exist)
genecount_path = f'{ORTHOFINDER_RESULTS}/Orthogroups/Orthogroups.GeneCount.tsv'
overlap_data = compute_species_overlap_from_genecount(genecount_path)

# Save the computed overlap data for reference
overlap_output_path = f'{OUTPUT_FIGURES}/Orthogroups_SpeciesOverlaps_computed.tsv'
overlap_data.to_csv(overlap_output_path, sep='\t')
print(f"Saved species overlap matrix to: {overlap_output_path}")
```

```{python}
# Create non-normalized species overlap heatmap
# Note: We pass the pre-computed overlap_data directly instead of a file path
import tempfile
import os

# Save overlap_data to a temporary file for the heatmap function
with tempfile.NamedTemporaryFile(mode='w', suffix='.tsv', delete=False) as f:
    overlap_data.to_csv(f, sep='\t')
    temp_overlap_path = f.name

overlap_result = create_species_overlap_heatmap(
    file_path=temp_overlap_path,
    save_path=f"{OUTPUT_FIGURES}/Orthogroups_SpeciesOverlaps",
    cmap='coolwarm',
    normalize=False,
    figsize=(12, 10),
    publication_ready=True
)

# Clean up temp file
os.unlink(temp_overlap_path)

# Print basic statistics
stats = analyze_species_overlap_statistics(overlap_data)
print("Species Overlap Statistics:")
print(f"Total species: {stats['total_species']}")
print(f"Mean overlap: {stats['mean_overlap']:.2f}")
print(f"Median overlap: {stats['median_overlap']:.2f}")
print(f"Range: {stats['min_overlap']} - {stats['max_overlap']}")
print(f"Standard deviation: {stats['std_overlap']:.2f}")
```

### 5.3 Species Overlap Heatmap (Normalized)

```{python}
# Create normalized species overlap heatmap
with tempfile.NamedTemporaryFile(mode='w', suffix='.tsv', delete=False) as f:
    overlap_data.to_csv(f, sep='\t')
    temp_overlap_path = f.name

create_species_overlap_heatmap(
    file_path=temp_overlap_path,
    save_path=f"{OUTPUT_FIGURES}/Orthogroups_SpeciesOverlaps_Normalized",
    cmap='coolwarm',
    normalize=True,
    figsize=(12, 10),
    publication_ready=True
)

os.unlink(temp_overlap_path)
```

## 6. Gene and Orthogroup Ratio Analysis

### 6.1 Ratio Analysis Functions

```{python}
import pandas as pd
import matplotlib.pyplot as plt
import numpy as np
from typing import Tuple, Optional, List

def create_gene_orthogroup_ratio_plot(
    file_path: str,
    save_path: str,
    plot_type: str = 'stacked_bar',
    figsize: Tuple[int, int] = (14, 10),
    colors: Optional[List[str]] = None,
    title: Optional[str] = None,
    xlabel: str = 'Species',
    ylabel: str = 'Counts',
    rotation: int = 45,
    dpi: int = 300,
    publication_ready: bool = True
) -> pd.DataFrame:
    """
    Create publication-quality plots for gene and orthogroup ratio analysis.
    
    Parameters:
    - file_path: str, path to the TSV/CSV file containing ratio data
    - save_path: str, base path to save the figure (saves as PNG and PDF)
    - plot_type: str, type of plot ('stacked_bar', 'grouped_bar', 'line')
    - figsize: tuple, figure size in inches
    - colors: list, optional custom colors for the plot
    - title: str, optional custom title
    - xlabel: str, x-axis label
    - ylabel: str, y-axis label
    - rotation: int, x-axis label rotation angle
    - dpi: int, resolution for saved figures
    - publication_ready: bool, apply publication-ready styling
    
    Returns:
    - pd.DataFrame: the loaded and processed data
    """
    try:
        # Load the data using the original working approach
        # Note: Despite .tsv extension, this file uses comma separators
        data = pd.read_csv(file_path, sep=',', index_col=0)
        
        # Transpose to have species as index and statistics as columns (original approach)
        data_transposed = data.T
        
        # Validate that we have data to plot
        if data_transposed.empty:
            raise ValueError(f"No data found in {file_path}")
        
        print(f"Data loaded successfully:")
        print(f"  - Shape: {data_transposed.shape}")
        print(f"  - Species: {list(data_transposed.index)}")
        print(f"  - Statistics: {list(data_transposed.columns)}")
        
        # Set publication-ready style
        if publication_ready:
            plt.style.use('default')
            plt.rcParams.update({
                'font.family': 'serif',
                'font.serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
                'font.size': 18,
                'axes.linewidth': 1.2,
                'figure.facecolor': 'white'
            })
        
        # Set default colors if none provided
        if colors is None:
            # Use a professional color palette
            # colors = ['#2E4057', '#048A81', '#54C6EB', '#F18F01', '#C73E1D', '#A8DADC', '#457B9D']
            # colors = ['#264653', '#2A9D8F', '#E9C46A', '#F4A261', '#E76F51', '#F8F9FA', '#495057']
            # colors = ['#FF6B6B', '#4ECDC4', '#45B7D1', '#96CEB4', '#FECA57', '#FF9FF3', '#6C5CE7'] #Better: Vibrant Harmony
            # colors = ['#0F3460', '#16537e', '#1A659E', '#00A8CC', '#7DD3FC', '#B3E5FC', '#E1F5FE']
            colors = ['#2D1B69', '#11998E', '#38EF7D', '#FFB75E', '#ED4A7B', '#945ECF', '#13547A'] #Better: Sunset Palette
            # colors = ['#2E3440', '#3B4252', '#5E81AC', '#88C0D0', '#8FBCBB', '#D8DEE9', '#ECEFF4']
            # colors = ['#1B4332', '#2D6A4F', '#40916C', '#52B788', '#74C69D', '#95D5B2', '#B7E4C7']
        
        # Create the plot
        fig, ax = plt.subplots(figsize=figsize)
        
        if plot_type == 'stacked_bar':
            data_transposed.plot(
                kind='bar', 
                stacked=True, 
                ax=ax, 
                color=colors[:len(data_transposed.columns)],
                alpha=0.8,
                edgecolor='white',
                linewidth=0.8
            )
        elif plot_type == 'grouped_bar':
            data_transposed.plot(
                kind='bar', 
                ax=ax, 
                color=colors[:len(data_transposed.columns)],
                alpha=0.8,
                edgecolor='white',
                linewidth=0.8
            )
        elif plot_type == 'line':
            for i, column in enumerate(data_transposed.columns):
                ax.plot(
                    range(len(data_transposed.index)),
                    data_transposed[column].values,
                    marker='o',
                    linewidth=2,
                    markersize=6,
                    color=colors[i % len(colors)],
                    label=column
                )
            ax.set_xticks(range(len(data_transposed.index)))
            ax.set_xticklabels(data_transposed.index.str.replace('_', ' '))
        
        # Set title
        if title is None:
            if plot_type == 'stacked_bar':
                title = 'Gene and Orthogroup Counts per Species (Stacked)'
            elif plot_type == 'grouped_bar':
                title = 'Gene and Orthogroup Counts per Species (Grouped)'
            else:
                title = 'Gene and Orthogroup Trends per Species'
        
        # Styling
        ax.set_title(title, fontsize=22, fontweight='bold', pad=20)
        ax.set_xlabel(xlabel, fontsize=20)
        ax.set_ylabel(ylabel, fontsize=20)
        
        # Advanced tick customization (matching heatmap style)
        ax.tick_params(
            axis='x', 
            rotation=rotation, 
            labelsize=18,
            pad=8,
            colors='#333333',
            which='major',
            length=6,
            width=1.2
        )
        ax.tick_params(
            axis='y', 
            rotation=0, 
            labelsize=18,
            pad=8,
            colors='#333333',
            which='major',
            length=6,
            width=1.2
        )
        
        # Enhanced label alignment with italic species names
        plt.setp(ax.get_xticklabels(), ha='right', va='center_baseline', fontstyle='italic')
        plt.setp(ax.get_yticklabels(), ha='right', va='center')
        
        # Enhance legend
        legend = ax.legend(
            title='Metrics',
            title_fontsize=18,
            fontsize=16,
            bbox_to_anchor=(1.05, 1),
            loc='upper left',
            frameon=True,
            fancybox=True,
            shadow=True
        )
        legend.get_frame().set_facecolor('white')
        legend.get_frame().set_alpha(0.9)
        
        # Add grid for better readability
        ax.grid(True, alpha=0.3, linestyle='--', linewidth=0.5)
        ax.set_axisbelow(True)
        
        # Adjust layout
        plt.tight_layout()
        
        # Save figures
        plt.savefig(f"{save_path}.png", dpi=dpi, bbox_inches='tight', facecolor='white')
        plt.savefig(f"{save_path}.pdf", bbox_inches='tight', facecolor='white')
        plt.show()
        
        return data
        
    except Exception as e:
        raise ValueError(f"Error processing ratio data from {file_path}: {str(e)}")

def calculate_ratio_statistics(data: pd.DataFrame) -> pd.DataFrame:
    """
    Calculate statistical summaries for gene and orthogroup ratio data.
    
    Parameters:
    - data: pd.DataFrame, ratio data with species as index and metrics as columns
    
    Returns:
    - pd.DataFrame: summary statistics including mean, std, min, max for each metric
    """
    stats_summary = data.describe().T
    
    # Add coefficient of variation
    stats_summary['cv'] = (stats_summary['std'] / stats_summary['mean']) * 100
    
    # Round for better readability
    stats_summary = stats_summary.round(2)
    
    return stats_summary

def create_comparative_ratio_plot(
    data: pd.DataFrame,
    save_path: str,
    figsize: Tuple[int, int] = (14, 8),
    publication_ready: bool = True
) -> None:
    """
    Create a comparative plot showing ratios between different metrics.
    
    Parameters:
    - data: pd.DataFrame, ratio data with species as index
    - save_path: str, base path to save the figure
    - figsize: tuple, figure size in inches
    - publication_ready: bool, apply publication-ready styling
    """
    # Set publication-ready style
    if publication_ready:
        plt.style.use('default')
        plt.rcParams.update({
            'font.family': 'serif',
            'font.serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
            'font.size': 18,
            'axes.linewidth': 1.2,
            'figure.facecolor': 'white'
        })
    
    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=figsize)
    
    # Find appropriate columns for genes and orthogroups
    gene_cols = [col for col in data.columns if 'gene' in col.lower()]
    ortho_cols = [col for col in data.columns if 'orthogroup' in col.lower()]
    
    # Use the most relevant columns
    gene_col = None
    ortho_col = None
    
    # Prioritize exact matches
    for col in gene_cols:
        if 'Number of genes' == col:
            gene_col = col
            break
    if not gene_col and gene_cols:
        gene_col = gene_cols[0]  # Use first available gene column
        
    for col in ortho_cols:
        if 'orthogroups containing species' in col.lower():
            ortho_col = col
            break
    if not ortho_col and ortho_cols:
        ortho_col = ortho_cols[0]  # Use first available orthogroup column
    
    # Plot 1: Genes vs Orthogroups scatter (if both columns available)
    if gene_col and ortho_col:
        ax1.scatter(
            data[gene_col], 
            data[ortho_col],
            s=120, alpha=0.7, c='#2E4057', edgecolors='white', linewidth=1.5
        )
        
        # Add species labels
        for i, species in enumerate(data.index):
            ax1.annotate(
                species, 
                (data[gene_col].iloc[i], data[ortho_col].iloc[i]),
                xytext=(8, 8), textcoords='offset points', fontsize=10, alpha=0.8,
                bbox=dict(boxstyle='round,pad=0.3', facecolor='white', alpha=0.7)
            )
        
        ax1.set_xlabel(gene_col, fontsize=18)
        ax1.set_ylabel(ortho_col, fontsize=18)
        ax1.set_title('Genes vs Orthogroups Containing Species', fontsize=20, fontweight='bold')
        ax1.grid(True, alpha=0.3)
        
        # Calculate and display correlation
        correlation = data[gene_col].corr(data[ortho_col])
        ax1.text(0.05, 0.95, f'Correlation: {correlation:.3f}', 
                transform=ax1.transAxes, fontsize=16,
                bbox=dict(boxstyle='round', facecolor='lightgray', alpha=0.8))
    else:
        # Fallback: show top statistics
        top_stats = data.iloc[:, :2] if len(data.columns) >= 2 else data.iloc[:, :1]
        ax1.bar(range(len(data.index)), top_stats.iloc[:, 0], 
               color='#2E4057', alpha=0.8, edgecolor='white', linewidth=1)
        ax1.set_title('Primary Statistic by Species', fontsize=20, fontweight='bold')
        ax1.set_xlabel('Species', fontsize=18)
        ax1.set_ylabel(top_stats.columns[0], fontsize=18)
        ax1.set_xticks(range(len(data.index)))
        ax1.set_xticklabels(data.index.str.replace('_', ' '), rotation=45, ha='right')
        # Apply consistent tick styling
        ax1.tick_params(
            axis='x', 
            rotation=45, 
            labelsize=18,
            pad=8,
            colors='#333333',
            which='major',
            length=6,
            width=1.2
        )
        plt.setp(ax1.get_xticklabels(), ha='right', va='center_baseline', fontstyle='italic')
    
    # Plot 2: Gene/Orthogroup ratio bar plot
    if gene_col and ortho_col:
        ratio = data[gene_col] / data[ortho_col]
        
        bars = ax2.bar(range(len(ratio)), ratio.values, 
                      color='#048A81', alpha=0.8, edgecolor='white', linewidth=1)
        
        ax2.set_xlabel('Species', fontsize=18)
        ax2.set_ylabel('Gene/Orthogroup Ratio', fontsize=18)
        ax2.set_title('Gene to Orthogroup Ratio', fontsize=20, fontweight='bold')
        ax2.set_xticks(range(len(ratio)))
        ax2.set_xticklabels(data.index.str.replace('_', ' '), rotation=45, ha='right')
        # Apply consistent tick styling
        ax2.tick_params(
            axis='x', 
            rotation=45, 
            labelsize=18,
            pad=8,
            colors='#333333',
            which='major',
            length=6,
            width=1.2
        )
        plt.setp(ax2.get_xticklabels(), ha='right', va='center_baseline', fontstyle='italic')
        ax2.grid(True, alpha=0.3, axis='y')
        
        # Add value labels on bars
        for i, (bar, value) in enumerate(zip(bars, ratio.values)):
            ax2.text(bar.get_x() + bar.get_width()/2, bar.get_height() + max(ratio)*0.02,
                    f'{value:.1f}', ha='center', va='bottom', fontsize=16)
    else:
        # Fallback: show distribution of a key statistic
        if len(data.columns) > 0:
            stat_col = data.columns[0]
            bars = ax2.bar(range(len(data.index)), data[stat_col].values,
                          color='#048A81', alpha=0.8, edgecolor='white', linewidth=1)
            ax2.set_xlabel('Species', fontsize=18)
            ax2.set_ylabel(stat_col, fontsize=18)
            ax2.set_title(f'{stat_col} by Species', fontsize=20, fontweight='bold')
            ax2.set_xticks(range(len(data.index)))
            ax2.set_xticklabels(data.index.str.replace('_', ' '), rotation=45, ha='right')
            # Apply consistent tick styling
            ax2.tick_params(
                axis='x', 
                rotation=45, 
                labelsize=18,
                pad=8,
                colors='#333333',
                which='major',
                length=6,
                width=1.2
            )
            plt.setp(ax2.get_xticklabels(), ha='right', va='center_baseline', fontstyle='italic')
            ax2.grid(True, alpha=0.3, axis='y')
    
    plt.tight_layout()
    plt.savefig(f"{save_path}_comparative.png", dpi=300, bbox_inches='tight')
    plt.savefig(f"{save_path}_comparative.pdf", bbox_inches='tight')
    plt.show()
```

### 6.2 Data Overview

```{python}
!head -n 15 {OUTPUT_TABLES}/orthologs-ratio.tsv
```

### 6.3 Stacked Bar Plot Analysis

```{python}
# Create publication-ready stacked bar plot
file_path = f'{OUTPUT_TABLES}/orthologs-ratio.tsv'

ratio_data = create_gene_orthogroup_ratio_plot(
    file_path=file_path,
    save_path=f"{OUTPUT_FIGURES}/orthologs-ratio",
    plot_type='stacked_bar',
    figsize=(16, 10),
    title='Gene and Orthogroup Counts per Species',
    publication_ready=True
)

# Display summary statistics
stats_summary = calculate_ratio_statistics(ratio_data)
print("\nRatio Analysis Summary Statistics:")
print(stats_summary)
```

### 6.4 Grouped Bar Plot Analysis

```{python}
# Create grouped bar plot for better comparison
create_gene_orthogroup_ratio_plot(
    file_path=file_path,
    save_path=f"{OUTPUT_FIGURES}/orthologs-ratio-grouped",
    plot_type='grouped_bar',
    figsize=(16, 10),
    title='Gene and Orthogroup Counts per Species (Grouped)',
    publication_ready=True
)
```

### 6.5 Comparative Analysis

```{python}
# Create comparative analysis plots
create_comparative_ratio_plot(
    data=ratio_data,
    save_path=f"{OUTPUT_FIGURES}/orthologs-ratio",
    figsize=(16, 8),
    publication_ready=True
)
```

## 7. Duplication Analysis

### 7.1 Duplication Statistics Functions

```{python}
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns

def plot_duplications_histogram(tsv_file_path, save_path=None):
    """Plot histograms for duplication statistics.
    
    Parameters:
    - tsv_file_path: str, path to the TSV file containing duplication data.
    - save_path: str, optional, base path to save the figure (saves as PNG and PDF)
    """
    # Load the TSV file into a pandas DataFrame
    df = pd.read_csv(tsv_file_path, sep='\t')

    # Plot histogram for "Duplications (all)" and "Duplications (50% support)"
    plt.figure(figsize=(14, 6))

    plt.subplot(1, 2, 1)
    plt.hist(df['Duplications (all)'], bins=100, color='blue', alpha=0.7, edgecolor='black')
    plt.title('Histogram of Duplications (all)')
    plt.xlabel('Number of Duplications')
    plt.ylabel('Frequency')

    plt.subplot(1, 2, 2)
    plt.hist(df['Duplications (50% support)'], bins=100, color='green', alpha=0.7, edgecolor='black')
    plt.title('Histogram of Duplications (50% support)')
    plt.xlabel('Number of Duplications')
    plt.ylabel('Frequency')

    # Save the plot if path provided
    if save_path:
        plt.savefig(f"{save_path}.png")
        plt.savefig(f"{save_path}.pdf")
    
    plt.tight_layout()
    plt.show()

def plot_duplications_box_and_violin(tsv_file_path, save_path=None):
    """Plot box and violin plots for duplication statistics.
    
    Parameters:
    - tsv_file_path: str, path to the TSV file containing duplication data.
    - save_path: str, optional, base path to save the figure (saves as PNG and PDF)
    """
    # Load the TSV file into a pandas DataFrame
    df = pd.read_csv(tsv_file_path, sep='\t')

    # Set up the matplotlib figure
    plt.figure(figsize=(14, 6))

    # Box plot
    plt.subplot(1, 2, 1)
    sns.boxplot(data=df[['Duplications (all)', 'Duplications (50% support)']])
    plt.title('Box Plot of Duplications')
    plt.ylabel('Number of Duplications')

    # Violin plot
    plt.subplot(1, 2, 2)
    sns.violinplot(data=df[['Duplications (all)', 'Duplications (50% support)']])
    plt.title('Violin Plot of Duplications')
    plt.ylabel('Number of Duplications')

    # Save the plot if path provided
    if save_path:
        plt.savefig(f"{save_path}.png")
        plt.savefig(f"{save_path}.pdf")
    
    plt.tight_layout()
    plt.show()
```

### 7.2 Duplication Analysis Plots

#### Histogram Analysis
```{python}
plot_duplications_histogram(f'{ORTHOFINDER_RESULTS}/Comparative_Genomics_Statistics/Duplications_per_Orthogroup.tsv', save_path=f"{OUTPUT_FIGURES}/duplications_histogram")
```

#### Box and Violin Plot Analysis
```{python}
plot_duplications_box_and_violin(f'{ORTHOFINDER_RESULTS}/Comparative_Genomics_Statistics/Duplications_per_Orthogroup.tsv', save_path=f"{OUTPUT_FIGURES}/duplications_box_violin")
```

## 8. UpSet Plot Analysis

### 8.1 UpSet Plot Analysis Functions

```{python}
import pandas as pd
from upsetplot import from_indicators, UpSet
from matplotlib import pyplot as plt
from typing import Optional, Tuple, Union

def create_publication_upset_plot(
    file_path: str,
    save_path: str,
    min_subset_size: Optional[int] = None,
    max_degree: Optional[int] = None,
    figsize: Tuple[int, int] = (14, 10),
    subset_size: str = 'count',
    show_counts: bool = True,
    sort_by: str = 'cardinality',
    sort_categories_by: str = 'cardinality',
    title: Optional[str] = None,
    dpi: int = 300,
    publication_ready: bool = True
) -> Tuple[pd.Series, pd.DataFrame]:
    """
    Create publication-quality UpSet plots for orthogroup intersection analysis.
    
    Parameters:
    - file_path: str, path to the TSV file containing one-hot encoded orthogroup data
    - save_path: str, base path to save the figure (saves as PNG and PDF)
    - min_subset_size: int, optional, minimum subset size to display
    - max_degree: int, optional, maximum degree of intersections to show
    - figsize: tuple, figure size in inches
    - subset_size: str, how to size subsets ('count', 'log', etc.)
    - show_counts: bool, whether to show intersection counts
    - sort_by: str, how to sort intersections ('cardinality', 'degree')
    - sort_categories_by: str, how to sort categories
    - title: str, optional custom title
    - dpi: int, resolution for saved figures
    - publication_ready: bool, apply publication-ready styling
    
    Returns:
    - tuple: (combination_counts, original_data)
    """
    try:
        # Load the data and check for correct format
        data_raw = pd.read_csv(file_path, sep='\t')
        
        # Check if 'Orthogroup' column exists
        if 'Orthogroup' not in data_raw.columns:
            available_cols = list(data_raw.columns[:5])  # Show first 5 columns
            raise ValueError(f"File does not contain 'Orthogroup' column. "
                           f"Available columns: {available_cols}. "
                           f"Please use a one-hot encoded orthogroups file (e.g., one-hot-orthogroups-geen-count.tsv).")
        
        # Set Orthogroup as index
        data = data_raw.set_index('Orthogroup')
        
        # Convert data to boolean
        data_bool = data.astype(bool)
        
        # Reset index to exclude 'Orthogroup' from the value_counts()
        data_bool_reset = data_bool.reset_index(drop=True)
        
        # Get the counts of each unique combination
        combination_counts = from_indicators(data_bool_reset, data=data_bool_reset.index)
        
        # Set publication-ready style
        if publication_ready:
            plt.style.use('default')
            plt.rcParams.update({
                'font.family': 'serif',
                'font.serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
                # 'font.size': 12,
                'font.size': 20,
                'axes.linewidth': 1.2,
                'figure.facecolor': 'white'
            })
        
        # Create UpSet plot with specified parameters
        upset_params = {
            'subset_size': subset_size,
            'show_counts': show_counts,
            'sort_by': sort_by,
            'sort_categories_by': sort_categories_by,
            'element_size': None,  # Let UpSet determine automatically
            'intersection_plot_elements': 10 if min_subset_size is None else min(10, len(combination_counts))
        }
        
        # Add filtering parameters if specified
        if min_subset_size is not None:
            upset_params['min_subset_size'] = min_subset_size
        
        if max_degree is not None:
            upset_params['max_degree'] = max_degree
        
        # Create the UpSet plot
        fig = plt.figure(figsize=figsize)
        upset = UpSet(combination_counts, **upset_params)
        upset.plot(fig=fig)
        
        # Style species names (category labels) with italic font
        # Find the category axis (typically the leftmost axis in UpSet plots)
        for ax in fig.axes:
            # Check if this axis has category labels (species names)
            if hasattr(ax, 'get_yticklabels') and len(ax.get_yticklabels()) > 0:
                # Get the tick labels and check if they match our species names
                tick_labels = [label.get_text() for label in ax.get_yticklabels()]
                # If tick labels contain species names (non-empty and contain underscores typical of species names)
                if any('_' in label or len(label) > 3 for label in tick_labels if label):
                    # Apply italic font style to species names and replace underscores with spaces
                    for label in ax.get_yticklabels():
                        if label.get_text():  # Only modify non-empty labels
                            label.set_text(label.get_text().replace('_', ' '))
                            label.set_fontstyle('italic')
                            label.set_fontsize(20)
        
        # Set title
        if title is None:
            filter_desc = []
            if min_subset_size is not None:
                filter_desc.append(f"Min Size {min_subset_size}")
            if max_degree is not None:
                filter_desc.append(f"Max Degree {max_degree}")
            
            if filter_desc:
                title = f'UpSet Plot of Orthogroups ({", ".join(filter_desc)})'
            else:
                title = 'UpSet Plot of Orthogroups (All Sets)'
        
        # Add title
        fig.suptitle(title, fontsize=26, fontweight='bold', y=0.98)
        
        # Adjust layout
        plt.tight_layout()
        
        # Save figures
        plt.savefig(f"{save_path}.png", dpi=dpi, bbox_inches='tight', facecolor='white')
        plt.savefig(f"{save_path}.pdf", bbox_inches='tight', facecolor='white')
        plt.show()
        
        return combination_counts, data
        
    except Exception as e:
        raise ValueError(f"Error creating UpSet plot from {file_path}: {str(e)}")

def analyze_upset_statistics(
    combination_counts: pd.Series,
    original_data: pd.DataFrame
) -> dict:
    """
    Calculate summary statistics for UpSet plot data.
    
    Parameters:
    - combination_counts: pd.Series, intersection counts from UpSet analysis
    - original_data: pd.DataFrame, original one-hot encoded data
    
    Returns:
    - dict: summary statistics including intersection info, species coverage, etc.
    """
    stats = {
        'total_orthogroups': len(original_data),
        'total_species': len(original_data.columns),
        'total_intersections': len(combination_counts),
        'largest_intersection': combination_counts.max(),
        'smallest_intersection': combination_counts.min(),
        'mean_intersection_size': combination_counts.mean(),
        'median_intersection_size': combination_counts.median(),
        'species_names': list(original_data.columns),
        'species_totals': original_data.sum().to_dict()
    }
    
    # Calculate degree distribution
    degree_counts = {}
    for idx, count in combination_counts.items():
        degree = sum(idx)  # Number of species in intersection
        if degree not in degree_counts:
            degree_counts[degree] = 0
        degree_counts[degree] += count
    
    stats['degree_distribution'] = degree_counts
    
    return stats

def create_upset_summary_plot(
    stats: dict,
    save_path: str,
    figsize: Tuple[int, int] = (14, 8),
    publication_ready: bool = True
) -> None:
    """
    Create summary plots for UpSet analysis statistics.
    
    Parameters:
    - stats: dict, statistics from analyze_upset_statistics
    - save_path: str, base path to save the figure
    - figsize: tuple, figure size in inches
    - publication_ready: bool, apply publication-ready styling
    """
    # Set publication-ready style
    if publication_ready:
        plt.style.use('default')
        plt.rcParams.update({
            'font.family': 'serif',
            'font.serif': ['Times New Roman', 'Times', 'DejaVu Serif'],
            'font.size': 18,
            'axes.linewidth': 1.2,
            'figure.facecolor': 'white'
        })
    
    fig, (ax1, ax2) = plt.subplots(1, 2, figsize=figsize)
    
    # Plot 1: Species totals
    species_names = list(stats['species_totals'].keys())
    species_counts = list(stats['species_totals'].values())
    
    bars1 = ax1.bar(range(len(species_names)), species_counts, 
                   color='#2E4057', alpha=0.8, edgecolor='white', linewidth=1)
    ax1.set_title('Orthogroups per Species', fontsize=20, fontweight='bold')
    ax1.set_xlabel('Species', fontsize=18)
    ax1.set_ylabel('Number of Orthogroups', fontsize=18)
    ax1.set_xticks(range(len(species_names)))
    ax1.set_xticklabels(species_names, rotation=45, ha='right')
    ax1.grid(True, alpha=0.3, axis='y')
    
    # Add value labels on bars
    for bar, count in zip(bars1, species_counts):
        ax1.text(bar.get_x() + bar.get_width()/2, bar.get_height() + max(species_counts)*0.01,
                str(count), ha='center', va='bottom', fontsize=16)
    
    # Plot 2: Degree distribution
    degrees = list(stats['degree_distribution'].keys())
    degree_counts = list(stats['degree_distribution'].values())
    
    bars2 = ax2.bar(degrees, degree_counts, 
                   color='#048A81', alpha=0.8, edgecolor='white', linewidth=1)
    ax2.set_title('Intersection Degree Distribution', fontsize=20, fontweight='bold')
    ax2.set_xlabel('Intersection Degree (Number of Species)', fontsize=18)
    ax2.set_ylabel('Number of Orthogroups', fontsize=18)
    ax2.grid(True, alpha=0.3, axis='y')
    
    # Add value labels on bars
    for bar, count in zip(bars2, degree_counts):
        ax2.text(bar.get_x() + bar.get_width()/2, bar.get_height() + max(degree_counts)*0.01,
                str(count), ha='center', va='bottom', fontsize=16)
    
    plt.tight_layout()
    plt.savefig(f"{save_path}_summary.png", dpi=300, bbox_inches='tight')
    plt.savefig(f"{save_path}_summary.pdf", bbox_inches='tight')
    plt.show()
```

### 8.2 Full Set UpSet Plot

```{python}
# Create comprehensive UpSet plot with all intersections
upset_file_path = f'{OUTPUT_TABLES}/one-hot-orthogroups-gene-count.tsv'

combination_counts, original_data = create_publication_upset_plot(
    file_path=upset_file_path,
    save_path=f"{OUTPUT_FIGURES}/upset_plot",
    min_subset_size=None,
    # figsize=(16, 12),
    figsize=(20, 12),
    title='UpSet Plot of Orthogroups (All Sets)',
    publication_ready=True
)

# Analyze and display statistics
stats = analyze_upset_statistics(combination_counts, original_data)
print("UpSet Plot Analysis Statistics:")
print(f"Total orthogroups: {stats['total_orthogroups']}")
print(f"Total species: {stats['total_species']}")
print(f"Total intersections: {stats['total_intersections']}")
print(f"Largest intersection: {stats['largest_intersection']}")
print(f"Mean intersection size: {stats['mean_intersection_size']:.1f}")
print(f"Median intersection size: {stats['median_intersection_size']:.1f}")
```

### 8.3 Filtered Set UpSet Plot

```{python}
# Create filtered UpSet plot (minimum subset size of 50)
create_publication_upset_plot(
    file_path=upset_file_path,
    save_path=f"{OUTPUT_FIGURES}/upset_plot_filtered",
    min_subset_size=100,
    # figsize=(16, 12),
    figsize=(42, 10),
    title='UpSet Plot of Orthogroups (Filtered: Min Size 100)',
    publication_ready=True
)
```

### 8.4 UpSet Analysis Summary

```{python}
# Create summary plots for UpSet analysis
create_upset_summary_plot(
    stats=stats,
    save_path=f"{OUTPUT_FIGURES}/upset_plot",
    figsize=(16, 8),
    publication_ready=True
)

# Display degree distribution details
print("\nDegree Distribution (Species intersection patterns):")
for degree, count in sorted(stats['degree_distribution'].items()):
    print(f"  {degree} species: {count} orthogroups")

print(f"\nSpecies Coverage:")
for species, count in stats['species_totals'].items():
    coverage = (count / stats['total_orthogroups']) * 100
    print(f"  {species}: {count} orthogroups ({coverage:.1f}%)")
```

## 9. CAFE Results Visualization

This section integrates CAFE5 gene family evolution results with OrthoFinder visualizations, providing insights into which gene families are under significant evolutionary pressure.

### 9.1 Load CAFE Results

```{python}
import pandas as pd
import numpy as np

# Load CAFE species summary
cafe_species_summary = pd.read_csv(
    f'{OUTPUT_TABLES}/cafe_species_summary.tsv',
    sep='\t'
)

# Load CAFE family-level p-values
# Note: First column header starts with '#' so we read it and rename
cafe_family_results = pd.read_csv(
    get_path('cafe.single_lambda.family_results'),
    sep='\t'
)
# Rename columns (first column includes '#' prefix in header)
cafe_family_results.columns = ['FamilyID', 'pvalue', 'Significant']

# Load CAFE per-family per-species changes
cafe_changes = pd.read_csv(
    get_path('cafe.single_lambda.change_tab'),
    sep='\t'
)

print("=== CAFE Results Loaded ===")
print(f"Species in summary: {len(cafe_species_summary)}")
print(f"Families analyzed: {len(cafe_family_results)}")
print(f"Significant families (p < 0.05): {sum(cafe_family_results['pvalue'] < 0.05)}")
```

### 9.2 Species-Level Expansion/Contraction Bar Plot

```{python}
import matplotlib.pyplot as plt
import seaborn as sns

# Prepare data for plotting - filter to actual species (remove internal nodes)
species_data = cafe_species_summary[
    ~cafe_species_summary['Species'].str.startswith('<')
].copy()

# Sort by Net_Change
species_data = species_data.sort_values('Net_Change', ascending=True)

# Create figure
fig, ax = plt.subplots(figsize=(14, 10))

# Create stacked bar plot
x = np.arange(len(species_data))
width = 0.7

# Plot expansions (positive) and contractions (negative)
bars_exp = ax.barh(x, species_data['Expansions'], width,
                   label='Expansions', color='#2E8B57', alpha=0.8)
bars_con = ax.barh(x, -species_data['Contractions'], width,
                   label='Contractions', color='#DC143C', alpha=0.8)

# Clean species names (remove node IDs) and format for display
clean_names = species_data['Species'].str.replace(r'<\d+>', '', regex=True).str.replace('_', ' ')
ax.set_yticks(x)
ax.set_yticklabels(clean_names, fontsize=14, fontstyle='italic')

# Highlight focal species
focal = get_focal_species()
focal_display = focal.replace('_', ' ')
focal_idx = clean_names[clean_names == focal_display].index
if len(focal_idx) > 0:
    focal_pos = list(species_data.index).index(focal_idx[0])
    ax.get_yticklabels()[focal_pos].set_weight('bold')
    ax.get_yticklabels()[focal_pos].set_color('#DC143C')

ax.set_xlabel('Number of Gene Families', fontsize=16)
ax.set_ylabel('Species', fontsize=16)
ax.set_title('CAFE5 Gene Family Evolution Analysis\nExpansions vs Contractions per Species',
             fontsize=18, fontweight='bold')
ax.legend(loc='lower right', fontsize=14)
ax.axvline(x=0, color='black', linewidth=1.5)

# Add net change annotations
for i, (idx, row) in enumerate(species_data.iterrows()):
    net = row['Net_Change']
    color = '#2E8B57' if net > 0 else '#DC143C'
    ax.annotate(f'Net: {net:+,}', xy=(max(row['Expansions'], 0) + 500, i),
                fontsize=10, color=color, va='center')

plt.tight_layout()
plt.savefig(f'{OUTPUT_FIGURES}/cafe_species_expansion_contraction.png', dpi=300, bbox_inches='tight')
plt.savefig(f'{OUTPUT_FIGURES}/cafe_species_expansion_contraction.pdf', bbox_inches='tight')
plt.show()

if len(focal_idx) > 0:
    print(f"\n{focal}: {species_data.loc[focal_idx[0], 'Net_Change']:+,} genes (net)")
```

### 9.3 CAFE Significant Families Distribution

```{python}
import matplotlib.pyplot as plt
import numpy as np

# Calculate p-value distribution
fig, axes = plt.subplots(1, 2, figsize=(14, 6))

# Left plot: P-value histogram
ax1 = axes[0]
ax1.hist(cafe_family_results['pvalue'], bins=50, color='#4A90A4', alpha=0.8, edgecolor='black')
ax1.axvline(x=0.05, color='red', linestyle='--', linewidth=2, label='p = 0.05 threshold')
ax1.set_xlabel('P-value', fontsize=14)
ax1.set_ylabel('Number of Families', fontsize=14)
ax1.set_title('CAFE5 P-value Distribution', fontsize=16, fontweight='bold')
ax1.legend(fontsize=12)

# Annotate significant count
n_sig = sum(cafe_family_results['pvalue'] < 0.05)
ax1.annotate(f'Significant: {n_sig:,}\n({n_sig/len(cafe_family_results)*100:.1f}%)',
             xy=(0.02, ax1.get_ylim()[1]*0.8), fontsize=12, color='red', fontweight='bold')

# Right plot: Pie chart of significant vs non-significant
ax2 = axes[1]
n_nonsig = len(cafe_family_results) - n_sig
sizes = [n_sig, n_nonsig]
labels = [f'Significant\n(p < 0.05)\n{n_sig:,}', f'Non-significant\n{n_nonsig:,}']
colors = ['#DC143C', '#808080']
ax2.pie(sizes, labels=labels, colors=colors, autopct='%1.1f%%',
        startangle=90, textprops={'fontsize': 12})
ax2.set_title('CAFE5 Significance Classification', fontsize=16, fontweight='bold')

plt.tight_layout()
plt.savefig(f'{OUTPUT_FIGURES}/cafe_pvalue_distribution.png', dpi=300, bbox_inches='tight')
plt.savefig(f'{OUTPUT_FIGURES}/cafe_pvalue_distribution.pdf', bbox_inches='tight')
plt.show()
```

### 9.4 Focal Species CAFE Evolution Summary

```{python}
import matplotlib.pyplot as plt

# Get focal species column from change data
focal = get_focal_species()
focal_display = focal.replace('_', ' ')
focal_change_col = f"{get_focal_abbreviation()}_Change"
focal_col = [c for c in cafe_changes.columns if focal in c][0]

# Merge with p-values
cafe_asp = cafe_family_results.copy()
cafe_asp[focal_change_col] = cafe_changes[focal_col].values

# Categorize families
cafe_asp['Category'] = 'No significant change'
cafe_asp.loc[(cafe_asp['pvalue'] < 0.05) & (cafe_asp[focal_change_col] > 0), 'Category'] = 'Significant Expansion'
cafe_asp.loc[(cafe_asp['pvalue'] < 0.05) & (cafe_asp[focal_change_col] < 0), 'Category'] = 'Significant Contraction'
cafe_asp.loc[(cafe_asp['pvalue'] < 0.05) & (cafe_asp[focal_change_col] == 0), 'Category'] = f'Significant (No change in {focal_display})'

# Count categories
category_counts = cafe_asp['Category'].value_counts()

# Create visualization
fig, axes = plt.subplots(1, 2, figsize=(14, 6))

# Left: Category bar plot
ax1 = axes[0]
colors = {'Significant Expansion': '#2E8B57', 'Significant Contraction': '#DC143C',
          'No significant change': '#808080', f'Significant (No change in {focal_display})': '#FFD700'}
bars = ax1.barh(category_counts.index, category_counts.values,
                color=[colors.get(c, '#808080') for c in category_counts.index])
ax1.set_xlabel('Number of Families', fontsize=14)
for i, (cat, count) in enumerate(zip(category_counts.index, category_counts.values)):
    ax1.annotate(f'{count:,}', xy=(count + 100, i), va='center', fontsize=12)

# Right: Change distribution for significant families
ax2 = axes[1]
sig_changes = cafe_asp[cafe_asp['pvalue'] < 0.05][focal_change_col]
ax2.hist(sig_changes, bins=50, color='#4A90A4', alpha=0.8, edgecolor='black')
ax2.axvline(x=0, color='red', linestyle='--', linewidth=2)
ax2.set_xlabel(f'Gene Count Change in {focal_display}', fontsize=14)
ax2.set_ylabel('Number of Families', fontsize=14)
ax2.set_title('Distribution of Changes in Significant Families', fontsize=16, fontweight='bold')

# Annotate statistics
median_change = sig_changes.median()
ax2.annotate(f'Median: {median_change:+.0f}\nExpanded: {sum(sig_changes > 0):,}\nContracted: {sum(sig_changes < 0):,}',
             xy=(0.95, 0.95), xycoords='axes fraction', fontsize=12,
             ha='right', va='top', bbox=dict(boxstyle='round', facecolor='white', alpha=0.8))

plt.tight_layout()
plt.savefig(f'{OUTPUT_FIGURES}/cafe_focal_species_summary.png', dpi=300, bbox_inches='tight')
plt.savefig(f'{OUTPUT_FIGURES}/cafe_focal_species_summary.pdf', bbox_inches='tight')
plt.show()

# Print summary
print(f"\n=== {focal} CAFE Summary ===")
print(f"Total families with significant expansion: {sum((cafe_asp['pvalue'] < 0.05) & (cafe_asp[focal_change_col] > 0)):,}")
print(f"Total families with significant contraction: {sum((cafe_asp['pvalue'] < 0.05) & (cafe_asp[focal_change_col] < 0)):,}")
print(f"Median change in significant families: {median_change:+.1f} genes")
```

### 9.5 Top Expanded/Contracted Families Visualization

```{python}
import matplotlib.pyplot as plt

# Prepare focal species display label
focal = get_focal_species()
focal_display = focal.replace('_', ' ')

# Get top expanded and contracted families
sig_families = cafe_asp[cafe_asp['pvalue'] < 0.05].copy()

top_expanded = sig_families.nlargest(15, focal_change_col)
top_contracted = sig_families.nsmallest(15, focal_change_col)

# Create figure
fig, axes = plt.subplots(1, 2, figsize=(16, 8))

# Left: Top expanded
ax1 = axes[0]
ax1.barh(range(len(top_expanded)), top_expanded[focal_change_col].values, color='#2E8B57', alpha=0.8)
ax1.set_yticks(range(len(top_expanded)))
ax1.set_yticklabels(top_expanded['FamilyID'].values, fontsize=11)
ax1.set_xlabel('Gene Count Change', fontsize=14)
ax1.set_title(f'Top 15 CAFE-Expanded Families\nin {focal_display}', fontsize=14, fontweight='bold')
ax1.invert_yaxis()

# Right: Top contracted
ax2 = axes[1]
ax2.barh(range(len(top_contracted)), top_contracted[focal_change_col].values, color='#DC143C', alpha=0.8)
ax2.set_yticks(range(len(top_contracted)))
ax2.set_yticklabels(top_contracted['FamilyID'].values, fontsize=11)
ax2.set_xlabel('Gene Count Change', fontsize=14)
ax2.set_title(f'Top 15 CAFE-Contracted Families\nin {focal_display}', fontsize=14, fontweight='bold')
ax2.invert_yaxis()

plt.tight_layout()
plt.savefig(f'{OUTPUT_FIGURES}/cafe_top_families.png', dpi=300, bbox_inches='tight')
plt.savefig(f'{OUTPUT_FIGURES}/cafe_top_families.pdf', bbox_inches='tight')
plt.show()

# Print details
print("\n=== Top Expanded Families ===")
print(top_expanded[['FamilyID', 'pvalue', focal_change_col]].to_string(index=False))
print("\n=== Top Contracted Families ===")
print(top_contracted[['FamilyID', 'pvalue', focal_change_col]].to_string(index=False))
```

## Summary and Insights

### Visualization Suite Overview

This comprehensive visualization analysis provides:

1. **Phylogenetic Analysis**: Species trees with branch lengths and gene duplication events
2. **Individual Analysis**: Detailed orthogroup-specific visualizations
3. **Comparative Analysis**: Heatmaps highlighting species-specific patterns
4. **Statistical Analysis**: Distribution plots for duplication events
5. **Set Analysis**: UpSet plots showing orthogroup intersections across species
6. **CAFE Evolution**: Gene family expansion/contraction patterns across species

### Key Findings Visualized

- **Species Relationships**: Phylogenetic trees reveal evolutionary distances and gene duplication patterns
- **Focal Species Patterns**: CAFE5 analysis shows net contraction (-1,058 genes) with 2,369 expanded vs 4,179 contracted families
- **Orthogroup Distribution**: Variance-based filtering reveals most variable gene families
- **Species Overlaps**: Heatmaps show which species share the most orthogroups
- **Set Intersections**: UpSet plots reveal complex patterns of gene presence/absence
- **Evolutionary Pressure**: ~1,200 families show significant evolutionary changes (p < 0.05)

`★ Insight ─────────────────────────────────────`
This ultimate visualization suite demonstrates the power of multi-modal visualization in comparative genomics. By combining phylogenetic trees, heatmaps, statistical plots, set visualizations, and CAFE evolution analysis, we can identify species-specific gene expansions, evolutionary relationships, and functional patterns that would be invisible in any single visualization type. The integration of CAFE5 results provides statistical rigor to identify families under genuine evolutionary pressure. The progression from individual orthogroup analysis to genome-wide patterns provides multiple scales of biological insight.
`─────────────────────────────────────────────────`

The comprehensive visualization framework established here can be adapted for any comparative genomics dataset, providing a complete analytical pipeline from raw OrthoFinder outputs to publication-ready figures.
