---
title: "Enrichment Analysis for Ka/Ks Selection Categories"
subtitle: "GO and KEGG enrichment for genes under positive and purifying selection"
jupyter:
  kernelspec:
    display_name: clusterprofiler R
    language: R
    name: ir_clusterprofiler_4100_env
---

## Overview

This notebook performs comprehensive GO and KO enrichment analysis for orthogroups identified by **Ka/Ks (dN/dS) selection analysis**. Unlike the CAFE-based enrichment which focuses on gene family size changes, this analysis focuses on **sequence-level selection pressure**.

### Analysis Categories (Selection-Based)

- **Positive Selection (tentative)**: Orthogroups where max omega > 1 in pairwise comparisons (yn00)
- **Relaxed Purifying**: Orthogroups with 0.5 < mean omega <= 1.0
- **Moderate Purifying**: Orthogroups with 0.1 < mean omega <= 0.5
- **Strong Purifying**: Orthogroups with mean omega <= 0.1 (essential genes)

### Why Selection-Based Enrichment?

1. **Complementary to CAFE**: CAFE detects gene family SIZE changes; Ka/Ks detects SEQUENCE changes
2. **Adaptive evolution**: Positive selection (omega > 1) indicates adaptive amino acid changes
3. **Essential genes**: Strong purifying selection identifies functionally critical genes
4. **Relaxed constraint**: May indicate niche-specific adaptation or pseudogenization

### Ka/Ks Results Summary for *A. laticeps*

- Total orthogroups analyzed: ~32,112
- yn00 successful: ~24,741
- Positive selection (tentative): ~14,920
- Relaxed purifying: ~6,132
- Strong purifying: ~2,162

## Environment Setup

```{r}
#| vscode: {languageId: r}
R.home()
```

## Methods

### Load Required Libraries

```{r}
#| vscode: {languageId: r}
library(dplyr)
library(stringr)
library(jsonlite)
library(purrr)
library(RCurl)
library(GO.db)
library(AnnotationForge)
library(clusterProfiler)
library(openxlsx)
library(ggplot2)
library(gridExtra)
library(R6)
library(enrichplot)
```

### Load Configuration

```{r}
#| vscode: {languageId: r}
# Load centralized path configuration
source("./Notebooks/config/load_config.R")
load_paths(config_path = "/home/jovyan/Notebooks/config/paths-set2.yaml", force_reload = TRUE)

# Display configuration summary
cat(sprintf("Configuration loaded: OrthoFinder Results_%s\n", get_version()))
cat(sprintf("OrthoFinder results: %s\n", get_path("orthofinder.results_dir")))

# Define commonly used paths for this notebook
ORTHOGROUPS_TSV <- get_path("orthofinder.orthogroups_tsv")
ORTHOGROUPS_COUNT <- get_path("orthofinder.orthogroups_count")
ORG_DB_PATH <- get_path("database.org_db")
OUTPUT_FIGURES <- get_path("outputs.figures")
OUTPUT_TABLES <- get_path("outputs.tables")
GO_ENRICHMENT_DIR <- get_path("outputs.go_enrichment")
KO_ENRICHMENT_DIR <- get_path("outputs.ko_enrichment")
EGGNOG_GENE2PATHWAY <- get_path("eggnog.gene2pathway")
EGGNOG_KEGG_INFO <- get_path("eggnog.kegg_info")
GENOMEWIDE_SUMMARY <- get_path("outputs.genomewide_selection.summary")

# Ka/Ks results paths
KAKS_MASTER <- paste0(GENOMEWIDE_SUMMARY, "/genome_wide_selection_master.tsv")
KAKS_TENTATIVE <- paste0(GENOMEWIDE_SUMMARY, "/positive_selection_tentative.tsv")
KAKS_SIGNIFICANT <- paste0(GENOMEWIDE_SUMMARY, "/positive_selection_significant.tsv")
```

### Initialize Custom Annotation Database

```{r}
#| vscode: {languageId: r}
# Force reinstall OrgDb to ensure the latest version is loaded
orgdb_pkg <- ORG_DB_PATH

# Check package exists
if (!dir.exists(orgdb_pkg)) {
  stop("OrgDb package not found at: ", orgdb_pkg)
}
cat("Package found at:", orgdb_pkg, "\n")
cat("Package contents:", paste(list.files(orgdb_pkg), collapse = ", "), "\n\n")

# Get the correct library path for this R environment
target_lib <- .libPaths()[1]
cat("Target library:", target_lib, "\n")
cat("AnnotationDbi available:", "AnnotationDbi" %in% installed.packages()[,"Package"], "\n\n")

orgdb_name <- get_orgdb_package_name()

# Remove old installation if present (from the correct library)
if (orgdb_name %in% installed.packages()[,"Package"]) {
  cat("Removing cached OrgDb installation...\n")
  remove.packages(orgdb_name, lib = target_lib)
}

# Install to the correct library using R CMD INSTALL with full path to conda R
cat("Installing OrgDb package...\n")
cat("========================================\n")

# Get the R executable path from this session
r_exe <- file.path(R.home("bin"), "R")
cat("Using R executable:", r_exe, "\n\n")

# Build the install command with explicit library path
install_cmd <- sprintf(
  "'%s' CMD INSTALL --library='%s' --no-test-load '%s' 2>&1",
  r_exe, target_lib, orgdb_pkg
)
cat("Command:", install_cmd, "\n\n")

# Run installation and capture output
install_output <- system(install_cmd, intern = TRUE)
cat(paste(install_output, collapse = "\n"), "\n")

cat("========================================\n")

# Verify installation succeeded
if (!orgdb_name %in% installed.packages(lib.loc = target_lib)[,"Package"]) {
  stop("Installation failed - package not found in library after install")
}
cat("Package installed successfully\n\n")

# Load the library
cat("Loading", orgdb_name, "...\n")
library(orgdb_name, character.only = TRUE)
orgdb_obj <- get(orgdb_name)

# Verify gene ID format (should be BRAKER3 format: g1.t1, g2.t1, ...)
cat("\n=== OrgDb Gene ID Verification ===\n")
sample_ids <- head(keys(orgdb_obj, keytype = "GID"), 10)
cat("Sample gene IDs:\n")
print(sample_ids)

# Check for expected format
if (any(grepl("^g\\d+\\.t\\d+$", sample_ids))) {
  cat("\nGene IDs are in BRAKER3 format (g*.t*) - CORRECT\n")
} else if (any(grepl("^evm\\.model\\.", sample_ids))) {
  cat("\nWARNING: Gene IDs are in EVM format - OLD DATABASE LOADED!\n")
  cat("  Please check that the OrgDb directory contains the new database.\n")
} else {
  cat("\n? Unknown gene ID format - please verify manually\n")
}

keytypes(orgdb_obj)
GO_database <- orgdb_name
```

## Load Ka/Ks Selection Results

### Data Sources

- **Ka/Ks Master Results**: `./Outputs/GenomewideSelection/summary/genome_wide_selection_master.tsv`
- **Tentative Positive Selection**: `./Outputs/GenomewideSelection/summary/positive_selection_tentative.tsv`
- **Orthogroups Information**: Configured via `orthofinder.orthogroups_tsv`

### Load Ka/Ks Analysis Results

```{r}
#| vscode: {languageId: r}
# Load Ka/Ks master results
kaks_master <- read.table(
  KAKS_MASTER,
  header = TRUE, sep = "\t", stringsAsFactors = FALSE,
  quote = "", fill = TRUE
)

cat("=== Ka/Ks Analysis Results Loaded ===\n")
cat(paste("Total orthogroups:", nrow(kaks_master), "\n"))
cat(paste("Columns:", paste(colnames(kaks_master)[1:10], collapse = ", "), "...\n"))

# Check status distribution
cat("\nyn00 Status distribution:\n")
print(table(kaks_master$status))
```

## Selection Category Classification

### Positive Selection (Tentative)

Orthogroups where **max omega > 1** in pairwise comparisons, indicating at least one species pair shows evidence of positive selection.

```{r}
#| vscode: {languageId: r}
# Filter to successful yn00 results
kaks_success <- kaks_master[kaks_master$status == "success", ]
cat(paste("Successful yn00 analyses:", nrow(kaks_success), "\n\n"))

# Positive selection (tentative) - max_omega > 1
positive_tentative_OG <- kaks_success$Orthogroup[
  !is.na(kaks_success$max_omega) &
  kaks_success$max_omega > 1
]

cat("=== POSITIVE SELECTION (TENTATIVE) ===\n")
cat(paste("Number of orthogroups with max omega > 1:", length(positive_tentative_OG), "\n"))

# Show distribution of max_omega for positive selection
if (length(positive_tentative_OG) > 0) {
  positive_data <- kaks_success[kaks_success$Orthogroup %in% positive_tentative_OG, ]
  cat(paste("Mean max_omega:", round(mean(positive_data$max_omega, na.rm = TRUE), 3), "\n"))
  cat(paste("Median max_omega:", round(median(positive_data$max_omega, na.rm = TRUE), 3), "\n"))
  cat(paste("Max max_omega:", round(max(positive_data$max_omega, na.rm = TRUE), 3), "\n"))
}
```

### Relaxed Purifying Selection

Orthogroups with **0.5 < mean omega <= 1.0**, indicating reduced functional constraint.

```{r}
#| vscode: {languageId: r}
# Relaxed purifying - 0.5 < mean_omega <= 1.0
relaxed_OG <- kaks_success$Orthogroup[
  !is.na(kaks_success$mean_omega) &
  kaks_success$mean_omega > 0.5 &
  kaks_success$mean_omega <= 1.0
]

cat("=== RELAXED PURIFYING SELECTION ===\n")
cat(paste("Number of orthogroups:", length(relaxed_OG), "\n"))

if (length(relaxed_OG) > 0) {
  relaxed_data <- kaks_success[kaks_success$Orthogroup %in% relaxed_OG, ]
  cat(paste("Mean omega:", round(mean(relaxed_data$mean_omega, na.rm = TRUE), 3), "\n"))
}
```

### Moderate Purifying Selection

Orthogroups with **0.1 < mean omega <= 0.5**, indicating moderate functional constraint.

```{r}
#| vscode: {languageId: r}
# Moderate purifying - 0.1 < mean_omega <= 0.5
moderate_purifying_OG <- kaks_success$Orthogroup[
  !is.na(kaks_success$mean_omega) &
  kaks_success$mean_omega > 0.1 &
  kaks_success$mean_omega <= 0.5
]

cat("=== MODERATE PURIFYING SELECTION ===\n")
cat(paste("Number of orthogroups:", length(moderate_purifying_OG), "\n"))

if (length(moderate_purifying_OG) > 0) {
  moderate_data <- kaks_success[kaks_success$Orthogroup %in% moderate_purifying_OG, ]
  cat(paste("Mean omega:", round(mean(moderate_data$mean_omega, na.rm = TRUE), 3), "\n"))
}
```

### Strong Purifying Selection

Orthogroups with **mean omega <= 0.1**, indicating strong functional constraint (essential genes).

```{r}
#| vscode: {languageId: r}
# Strong purifying - mean_omega <= 0.1
strong_purifying_OG <- kaks_success$Orthogroup[
  !is.na(kaks_success$mean_omega) &
  kaks_success$mean_omega <= 0.1
]

cat("=== STRONG PURIFYING SELECTION ===\n")
cat(paste("Number of orthogroups:", length(strong_purifying_OG), "\n"))

if (length(strong_purifying_OG) > 0) {
  strong_data <- kaks_success[kaks_success$Orthogroup %in% strong_purifying_OG, ]
  cat(paste("Mean omega:", round(mean(strong_data$mean_omega, na.rm = TRUE), 3), "\n"))
}
```

### Summary of Selection Categories

```{r}
#| vscode: {languageId: r}
cat("\n========================================\n")
cat("    SELECTION CATEGORY SUMMARY\n")
cat("========================================\n\n")
cat(paste("Positive selection (tentative):  ", length(positive_tentative_OG), "\n"))
cat(paste("Relaxed purifying:               ", length(relaxed_OG), "\n"))
cat(paste("Moderate purifying:              ", length(moderate_purifying_OG), "\n"))
cat(paste("Strong purifying:                ", length(strong_purifying_OG), "\n"))
cat(paste("----------------------------------------\n"))
cat(paste("Total successful analyses:       ", nrow(kaks_success), "\n"))
cat("\nNote: Categories may overlap (a gene can have max_omega > 1\n")
cat("but mean_omega < 0.5 if selection is heterogeneous across species)\n")
```

#### Out[1]: ✓ Done 0.0s
```
========================================

    SELECTION CATEGORY SUMMARY

========================================


Positive selection (tentative):   14920 

Relaxed purifying:                6132 

Moderate purifying:               14558 

Strong purifying:                 2162 

----------------------------------------

Total successful analyses:        24741 


Note: Categories may overlap (a gene can have max_omega > 1

but mean_omega < 0.5 if selection is heterogeneous across species)
```


## GO Enrichment Analysis

### GO Enrichment Function

This function performs Gene Ontology enrichment analysis for specific orthogroups.

```{r}
#| vscode: {languageId: r}
enrich_orthogroups <- function(input_path, ont, output_path, target_ogs, org_db,
                              key_type="GID", plot_dir=NULL,
                              p_cutoff=0.01, q_cutoff=0.05,
                              # Publication-quality plot parameters
                              show_category=10, plot_width=12, plot_height=8,
                              font_size=18, title_size=22, axis_title_size=20,
                              color_palette="viridis", output_formats=c("png", "pdf"),
                              dpi=300, theme_style="publication") {

  # Resolve default plot_dir from config
  if (is.null(plot_dir)) {
    plot_dir <- paste0(GO_ENRICHMENT_DIR, "/Plots")
  }

  # Load additional libraries for enhanced plotting
  if (!requireNamespace("viridis", quietly = TRUE)) {
    install.packages("viridis")
  }
  if (!requireNamespace("RColorBrewer", quietly = TRUE)) {
    install.packages("RColorBrewer")
  }
  library(viridis)
  library(RColorBrewer)

  # Read OrthoFinder results
  orthogroups_df <- read.delim(input_path, header = TRUE)

  # Extract gene IDs from target orthogroups
  genes_of_interest <- c()
  filtered_rows <- orthogroups_df[orthogroups_df$Orthogroup %in% target_ogs, ]

  for (row in 1:nrow(filtered_rows)) {
    selected_row <- filtered_rows[row, ]

    for (colname in names(selected_row)) {
      if (colname != "Orthogroup") {
        genes_split <- unlist(strsplit(as.character(selected_row[[colname]]), ", "))
        genes_of_interest <- c(genes_of_interest, genes_split)
      }
    }
  }

  # Remove duplicates and empty strings
  genes_of_interest <- unique(genes_of_interest[genes_of_interest != ""])

  if (length(genes_of_interest) == 0) {
    warning("No genes found for the specified orthogroups")
    return(NULL)
  }

  # Perform GO enrichment analysis
  tryCatch({
    ego_result <- enrichGO(
      gene          = genes_of_interest,
      OrgDb         = org_db,
      keyType       = key_type,
      ont           = ont,
      pAdjustMethod = "BH",
      pvalueCutoff  = p_cutoff,
      qvalueCutoff  = q_cutoff
    )

    # Create output directory
    if (!dir.exists(plot_dir)) {
      dir.create(plot_dir, recursive = TRUE)
    }

    # Save results table
    write.csv(as.data.frame(ego_result), output_path, row.names=FALSE)

    # Generate and save visualizations if results exist
    if (nrow(ego_result) > 0) {

      # Define publication-quality theme
      publication_theme <- function() {
        theme_minimal() +
        theme(
          text = element_text(size = font_size, family = "sans"),
          plot.title = element_text(size = title_size, face = "bold", hjust = 0.5, margin = margin(b = 20)),
          axis.title.x = element_text(size = axis_title_size, face = "bold", margin = margin(t = 15)),
          axis.title.y = element_text(size = axis_title_size, face = "bold", margin = margin(r = 15)),
          axis.text.x = element_text(size = font_size, color = "black"),
          axis.text.y = element_text(size = font_size, color = "black", hjust = 1),
          legend.title = element_text(size = font_size + 2, face = "bold"),
          legend.text = element_text(size = font_size),
          legend.position = "right",
          legend.margin = margin(l = 20),
          legend.key.size = unit(0.8, "cm"),
          panel.grid.major = element_line(color = "grey90", size = 0.5),
          panel.grid.minor = element_line(color = "grey95", size = 0.3),
          panel.border = element_rect(color = "black", fill = NA, size = 0.8),
          plot.margin = margin(20, 20, 20, 20),
          strip.text = element_text(size = font_size, face = "bold", margin = margin(5, 5, 5, 5))
        )
      }

      # Define color palette
      get_color_palette <- function(palette_name, n_colors) {
        if (palette_name == "viridis") {
          return(viridis::viridis(n_colors))
        } else if (palette_name == "plasma") {
          return(viridis::plasma(n_colors))
        } else if (palette_name %in% rownames(RColorBrewer::brewer.pal.info)) {
          max_colors <- RColorBrewer::brewer.pal.info[palette_name, "maxcolors"]
          if (n_colors <= max_colors) {
            return(RColorBrewer::brewer.pal(max(3, n_colors), palette_name))
          } else {
            return(colorRampPalette(RColorBrewer::brewer.pal(max_colors, palette_name))(n_colors))
          }
        } else {
          return(RColorBrewer::brewer.pal(max(3, min(n_colors, 8)), "Set2"))
        }
      }

      # Enhanced Bar plot
      ont_full_name <- switch(ont,
                             "BP" = "Biological Process",
                             "MF" = "Molecular Function",
                             "CC" = "Cellular Component",
                             ont)

      barplot_plot <- barplot(ego_result, showCategory = show_category, x = "Count") +
        scale_fill_gradientn(colors = get_color_palette(color_palette, 100),
                           name = "-log10(p.adjust)",
                           guide = guide_colorbar(title.position = "top")) +
        labs(
          title = paste("GO", ont_full_name, "Enrichment Analysis"),
          x = "Gene Count",
          y = "GO Terms",
          subtitle = paste("Top", min(show_category, nrow(ego_result)), "enriched terms")
        ) +
        publication_theme() +
        theme(
          axis.text.y = element_text(angle = 0, hjust = 1, size = font_size),
          plot.subtitle = element_text(size = font_size + 2, hjust = 0.5, margin = margin(b = 15))
        )

      # Enhanced Dot plot
      dotplot_plot <- dotplot(ego_result, showCategory = show_category, x = "GeneRatio") +
        scale_color_gradientn(colors = get_color_palette(color_palette, 100),
                            name = "-log10(p.adjust)",
                            guide = guide_colorbar(title.position = "top")) +
        scale_size_continuous(name = "Gene Count", range = c(3, 8)) +
        labs(
          title = paste("GO", ont_full_name, "Enrichment Analysis"),
          x = "Gene Ratio",
          y = "GO Terms",
          subtitle = paste("Top", min(show_category, nrow(ego_result)), "enriched terms")
        ) +
        publication_theme() +
        theme(
          axis.text.y = element_text(angle = 0, hjust = 1, size = font_size),
          plot.subtitle = element_text(size = font_size + 2, hjust = 0.5, margin = margin(b = 15)),
          legend.box = "horizontal",
          legend.margin = margin(t = 10)
        )

      # Save plots in multiple formats
      for (format in output_formats) {
        barplot_file <- file.path(plot_dir, paste0("GO_enrichment_barplot.", format))
        if (format == "pdf") {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height,
                 device = "pdf", useDingbats = FALSE)
        } else {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height,
                 dpi = dpi, device = format)
        }

        dotplot_file <- file.path(plot_dir, paste0("GO_enrichment_dotplot.", format))
        if (format == "pdf") {
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height,
                 device = "pdf", useDingbats = FALSE)
        } else {
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height,
                 dpi = dpi, device = format)
        }
      }

      print(paste("GO enrichment completed. Results saved to:", output_path))
      print(paste("Publication-quality plots saved to:", plot_dir))
    } else {
      print("No significant GO terms found for the given cutoffs")
    }

    return(ego_result)

  }, error = function(e) {
    print(paste("Error in GO enrichment:", e$message))
    return(NULL)
  })
}
```

### GO Enrichment: Positive Selection (Tentative)

Genes showing evidence of positive selection (omega > 1) in at least one pairwise comparison. Enrichment in these families suggests functional categories under **adaptive evolution**.

#### Biological Process (BP)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = ORTHOGROUPS_TSV,
  ont = "BP",
  output_path = paste0(OUTPUT_TABLES, "/GO_enrichment_KaKs_positive_tentative_BP_results.csv"),
  target_ogs = positive_tentative_OG,
  org_db = orgdb_obj,
  color_palette="Set2",
  plot_height = 12,
  plot_width = 12,
  p_cutoff = 0.05,
  q_cutoff = 0.1,
  plot_dir = paste0(OUTPUT_FIGURES, "/GO_Enrichment/KaKs_Positive_Tentative_BP_Plots")
)
```

#### Molecular Function (MF)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = ORTHOGROUPS_TSV,
  ont = "MF",
  output_path = paste0(OUTPUT_TABLES, "/GO_enrichment_KaKs_positive_tentative_MF_results.csv"),
  target_ogs = positive_tentative_OG,
  org_db = orgdb_obj,
  color_palette="Set2",
  plot_height = 12,
  plot_width = 12,
  p_cutoff = 0.05,
  q_cutoff = 0.1,
  plot_dir = paste0(OUTPUT_FIGURES, "/GO_Enrichment/KaKs_Positive_Tentative_MF_Plots")
)
```

#### Cellular Component (CC)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = ORTHOGROUPS_TSV,
  ont = "CC",
  output_path = paste0(OUTPUT_TABLES, "/GO_enrichment_KaKs_positive_tentative_CC_results.csv"),
  target_ogs = positive_tentative_OG,
  org_db = orgdb_obj,
  color_palette="Set2",
  plot_height = 12,
  plot_width = 12,
  p_cutoff = 0.05,
  q_cutoff = 0.1,
  plot_dir = paste0(OUTPUT_FIGURES, "/GO_Enrichment/KaKs_Positive_Tentative_CC_Plots")
)
```

### GO Enrichment: Relaxed Purifying Selection

Genes with reduced functional constraint (0.5 < omega <= 1.0). May indicate **niche-specific adaptation** or genes under relaxed selection pressure.

#### Biological Process (BP)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = ORTHOGROUPS_TSV,
  ont = "BP",
  output_path = paste0(OUTPUT_TABLES, "/GO_enrichment_KaKs_relaxed_purifying_BP_results.csv"),
  target_ogs = relaxed_OG,
  org_db = orgdb_obj,
  color_palette="Set2",
  plot_height = 12,
  plot_width = 12,
  plot_dir = paste0(OUTPUT_FIGURES, "/GO_Enrichment/KaKs_Relaxed_Purifying_BP_Plots")
)
```

#### Molecular Function (MF)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = ORTHOGROUPS_TSV,
  ont = "MF",
  output_path = paste0(OUTPUT_TABLES, "/GO_enrichment_KaKs_relaxed_purifying_MF_results.csv"),
  target_ogs = relaxed_OG,
  org_db = orgdb_obj,
  color_palette="Set2",
  plot_dir = paste0(OUTPUT_FIGURES, "/GO_Enrichment/KaKs_Relaxed_Purifying_MF_Plots")
)
```

#### Cellular Component (CC)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = ORTHOGROUPS_TSV,
  ont = "CC",
  output_path = paste0(OUTPUT_TABLES, "/GO_enrichment_KaKs_relaxed_purifying_CC_results.csv"),
  target_ogs = relaxed_OG,
  org_db = orgdb_obj,
  color_palette="Set2",
  plot_dir = paste0(OUTPUT_FIGURES, "/GO_Enrichment/KaKs_Relaxed_Purifying_CC_Plots")
)
```

### GO Enrichment: Strong Purifying Selection

Genes under strong functional constraint (omega <= 0.1). These are likely **essential genes** critical for basic cellular functions.

#### Biological Process (BP)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = ORTHOGROUPS_TSV,
  ont = "BP",
  output_path = paste0(OUTPUT_TABLES, "/GO_enrichment_KaKs_strong_purifying_BP_results.csv"),
  target_ogs = strong_purifying_OG,
  org_db = orgdb_obj,
  color_palette="Set2",
  plot_height = 12,
  plot_width = 12,
  plot_dir = paste0(OUTPUT_FIGURES, "/GO_Enrichment/KaKs_Strong_Purifying_BP_Plots")
)
```

#### Molecular Function (MF)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = ORTHOGROUPS_TSV,
  ont = "MF",
  output_path = paste0(OUTPUT_TABLES, "/GO_enrichment_KaKs_strong_purifying_MF_results.csv"),
  target_ogs = strong_purifying_OG,
  org_db = orgdb_obj,
  color_palette="Set2",
  plot_dir = paste0(OUTPUT_FIGURES, "/GO_Enrichment/KaKs_Strong_Purifying_MF_Plots")
)
```

#### Cellular Component (CC)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = ORTHOGROUPS_TSV,
  ont = "CC",
  output_path = paste0(OUTPUT_TABLES, "/GO_enrichment_KaKs_strong_purifying_CC_results.csv"),
  target_ogs = strong_purifying_OG,
  org_db = orgdb_obj,
  color_palette="Set2",
  plot_dir = paste0(OUTPUT_FIGURES, "/GO_Enrichment/KaKs_Strong_Purifying_CC_Plots")
)
```

## KEGG Pathway (KO) Enrichment Analysis

### KO Enrichment Function

```{r}
#| vscode: {languageId: r}
enrich_orthogroups_with_enricher <- function(input_path, data_path, KEGG_database, target_ogs,
                                            output_path, plot_dir=NULL,
                                            p_cutoff=0.05, q_cutoff=0.05, min_gs_size=1,
                                            show_category=10, plot_width=14, plot_height=10,
                                            font_size=18, title_size=22, axis_title_size=20,
                                            color_palette="Set2", output_formats=c("png", "pdf"),
                                            dpi=300, theme_style="publication") {

  # Resolve default plot_dir from config
  if (is.null(plot_dir)) {
    plot_dir <- paste0(KO_ENRICHMENT_DIR, "/Plots")
  }

  # Load additional libraries
  if (!requireNamespace("viridis", quietly = TRUE)) {
    install.packages("viridis")
  }
  if (!requireNamespace("RColorBrewer", quietly = TRUE)) {
    install.packages("RColorBrewer")
  }
  library(viridis)
  library(RColorBrewer)

  tryCatch({
    # Load gene-to-pathway mappings
    gene2pathway_name <- read.table(data_path, sep="\t", header=TRUE, stringsAsFactors=FALSE)
    term2gene <- gene2pathway_name[, c("Pathway", "GID")]

    # Extract genes from target orthogroups
    orthogroups_df <- read.delim(input_path, header = TRUE)
    genes_of_interest <- c()
    filtered_rows <- orthogroups_df[orthogroups_df$Orthogroup %in% target_ogs, ]

    for (row in 1:nrow(filtered_rows)) {
      selected_row <- filtered_rows[row, ]

      for (colname in names(selected_row)) {
        if (colname != "Orthogroup") {
          genes_split <- unlist(strsplit(as.character(selected_row[[colname]]), ", "))
          genes_of_interest <- c(genes_of_interest, genes_split)
        }
      }
    }

    # Remove duplicates and empty strings
    genes_of_interest <- unique(genes_of_interest[genes_of_interest != ""])

    if (length(genes_of_interest) == 0) {
      warning("No genes found for KO enrichment analysis")
      return(NULL)
    }

    # Load KEGG database
    load(file=KEGG_database)
    pathway2name$Name <- gsub("\\s*\\[BR:ko[0-9]+\\]", "", pathway2name$Name)

    # Perform KEGG enrichment analysis
    KEGG_result <- enricher(
      gene          = genes_of_interest,
      TERM2GENE     = term2gene,
      TERM2NAME     = pathway2name,
      pvalueCutoff  = p_cutoff,
      qvalueCutoff  = q_cutoff,
      pAdjustMethod = "BH",
      minGSSize     = min_gs_size
    )

    # Save results
    write.table(as.data.frame(KEGG_result), output_path, sep="\t", row.names=FALSE, quote=FALSE)

    # Create plot directory
    if (!dir.exists(plot_dir)) {
      dir.create(plot_dir, recursive=TRUE)
    }

    # Generate visualizations if results exist
    if (nrow(KEGG_result) > 0) {

      # Define publication-quality theme
      kegg_publication_theme <- function() {
        theme_minimal() +
        theme(
          text = element_text(size = font_size, family = "sans"),
          plot.title = element_text(size = title_size, face = "bold", hjust = 0.5, margin = margin(b = 20)),
          axis.title.x = element_text(size = axis_title_size, face = "bold", margin = margin(t = 15)),
          axis.title.y = element_text(size = axis_title_size, face = "bold", margin = margin(r = 15)),
          axis.text.x = element_text(size = font_size - 1, color = "black", angle = 0),
          axis.text.y = element_text(size = font_size - 2, color = "black", hjust = 1),
          legend.title = element_text(size = font_size + 2, face = "bold"),
          legend.text = element_text(size = font_size),
          legend.position = "right",
          legend.margin = margin(l = 20),
          legend.key.size = unit(0.8, "cm"),
          panel.grid.major.x = element_line(color = "grey90", size = 0.5),
          panel.grid.minor.x = element_line(color = "grey95", size = 0.3),
          panel.grid.major.y = element_line(color = "grey90", size = 0.3),
          panel.border = element_rect(color = "black", fill = NA, size = 0.8),
          plot.margin = margin(20, 25, 20, 20)
        )
      }

      # Define color palette function
      get_kegg_color_palette <- function(palette_name, n_colors) {
        if (palette_name == "viridis") {
          return(viridis::viridis(n_colors))
        } else if (palette_name %in% rownames(RColorBrewer::brewer.pal.info)) {
          max_colors <- RColorBrewer::brewer.pal.info[palette_name, "maxcolors"]
          if (n_colors <= max_colors) {
            return(RColorBrewer::brewer.pal(max(3, n_colors), palette_name))
          } else {
            return(colorRampPalette(RColorBrewer::brewer.pal(max_colors, palette_name))(n_colors))
          }
        } else {
          return(RColorBrewer::brewer.pal(max(3, min(n_colors, 8)), "Set2"))
        }
      }

      # Enhanced Bar plot
      barplot_plot <- barplot(KEGG_result, showCategory = show_category, x = "Count") +
        scale_fill_gradientn(
          colors = get_kegg_color_palette(color_palette, 100),
          name = "-log10(p.adjust)",
          guide = guide_colorbar(title.position = "top", title.hjust = 0.5)
        ) +
        labs(
          title = "KEGG Pathway Enrichment Analysis",
          x = "Gene Count",
          y = "KEGG Pathways",
          subtitle = paste("Top", min(show_category, nrow(KEGG_result)), "enriched pathways")
        ) +
        kegg_publication_theme()

      # Enhanced Dot plot
      dotplot_plot <- dotplot(KEGG_result, showCategory = show_category, x = "GeneRatio") +
        scale_color_gradientn(
          colors = get_kegg_color_palette(color_palette, 100),
          name = "-log10(p.adjust)",
          guide = guide_colorbar(title.position = "top", title.hjust = 0.5)
        ) +
        scale_size_continuous(name = "Gene Count", range = c(3, 10)) +
        labs(
          title = "KEGG Pathway Enrichment Analysis",
          x = "Gene Ratio",
          y = "KEGG Pathways",
          subtitle = paste("Top", min(show_category, nrow(KEGG_result)), "enriched pathways")
        ) +
        kegg_publication_theme() +
        theme(legend.box = "horizontal", legend.margin = margin(t = 15))

      # Save plots
      for (format in output_formats) {
        barplot_file <- file.path(plot_dir, paste0("KO_enrichment_barplot.", format))
        dotplot_file <- file.path(plot_dir, paste0("KO_enrichment_dotplot.", format))

        if (format == "pdf") {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height,
                 device = "pdf", useDingbats = FALSE)
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height,
                 device = "pdf", useDingbats = FALSE)
        } else {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height,
                 dpi = dpi, device = format)
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height,
                 dpi = dpi, device = format)
        }
      }

      print(paste("KO enrichment completed. Results saved to:", output_path))
      print(paste("Publication-quality plots saved to:", plot_dir))
    } else {
      print("No significant KEGG pathways found for the given cutoffs")
    }

    return(KEGG_result)

  }, error = function(e) {
    print(paste("Error in KO enrichment analysis:", e$message))
    return(NULL)
  })
}
```

### KO Enrichment: Positive Selection (Tentative)

KEGG pathway enrichment for genes showing evidence of positive selection.

```{r}
#| vscode: {languageId: r}
enrich_orthogroups_with_enricher(
  input_path = ORTHOGROUPS_TSV,
  data_path = EGGNOG_GENE2PATHWAY,
  KEGG_database = EGGNOG_KEGG_INFO,
  output_path = paste0(OUTPUT_TABLES, "/KO_enrichment_KaKs_positive_tentative_results.tsv"),
  target_ogs = positive_tentative_OG,
  plot_width = 12,
  plot_height = 12,
  plot_dir = paste0(OUTPUT_FIGURES, "/KO_Enrichment/KaKs_Positive_Tentative_KO_Plots")
)
```

### KO Enrichment: Relaxed Purifying Selection

KEGG pathway enrichment for genes with reduced functional constraint.

```{r}
#| vscode: {languageId: r}
enrich_orthogroups_with_enricher(
  input_path = ORTHOGROUPS_TSV,
  data_path = EGGNOG_GENE2PATHWAY,
  KEGG_database = EGGNOG_KEGG_INFO,
  output_path = paste0(OUTPUT_TABLES, "/KO_enrichment_KaKs_relaxed_purifying_results.tsv"),
  target_ogs = relaxed_OG,
  plot_width = 12,
  plot_height = 12,
  plot_dir = paste0(OUTPUT_FIGURES, "/KO_Enrichment/KaKs_Relaxed_Purifying_KO_Plots")
)
```

### KO Enrichment: Strong Purifying Selection

KEGG pathway enrichment for essential genes under strong purifying selection.

```{r}
#| vscode: {languageId: r}
enrich_orthogroups_with_enricher(
  input_path = ORTHOGROUPS_TSV,
  data_path = EGGNOG_GENE2PATHWAY,
  KEGG_database = EGGNOG_KEGG_INFO,
  output_path = paste0(OUTPUT_TABLES, "/KO_enrichment_KaKs_strong_purifying_results.tsv"),
  target_ogs = strong_purifying_OG,
  plot_width = 12,
  plot_height = 12,
  plot_dir = paste0(OUTPUT_FIGURES, "/KO_Enrichment/KaKs_Strong_Purifying_KO_Plots")
)
```

## Summary and Results Organization

The Ka/Ks selection-based enrichment analysis generates the following outputs:

### GO Enrichment Results

- **Tables**: `./Notebooks/Tables/GO_enrichment_KaKs_*_results.csv`
- **Figures**: `./Notebooks/Figures/GO_Enrichment/KaKs_*/`

### KO Enrichment Results

- **Tables**: `./Notebooks/Tables/KO_enrichment_KaKs_*_results.tsv`
- **Figures**: `./Notebooks/Figures/KO_Enrichment/KaKs_*/`

### Selection Categories Summary

```{r}
#| vscode: {languageId: r}
cat("\n========================================\n")
cat("    FINAL ANALYSIS SUMMARY\n")
cat("========================================\n\n")

cat("=== Selection Categories ===\n")
cat(paste("Positive selection (tentative):  ", length(positive_tentative_OG), "orthogroups\n"))
cat(paste("Relaxed purifying:               ", length(relaxed_OG), "orthogroups\n"))
cat(paste("Moderate purifying:              ", length(moderate_purifying_OG), "orthogroups\n"))
cat(paste("Strong purifying:                ", length(strong_purifying_OG), "orthogroups\n"))

cat("\n=== Output Files ===\n")
cat("GO Enrichment tables: ./Notebooks/Tables/GO_enrichment_KaKs_*.csv\n")
cat("GO Enrichment figures: ./Notebooks/Figures/GO_Enrichment/KaKs_*/\n")
cat("KO Enrichment tables: ./Notebooks/Tables/KO_enrichment_KaKs_*.tsv\n")
cat("KO Enrichment figures: ./Notebooks/Figures/KO_Enrichment/KaKs_*/\n")
```

### Interpretation Guidelines

#### Positive Selection (Tentative)

- Enriched GO terms/pathways suggest **functional categories under adaptive evolution**
- May indicate **ecological specialization** in *A. laticeps*
- Examples: immune response, sensory perception, metabolism genes

#### Relaxed Purifying Selection

- Enriched GO terms/pathways may indicate **reduced functional constraint**
- Could reflect **niche-specific adaptation** where certain functions are less critical
- May identify genes undergoing **subfunctionalization** or **pseudogenization**

#### Strong Purifying Selection

- Enriched GO terms/pathways identify **essential genes**
- Core cellular functions that cannot tolerate amino acid changes
- Examples: ribosomal proteins, DNA replication, core metabolism

## Data Sources and References

### Input Data

- **Ka/Ks Results**: `./Outputs/GenomewideSelection/summary/` (from genome-wide selection screen)
- **Orthogroups Information**: Configured via `orthofinder.orthogroups_tsv`
- **EggNOG Annotation Data**: `./Outputs/Eggnog/`

### Methods Reference

- **PAML yn00**: Yang Z, Nielsen R. Estimating synonymous and nonsynonymous substitution rates under realistic evolutionary models. Mol Biol Evol. 2000;17(1):32-43.
- **clusterProfiler**: Wu T, Hu E, Xu S, Chen M, Guo P, Dai Z, et al. clusterProfiler 4.0: A universal enrichment tool for interpreting omics data. The Innovation. 2021;2(3):100141.

---

*This analysis integrates Ka/Ks selection pressure results with functional enrichment to identify biological processes under different selection regimes in the focal species.*
