---
title: "Ultimate Enrichment Analysis Preparation - Annotation and Database Setup"
format: html
jupyter:
  kernelspec:
    display_name: clusterprofiler R
    language: R
    name: ir_clusterprofiler_481_env
---

# Enrichment Analysis Preparation

This notebook provides a comprehensive workflow for preparing enrichment analysis by setting up functional annotations using eggnog-mapper and creating custom organism databases. This is essential for downstream GO and KEGG pathway enrichment analysis of OrthoFinder results.

## Overview

**Purpose**: Prepare functional annotations and custom databases for comprehensive enrichment analysis of comparative genomics results.

**Key Steps**:
1. Generate functional annotations using eggnog-mapper
2. Extract GO and KEGG pathway information
3. Create custom OrgDb database for organism-specific analysis
4. Validate setup with test enrichment analysis

**Input Files**:
- CDS sequences for annotation (e.g., AL.cds.fa)
- KEGG hierarchy file (ko00001.json)

**Output Files**:
- eggnog annotation files
- Custom OrgDb database package
- Gene-to-GO and gene-to-pathway mapping files

## Load Configuration

```{r}
#| vscode: {languageId: r}
# Install yaml package if not available (required for config loading)
if (!requireNamespace("yaml", quietly = TRUE)) {
  cat("Installing required 'yaml' package...\n")
  install.packages("yaml", repos = "https://cloud.r-project.org")
}

# Load centralized path configuration
source("./Notebooks/config/load_config.R")
load_paths(config_path = "/home/jovyan/Notebooks/config/paths-set2.yaml", force_reload = TRUE)

# Display configuration summary
cat(sprintf("Configuration loaded: OrthoFinder Results_%s\n", get_version()))

# Define paths from centralized config
ORTHOFINDER_OUTPUTS <- get_path("orthofinder.results_dir")
FOCAL_CDS <- get_path(paste0("cds.", get_focal_species()))
KEGG_JSON <- get_path("database.kegg_json")
EGGNOG_DB <- get_path("database.eggnog")
ORG_DB_PATH <- get_path("database.org_db")
EGGNOG_OUTPUT <- get_path("eggnog.output_dir")
GO_ENRICHMENT_DIR <- get_path("outputs.go_enrichment")

# Display configuration summary
cat(sprintf("OrthoFinder results: %s\n", ORTHOFINDER_OUTPUTS))
cat(sprintf("Focal species CDS: %s\n", FOCAL_CDS))
cat(sprintf("KEGG JSON: %s\n", KEGG_JSON))
cat(sprintf("Eggnog output: %s\n", EGGNOG_OUTPUT))
```

## System Setup and Requirements

```{r}
# Check system information and setup directories
R.home()
output <- system("pwd", intern = TRUE)
print(paste("Working directory:", output))

# Create necessary output directories using config paths
system(paste("mkdir -p", shQuote(EGGNOG_OUTPUT)))
system(paste("mkdir -p", shQuote(get_path("outputs.go_enrichment"))))
system(paste("mkdir -p", shQuote(get_path("outputs.ko_enrichment"))))

# Check eggnog-mapper version and log it
system("conda run -n eggnog_env emapper.py --version > ./Apps/eggnog.version")
```

## Functional Annotation with eggnog-mapper

### Annotate CDS Sequences

```{r}
# Annotate Diptychus_maculatus CDS sequences using eggnog-mapper
# This step performs comprehensive functional annotation including GO terms and KEGG pathways
# Uses BRAKER3 CDS output from centralized config

# =============================================================================
# DEBUG: Verify all paths before running
# =============================================================================
cat("=== Path Verification ===\n\n")

cat("1. Input CDS file (FOCAL_CDS):\n")
cat("   Path:", FOCAL_CDS, "\n")
cat("   Exists:", file.exists(FOCAL_CDS), "\n")
if (file.exists(FOCAL_CDS)) {
  file_info <- file.info(FOCAL_CDS)
  cat("   Size:", format(file_info$size, big.mark=","), "bytes\n")
}

cat("\n2. Eggnog database directory (EGGNOG_DB):\n")
cat("   Path:", EGGNOG_DB, "\n")
cat("   Exists:", dir.exists(EGGNOG_DB), "\n")
if (dir.exists(EGGNOG_DB)) {
  db_files <- list.files(EGGNOG_DB, pattern = "\\.db$|\\.dmnd$", full.names = FALSE)
  cat("   Database files found:", length(db_files), "\n")
  if (length(db_files) > 0) {
    cat("   Files:", paste(head(db_files, 5), collapse = ", "),
        ifelse(length(db_files) > 5, "...", ""), "\n")
  }
}

cat("\n3. Output directory (EGGNOG_OUTPUT):\n")
cat("   Path:", EGGNOG_OUTPUT, "\n")
cat("   Exists:", dir.exists(EGGNOG_OUTPUT), "\n")

# =============================================================================
# Build and display the annotation command
# =============================================================================
output_prefix <- paste0(EGGNOG_OUTPUT, "/Eggnog")

annotation_command <- paste(
  "conda run -n eggnog_env emapper.py",
  paste("--data_dir", shQuote(EGGNOG_DB)),
  "--itype CDS",
  paste("-i", shQuote(FOCAL_CDS)),
  paste("--output", shQuote(output_prefix)),
  "-m diamond",
  "--cpu 12",
  "--override"  # Allow overwriting previous incomplete runs
)

cat("\n=== Command to Execute ===\n")
cat(annotation_command, "\n\n")

# =============================================================================
# Execute with proper error capture
# =============================================================================
# Check prerequisites before running
if (!file.exists(FOCAL_CDS)) {
  stop("ERROR: Input CDS file does not exist: ", FOCAL_CDS)
}

if (!dir.exists(EGGNOG_DB)) {
  stop("ERROR: Eggnog database directory does not exist: ", EGGNOG_DB)
}

# Create output directory if needed
if (!dir.exists(EGGNOG_OUTPUT)) {
  dir.create(EGGNOG_OUTPUT, recursive = TRUE)
  cat("Created output directory:", EGGNOG_OUTPUT, "\n")
}

cat("=== Running eggnog-mapper ===\n")
cat("This may take 10-60 minutes depending on input size...\n\n")

# Use system2() to capture both stdout and stderr
result <- system2(
  "bash",
  args = c("-c", shQuote(annotation_command)),
  stdout = TRUE,
  stderr = TRUE
)

# Display output
cat("=== Command Output ===\n")
cat(result, sep = "\n")

# Check exit status
exit_status <- attr(result, "status")
if (!is.null(exit_status) && exit_status != 0) {
  cat("\n=== ERROR: Command failed with exit status", exit_status, "===\n")
} else {
  cat("\n=== Annotation completed successfully ===\n")

  # Verify output files were created
  expected_output <- paste0(output_prefix, ".emapper.annotations")
  if (file.exists(expected_output)) {
    output_info <- file.info(expected_output)
    cat("Output file:", expected_output, "\n")
    cat("Size:", format(output_info$size, big.mark=","), "bytes\n")
  } else {
    cat("WARNING: Expected output file not found:", expected_output, "\n")
  }
}
```

## R Environment Setup

### Install Required Packages

```{r}
# Install BiocManager if not available
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Install required packages
install.packages("openxlsx")
BiocManager::install("AnnotationForge")
```

### Load Required Libraries

```{r}
# Core data manipulation and analysis libraries
library(dplyr)
library(stringr)
library(jsonlite)
library(purrr)
library(RCurl)

# Bioconductor packages for functional annotation
library(GO.db)
library(AnnotationForge)
library(clusterProfiler)
library(enrichplot)

# Additional utility libraries
library(openxlsx)
library(ggplot2)
library(gridExtra)
library(R6)
```

## Annotation Data Processing

### Process eggnog-mapper Output

```{r}
# Function to read and preprocess eggnog-mapper annotation results
read_and_process_eggnog <- function(file_path = NULL) {
  # Read and preprocess eggnog-mapper annotation file.
  #
  # Parameters:
  #   file_path: Path to the eggnog annotation file (uses config default if NULL)
  #
  # Returns:
  #   data.frame: Processed annotation data with empty strings converted to NA

  # Use config default if not provided
  if (is.null(file_path)) {
    file_path <- get_path("eggnog.annotations")
  }

  # Read tab-separated annotation file
  egg <- read.csv(file_path, header = TRUE, sep = "\t")
  
  # Convert empty strings to NA for better data handling
  egg[egg == ""] <- NA
  
  # Display column structure for verification
  cat("Annotation file columns:\n")
  print(colnames(egg))
  cat("\nDimensions:", nrow(egg), "rows x", ncol(egg), "columns\n")
  
  return(egg)
}

# Prepare annotation file by removing comment lines and formatting
prepare_annotation_file <- function() {
  # Prepare eggnog annotation file by removing header comments and formatting.
  # Uses paths from centralized config.

  cat("Preparing annotation file...\n")
  raw_anno <- paste0(EGGNOG_OUTPUT, "/Eggnog.emapper.annotations")
  clean_anno <- get_path("eggnog.annotations")

  system(paste("cp", shQuote(raw_anno), shQuote(clean_anno)))
  system(paste("sed -i '/^##/d'", shQuote(clean_anno)))  # Remove comment lines
  system(paste("sed -i 's/#//g'", shQuote(clean_anno)))  # Remove remaining # symbols
  cat("Annotation file prepared successfully.\n")
}

# Execute file preparation and read annotations
prepare_annotation_file()
egg <- read_and_process_eggnog(get_path("eggnog.annotations"))
```

## GO Annotation Processing

### Extract and Format GO Information

```{r}
create_gene2go <- function(egg, output_file) {
  # Extract GO annotations from eggnog results and create gene-to-GO mapping.
  #
  # Parameters:
  #   egg: eggnog annotation results (data.frame)
  #   output_file: Path to save gene2go mapping file
  #
  # Returns:
  #   list: Contains gene_info and gene2go data frames

  cat("Processing GO annotations...\n")
  
  # Extract gene information (GID to gene name mapping)
  gene_info <- egg %>%
    dplyr::select(GID = query, GENENAME = seed_ortholog) %>%
    na.omit()
  
  cat("Found", nrow(gene_info), "genes with name annotations\n")
  
  # Extract GO terms from annotations
  goterms <- egg %>%
    dplyr::select(query, GOs) %>%
    na.omit() %>%
    filter(str_detect(GOs, "GO"))
  
  cat("Found", nrow(goterms), "genes with GO annotations\n")
  
  # Process GO terms by splitting comma-separated values
  all_go_list <- str_split(goterms$GOs, ",")
  
  # Create gene-to-GO mapping with evidence codes
  gene2go <- data.frame(
    GID = rep(goterms$query, times = sapply(all_go_list, length)),
    GO = unlist(all_go_list),
    EVIDENCE = "IEA"  # Inferred from Electronic Annotation
  )
  
  cat("Created", nrow(gene2go), "gene-to-GO mappings\n")
  
  # Save to file
  write.table(gene2go, file = output_file, sep = "\t", row.names = FALSE, quote = FALSE)
  cat("Gene-to-GO mapping saved to:", output_file, "\n")
  
  # Display sample data
  cat("\nSample gene2go data:\n")
  print(head(gene2go))
  
  return(list(gene_info = gene_info, gene2go = gene2go))
}

# Process GO annotations
results <- create_gene2go(egg, get_path("eggnog.gene2go"))
gene_info <- results$gene_info
gene2go <- results$gene2go
```

## KEGG Pathway Processing

### Generate KEGG Hierarchy Information

```{r}
update_kegg <- function(json = NULL, file = NULL) {
  # Process KEGG hierarchy JSON to create pathway and KO mappings.
  #
  # This function parses the KEGG ko00001.json file to extract:
  #   - Pathway ID to pathway name mappings
  #   - KO (KEGG Orthology) to pathway mappings
  #
  # Parameters:
  #   json: Path to ko00001.json file from KEGG (uses KEGG_JSON from config if NULL)
  #   file: Path to save processed KEGG information
  #
  # References:
  #   KEGG BRITE hierarchy: https://www.genome.jp/kegg-bin/get_htext?ko00001

  # Use config default if not provided
  if (is.null(json)) {
    json <- KEGG_JSON
  }

  cat("Processing KEGG hierarchy information...\n")
  
  # Initialize data structures
  pathway2name <- tibble(Pathway = character(), Name = character())
  ko2pathway <- tibble(Ko = character(), Pathway = character())
  
  # Load KEGG JSON data
  kegg <- fromJSON(json)
  
  # Parse hierarchical structure: Level A -> Level B -> Level C (pathways) -> KOs
  for (a in seq_along(kegg[["children"]][["children"]])) {
    A <- kegg[["children"]][["name"]][[a]]  # Level A category
    
    for (b in seq_along(kegg[["children"]][["children"]][[a]][["children"]])) {
      B <- kegg[["children"]][["children"]][[a]][["name"]][[b]]  # Level B category
      
      for (c in seq_along(kegg[["children"]][["children"]][[a]][["children"]][[b]][["children"]])) {
        # Extract pathway information (Level C)
        pathway_info <- kegg[["children"]][["children"]][[a]][["children"]][[b]][["name"]][[c]]
        
        # Extract pathway ID (e.g., ko00010)
        pathway_id <- str_match(pathway_info, "ko[0-9]{5}")[1]
        
        # Clean pathway name by removing ID suffix
        pathway_name <- str_replace(pathway_info, " \\[PATH:ko[0-9]{5}\\]", "") %>% 
                       str_replace("[0-9]{5} ", "")
        
        # Store pathway information
        pathway2name <- rbind(pathway2name, tibble(Pathway = pathway_id, Name = pathway_name))
        
        # Extract KO identifiers for this pathway
        kos_info <- kegg[["children"]][["children"]][[a]][["children"]][[b]][["children"]][[c]][["name"]]
        kos <- str_match(kos_info, "K[0-9]*")[,1]
        kos <- kos[!is.na(kos)]  # Remove NA values
        
        if (length(kos) > 0) {
          ko2pathway <- rbind(ko2pathway, tibble(Ko = kos, Pathway = rep(pathway_id, length(kos))))
        }
      }
    }
  }
  
  cat("Processed", nrow(pathway2name), "pathways and", nrow(ko2pathway), "KO-pathway mappings\n")
  
  # Save processed data
  if (!is.null(file)) {
    save(pathway2name, ko2pathway, file = file)
    cat("KEGG information saved to:", file, "\n")
  }
  
  return(list(pathway2name = pathway2name, ko2pathway = ko2pathway))
}

# Generate KEGG information
# Set FORCE_REGENERATE to TRUE to overwrite existing files
FORCE_REGENERATE <- TRUE

kegg_info_output <- get_path("eggnog.kegg_info")
if (FORCE_REGENERATE || !file.exists(kegg_info_output)) {
  cat("Generating KEGG hierarchy information...\n")
  update_kegg(json = KEGG_JSON, file = kegg_info_output)
  cat("KEGG hierarchy information saved to:", kegg_info_output, "\n")
} else {
  cat("KEGG hierarchy information already exists. Set FORCE_REGENERATE <- TRUE to overwrite.\n")
}
```

### Extract KEGG Annotations from eggnog Results

```{r}
process_egg_for_koterms <- function(egg, kegg_info_path = NULL) {
  # Extract KEGG Orthology (KO) terms from eggnog annotation results.
  #
  # Parameters:
  #   egg: eggnog annotation results (data.frame)
  #   kegg_info_path: Path to KEGG hierarchy information file (uses config default if NULL)
  #
  # Returns:
  #   data.frame: Gene-to-KO mappings

  # Use config default if not provided
  if (is.null(kegg_info_path)) {
    kegg_info_path <- get_path("eggnog.kegg_info")
  }

  cat("Processing KEGG annotations...\n")

  # Extract gene KEGG annotations
  koterms <- egg %>%
    dplyr::select(GID = query, KO = KEGG_ko) %>%
    na.omit() %>%
    filter(str_detect(KO, "ko"))

  cat("Found", nrow(koterms), "genes with KEGG annotations\n")

  # Display sample data
  cat("\nSample KO terms:\n")
  print(head(koterms))

  # Ensure KEGG information file exists
  if (!file.exists(kegg_info_path)) {
    cat("KEGG information file not found. Generating...\n")
    update_kegg(json = KEGG_JSON, file = kegg_info_path)
  }

  return(koterms)
}

# Process KEGG annotations
koterms <- process_egg_for_koterms(egg, get_path("eggnog.kegg_info"))
```

### Create Gene-to-Pathway Mappings

```{r}
processKEGGData <- function(koterms, kegg_info_path = NULL,
                           pathway2name_path = NULL,
                           gene2pathway_name_path = NULL) {
  # Process KEGG data to create comprehensive gene-to-pathway mappings.
  #
  # Parameters:
  #   koterms: Gene-to-KO mappings from eggnog (data.frame)
  #   kegg_info_path: Path to processed KEGG hierarchy data (uses config default if NULL)
  #   pathway2name_path: Output path for pathway name mappings (uses config default if NULL)
  #   gene2pathway_name_path: Output path for gene-pathway-name mappings (uses config default if NULL)
  #
  # Returns:
  #   data.frame: Gene-to-pathway mappings

  # Use config defaults if not provided
  if (is.null(kegg_info_path)) {
    kegg_info_path <- get_path("eggnog.kegg_info")
  }
  if (is.null(pathway2name_path)) {
    pathway2name_path <- paste0(EGGNOG_OUTPUT, "/pathway2name.txt")
  }
  if (is.null(gene2pathway_name_path)) {
    gene2pathway_name_path <- get_path("eggnog.gene2pathway")
  }

  cat("Processing KEGG pathway mappings...\n")

  # Load KEGG hierarchy information
  load(kegg_info_path)

  cat("Loaded KEGG data:\n")
  cat("- Pathways:", nrow(pathway2name), "\n")
  cat("- KO-pathway mappings:", nrow(ko2pathway), "\n")

  # Save pathway names for reference
  write.table(pathway2name, file = pathway2name_path, sep = "\t", row.names = FALSE, quote = FALSE)

  # Standardize column names for joining
  colnames(ko2pathway) <- c("KO", "Pathway")

  # Remove "ko:" prefix from KO terms to match ko2pathway format
  koterms$KO <- str_replace_all(koterms$KO, "ko:", "")

  # Create gene-to-pathway mappings by joining with KO-pathway data
  gene2pathway <- koterms %>%
    left_join(ko2pathway, by = "KO") %>%
    dplyr::select(GID, Pathway) %>%
    na.omit()

  cat("Created", nrow(gene2pathway), "gene-to-pathway mappings\n")

  # Add pathway names for comprehensive mapping
  gene2pathway_name <- left_join(gene2pathway, pathway2name, by = "Pathway")

  # Save comprehensive mapping
  write.table(gene2pathway_name, file = gene2pathway_name_path, sep = "\t", row.names = FALSE, quote = FALSE)

  cat("Files saved:\n")
  cat("- Pathway names:", pathway2name_path, "\n")
  cat("- Gene-pathway-names:", gene2pathway_name_path, "\n")

  # Display sample data
  cat("\nSample gene2pathway data:\n")
  print(head(gene2pathway))
  cat("\nSample gene2pathway_name data:\n")
  print(head(gene2pathway_name))

  return(gene2pathway)
}

# Create comprehensive KEGG mappings (uses config paths by default)
gene2pathway <- processKEGGData(koterms)
```

## Custom Organism Database Creation

### Build OrgDb Package

> Using a lot of RAM and time, so this step must run on a HPC node with sufficient resources. Here is the script: `06_Fish_ComparativeGenomics4AL_Publication/Jobs/submit_create_orgdb.sh`

```{r}
# Create custom organism database package for the focal species
# This enables species-specific enrichment analysis using clusterProfiler

# Set to TRUE to regenerate even if package exists
FORCE_REGENERATE_ORGDB <- TRUE

org_pkg_dir <- get_path("database.org_db")
org_sqlite_file <- file.path(
  get_path("base.database"),
  sub("\\.eg\\.db$", ".eg.sqlite", basename(get_path("database.org_db")))
)

# Check if we need to create the package
if (FORCE_REGENERATE_ORGDB || !dir.exists(org_pkg_dir)) {

  cat("=== Creating custom OrgDb package ===\n\n")

  # Clean up any existing partial files to avoid conflicts
  if (file.exists(org_sqlite_file)) {
    cat("Removing stale SQLite file:", org_sqlite_file, "\n")
    file.remove(org_sqlite_file)
  }

  if (dir.exists(org_pkg_dir)) {
    cat("Removing existing package directory:", org_pkg_dir, "\n")
    unlink(org_pkg_dir, recursive = TRUE)
  }

  cat("\nBuilding organism package with annotation data...\n")
  cat("  - Genes:", nrow(gene_info), "\n")
  cat("  - GO mappings:", nrow(gene2go), "\n")
  cat("  - KO mappings:", nrow(koterms), "\n")
  cat("  - Pathway mappings:", nrow(gene2pathway), "\n\n")

  # Build organism package with all annotation data
  # Tax ID from config corresponds to focal species in NCBI Taxonomy
  # Reference: https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi
  makeOrgPackage(
    gene_info = gene_info,      # Gene ID to gene name mappings
    go = gene2go,              # Gene-to-GO term mappings
    ko = koterms,              # Gene-to-KO mappings
    pathway = gene2pathway,     # Gene-to-pathway mappings
    version = "0.0.4",
    maintainer = 'HUANGYuxiang <huang-yuxiang951@g.ecc.u-tokyo.ac.jp>',
    author = 'HUANGYuxiang <huang-yuxiang951@g.ecc.u-tokyo.ac.jp>',
    outputDir = "./DB",
    tax_id = get_focal_tax_id(),
    genus = get_focal_genus(),
    species = get_focal_species_epithet(),
    goTable = "go"
  )

  # Clean up the intermediate sqlite file (it's now inside the package)
  if (file.exists(org_sqlite_file)) {
    file.remove(org_sqlite_file)
  }

  cat(sprintf("\n=== Custom OrgDb package created successfully ===\n"))
  cat(sprintf("Location: %s\n", org_pkg_dir))

} else {
  cat("OrgDb package already exists at:", org_pkg_dir, "\n")
  cat("Set FORCE_REGENERATE_ORGDB <- TRUE to rebuild.\n")
}
```

## Database Installation and Testing

### Install Custom Database

```{r}
# Install custom organism database with execution control
# Uses flag file to prevent repeated installation

flag_file <- get_path("eggnog.executed_flag")

# Execute command only if flag file does not exist
if (!file.exists(flag_file)) {
  cat("Installing custom organism database...\n")
  cat(sprintf("  Source: %s\n", ORG_DB_PATH))

  # Install custom library from local source
  install.packages(ORG_DB_PATH, repos = NULL, type = "sources")

  # Create flag file after execution to prevent re-installation
  file.create(flag_file)
  cat("Custom database installed successfully.\n")
} else {
  cat("Custom database installation has already been completed.\n")
}

# Load custom library
orgdb_name <- get_orgdb_package_name()
library(orgdb_name, character.only = TRUE)
orgdb_obj <- get(orgdb_name)

# Display available key types
cat("Available key types in custom database:\n")
print(keytypes(orgdb_obj))

# Set database reference
GO_database <- orgdb_name
```

## Validation and Testing

### Test GO Enrichment Analysis

```{r}
# Load OrthoFinder results for testing enrichment analysis
# Reference: OrthoGroups Info from centralized config (paths.yaml)

cat("Loading OrthoFinder results for validation...\n")
orthogroups_df <- read.delim(get_path("orthofinder.orthogroups_tsv"), header = TRUE)

cat("OrthoFinder data loaded:", nrow(orthogroups_df), "orthogroups\n")
```

### Extract Genes from Specific Orthogroup

```{r}
# Example: Extract genes from a specific orthogroup for enrichment testing
extract_orthogroup_genes <- function(orthogroups_df, target_og = "OG0000000") {
  # Extract all gene IDs from a specific orthogroup across all species.
  #
  # Parameters:
  #   orthogroups_df: OrthoFinder orthogroups data (data.frame)
  #   target_og: Target orthogroup ID (default: "OG0000000")
  #
  # Returns:
  #   vector: Gene IDs from the specified orthogroup

  cat("Extracting genes from orthogroup:", target_og, "\n")
  
  # Select the target orthogroup row
  selected_row <- orthogroups_df[orthogroups_df$Orthogroup == target_og, ]
  
  if (nrow(selected_row) == 0) {
    cat("Warning: Orthogroup", target_og, "not found\n")
    return(character(0))
  }
  
  # Extract gene IDs from all species columns
  genes_of_interest <- c()
  
  for (colname in names(selected_row)) {
    if (colname != "Orthogroup") {
      # Split comma-separated gene IDs and combine
      genes_split <- unlist(strsplit(as.character(selected_row[[colname]]), ", "))
      genes_split <- genes_split[genes_split != ""]  # Remove empty strings
      genes_of_interest <- c(genes_of_interest, genes_split)
    }
  }
  
  # Remove duplicates and empty values
  genes_of_interest <- unique(genes_of_interest[genes_of_interest != ""])
  
  cat("Found", length(genes_of_interest), "unique genes in", target_og, "\n")
  
  return(genes_of_interest)
}

# Extract genes from test orthogroup
target_og <- "OG0000000"
genes_of_interest <- extract_orthogroup_genes(orthogroups_df, target_og)

cat("Sample genes from", target_og, ":\n")
print(head(genes_of_interest))
```

### Perform Test GO Enrichment

```{r}
# Perform GO enrichment analysis using custom database
if (length(genes_of_interest) > 0) {
  cat("Performing test GO enrichment analysis...\n")
  
  # GO enrichment analysis with relaxed parameters for testing
  ego_result <- enrichGO(
    gene          = unique(genes_of_interest),
    OrgDb         = orgdb_obj,
    keyType       = "GID",           # Gene ID type in custom database
    ont           = "BP",            # Biological Process ontology
    pAdjustMethod = "BH",           # Benjamini-Hochberg correction
    pvalueCutoff  = 1,              # Relaxed p-value for testing
    qvalueCutoff  = 1               # Relaxed q-value for testing
  )
  
  # Display results summary
  cat("GO enrichment results:\n")
  print(head(ego_result))
  
  # Create visualizations if results are available
  if (nrow(as.data.frame(ego_result)) > 0) {
    cat("Creating visualizations...\n")
    
    # Bar plot of top enriched terms
    p1 <- barplot(ego_result, showCategory = 20, title = "GO Enrichment - Bar Plot")
    print(p1)
    ggsave(paste0(GO_ENRICHMENT_DIR, "/GO_enrichment_test_barplot.png"), plot = p1, width = 12, height = 8, dpi = 300)
    ggsave(paste0(GO_ENRICHMENT_DIR, "/GO_enrichment_test_barplot.pdf"), plot = p1, width = 12, height = 8)
    
    # Dot plot of enriched terms
    p2 <- dotplot(ego_result, showCategory = 20, title = "GO Enrichment - Dot Plot")
    print(p2)
    ggsave(paste0(GO_ENRICHMENT_DIR, "/GO_enrichment_test_dotplot.png"), plot = p2, width = 12, height = 8, dpi = 300)
    ggsave(paste0(GO_ENRICHMENT_DIR, "/GO_enrichment_test_dotplot.pdf"), plot = p2, width = 12, height = 8)
    
    # Save results to file
    write.csv(as.data.frame(ego_result), 
             paste0(GO_ENRICHMENT_DIR, "/GO_enrichment_test_results.csv"), 
             row.names = FALSE)
    
    cat("Test enrichment results saved to: ./Outputs/GO_Enrichment/GO_enrichment_test_results.csv\n")
    cat("Test enrichment plots saved to: ./Outputs/GO_Enrichment/GO_enrichment_test_*.png and *.pdf\n")
  } else {
    cat("No significant enrichment found with current parameters.\n")
  }
} else {
  cat("No genes found for enrichment analysis.\n")
}
```

## Quality Control and Validation

### Database Statistics

```{r}
# Provide comprehensive statistics about the created database
cat("=== Database Quality Control Report ===\n\n")

cat("1. Annotation Coverage:\n")
cat("   - Total genes with annotations:", nrow(gene_info), "\n")
cat("   - Genes with GO annotations:", length(unique(gene2go$GID)), "\n")
cat("   - Genes with KEGG annotations:", length(unique(koterms$GID)), "\n")
cat("   - Genes with pathway annotations:", length(unique(gene2pathway$GID)), "\n\n")

cat("2. GO Annotation Statistics:\n")
cat("   - Total GO terms assigned:", nrow(gene2go), "\n")
cat("   - Unique GO terms:", length(unique(gene2go$GO)), "\n")
cat("   - Average GO terms per gene:", round(nrow(gene2go) / length(unique(gene2go$GID)), 2), "\n\n")

cat("3. KEGG Annotation Statistics:\n")
cat("   - Total KO assignments:", nrow(koterms), "\n")
cat("   - Unique KO terms:", length(unique(koterms$KO)), "\n")
cat("   - Total pathway assignments:", nrow(gene2pathway), "\n")
cat("   - Unique pathways:", length(unique(gene2pathway$Pathway)), "\n\n")

cat("4. Database Verification:\n")
cat("   - Custom OrgDb package:", get_orgdb_package_name(), "\n")
cat("   - Available key types:", paste(keytypes(orgdb_obj), collapse = ", "), "\n")
cat("   - Database ready for enrichment analysis: TRUE\n")
```

### Out[1]: ✓ Done 0.49s
```
=== Database Quality Control Report ===


1. Annotation Coverage:

   - Total genes with annotations: 92683 

   - Genes with GO annotations: 43245 

   - Genes with KEGG annotations: 45567 

   - Genes with pathway annotations: 30789 


2. GO Annotation Statistics:

   - Total GO terms assigned: 6202355 

   - Unique GO terms: 23339 

   - Average GO terms per gene: 143.42 


3. KEGG Annotation Statistics:

   - Total KO assignments: 45567 

   - Unique KO terms: 8867 

   - Total pathway assignments: 124669 

   - Unique pathways: 456 


4. Database Verification:

   - Custom OrgDb package: org.Dmaculatus.eg.db

   - Available key types: EVIDENCE, EVIDENCEALL, GENENAME, GID, GO, GOALL, KO, ONTOLOGY, ONTOLOGYALL, Pathway 

   - Database ready for enrichment analysis: TRUE
```


## Output Files Summary

The following files have been generated for downstream enrichment analysis:

```{r}
# List all generated output files (using config paths)
output_files <- c(
  get_path("eggnog.annotations"),
  get_path("eggnog.gene2go"),
  get_path("eggnog.kegg_info"),
  paste0(EGGNOG_OUTPUT, "/pathway2name.txt"),
  get_path("eggnog.gene2pathway"),
  ORG_DB_PATH,
  paste0(GO_ENRICHMENT_DIR, "/GO_enrichment_test_results.csv")
)

cat("Generated output files:\n")
for (file in output_files) {
  if (file.exists(file) || dir.exists(file)) {
    cat("✓", file, "\n")
  } else {
    cat("✗", file, "(not found)\n")
  }
}
```

### Out[1]: ✓ Done 0.0s
```
Generated output files:

✓ /home/jovyan/Outputs/Eggnog/eggnog.anno 
✓ /home/jovyan/Outputs/Eggnog/gene2go.txt 
✓ /home/jovyan/Outputs/Eggnog/kegg_info.RData 
✓ /home/jovyan/Outputs/Eggnog/pathway2name.txt 
✓ /home/jovyan/Outputs/Eggnog/gene2pathway_name.txt 
✓ /home/jovyan/DB/org.Dmaculatus.eg.db 
✗ ./Outputs/GO_Enrichment/GO_enrichment_test_results.csv (not found)

```

Generated output files:

✓ /home/jovyan/Outputs/Eggnog/eggnog.anno 
✓ /home/jovyan/Outputs/Eggnog/gene2go.txt 
✓ /home/jovyan/Outputs/Eggnog/kegg_info.RData 
✓ /home/jovyan/Outputs/Eggnog/pathway2name.txt 
✓ /home/jovyan/Outputs/Eggnog/gene2pathway_name.txt 
✓ /home/jovyan/DB/org.Dmaculatus.eg.db 
✗ ./Outputs/GO_Enrichment/GO_enrichment_test_results.csv (not found)

```

## Next Steps

After completing the annotation and database setup:

1. **Use the custom database** (`org.Dmaculatus.eg.db`) for species-specific enrichment analysis
2. **Apply enrichment analysis** to OrthoFinder results using the prepared annotations
3. **Perform comparative analysis** across different orthogroups and species
4. **Validate results** using the generated pathway and GO mappings

## Notes

- The custom OrgDb database enables organism-specific enrichment analysis
- All annotation files are saved in standardized formats for downstream analysis
- The flag file system prevents accidental re-installation of the database
- Test results validate the functionality of the annotation pipeline

## References

- eggnog-mapper: https://github.com/eggnogdb/eggnog-mapper
- clusterProfiler: https://bioconductor.org/packages/clusterProfiler/
- KEGG Database: https://www.genome.jp/kegg/
- GO Consortium: http://geneontology.org/
- NCBI Taxonomy: https://www.ncbi.nlm.nih.gov/Taxonomy/
