---
title: "Ultimate Enrichment Analysis for Specific Orthogroups"
jupyter:
  kernelspec:
    display_name: clusterprofiler R
    language: R
    name: ir_clusterprofiler_4100_env
---

## Overview
This notebook performs comprehensive GO and KO enrichment analysis for specific orthogroups identified in the comparative genomics analysis. It analyzes three categories of orthogroups:
- **Increased gene families**: Where *Aspiorhynchus_laticeps* has the maximum gene count
- **Decreased gene families**: Where *Aspiorhynchus_laticeps* has the minimum gene count  
- **Species-specific families**: Where only *Aspiorhynchus_laticeps* has genes (unique families)

## Environment Setup

```{r}
#| vscode: {languageId: r}
R.home()
```

## Methods

### Load Required Libraries

```{r}
#| vscode: {languageId: r}
library(dplyr)
library(stringr)
library(jsonlite)
library(purrr)
library(RCurl)
library(GO.db)
library(AnnotationForge)
library(clusterProfiler)
library(openxlsx)
library(ggplot2)
library(gridExtra)
library(R6)
library(enrichplot)
```

### Initialize Custom Annotation Database

```{r}
#| vscode: {languageId: r}
flag_file <- "./Outputs/Eggnog/executed.flag"

# Install custom database only if not already done
if (!file.exists(flag_file)) {
  install.packages("./DB/org.Alaticeps.eg.db", repos=NULL, type="sources")
  file.create(flag_file)
} else {
  print("Custom annotation database already installed.")
}

library(org.Alaticeps.eg.db)
keytypes(org.Alaticeps.eg.db)
GO_database <- 'org.Alaticeps.eg.db'
```

## Orthogroup Classification

### Data Loading and Classification
[Orthogroups Gene Count Data](./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.GeneCount.tsv)

#### Increased Gene Families
Orthogroups where *Aspiorhynchus_laticeps* has the maximum gene count compared to other species.

```{r}
#| vscode: {languageId: r}
# Read the orthogroup gene count data
data <- read.table("./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.GeneCount.tsv", 
                   header = TRUE, stringsAsFactors = FALSE)

# Find orthogroups where Aspiorhynchus_laticeps has the maximum gene count
orthogroups_with_max <- data$Orthogroup[apply(data[, -c(1, ncol(data))], 1, function(row) {
  row["Aspiorhynchus_laticeps"] == max(row)
})]

# Store as vector for analysis
increased_genes_OG <- c(orthogroups_with_max)
print(paste("Number of increased gene families:", length(increased_genes_OG)))
```

#### Decreased Gene Families
Orthogroups where *Aspiorhynchus_laticeps* has the minimum gene count compared to other species.

```{r}
#| vscode: {languageId: r}
# Find orthogroups where Aspiorhynchus_laticeps has the minimum gene count
orthogroups_with_min <- data$Orthogroup[apply(data[, -c(1, ncol(data))], 1, function(row) {
  row["Aspiorhynchus_laticeps"] == min(row)
})]

# Store as vector for analysis
decreased_genes_OG <- c(orthogroups_with_min)
print(paste("Number of decreased gene families:", length(decreased_genes_OG)))
```

#### Species-Specific Gene Families
Orthogroups where only *Aspiorhynchus_laticeps* has genes (all other species have zero).

```{r}
#| vscode: {languageId: r}
# Find orthogroups unique to Aspiorhynchus_laticeps
orthogroups_with_unique_nonzero <- data$Orthogroup[apply(data[, -c(1, ncol(data))], 1, function(row) {
  asp_value <- row["Aspiorhynchus_laticeps"]
  other_values <- row[names(row) != "Aspiorhynchus_laticeps"]
  asp_value > 0 && all(other_values == 0)
})]

# Store as vector for analysis
unique_nonzero_genes_OG <- c(orthogroups_with_unique_nonzero)
print(paste("Number of species-specific gene families:", length(unique_nonzero_genes_OG)))
print("First 10 unique orthogroups:")
print(head(unique_nonzero_genes_OG, 10))
```

## GO Enrichment Analysis

### GO Enrichment Function
This function performs Gene Ontology enrichment analysis for specific orthogroups.

```{r}
#| vscode: {languageId: r}
enrich_orthogroups <- function(input_path, ont, output_path, target_ogs, org_db, 
                              key_type="GID", plot_dir="./Outputs/GO_Enrichment/Plots",
                              p_cutoff=0.01, q_cutoff=0.05, 
                              # Publication-quality plot parameters
                              show_category=20, plot_width=12, plot_height=8, 
                              font_size=18, title_size=22, axis_title_size=20,
                              color_palette="viridis", output_formats=c("png", "pdf"),
                              dpi=300, theme_style="publication") {
  
  # Load additional libraries for enhanced plotting
  if (!requireNamespace("viridis", quietly = TRUE)) {
    install.packages("viridis")
  }
  if (!requireNamespace("RColorBrewer", quietly = TRUE)) {
    install.packages("RColorBrewer")
  }
  library(viridis)
  library(RColorBrewer)
  
  # Read OrthoFinder results
  orthogroups_df <- read.delim(input_path, header = TRUE)
  
  # Extract gene IDs from target orthogroups
  genes_of_interest <- c()
  filtered_rows <- orthogroups_df[orthogroups_df$Orthogroup %in% target_ogs, ]
  
  for (row in 1:nrow(filtered_rows)) {
    selected_row <- filtered_rows[row, ]
    
    for (colname in names(selected_row)) {
      if (colname != "Orthogroup") {
        genes_split <- unlist(strsplit(as.character(selected_row[[colname]]), ", "))
        genes_of_interest <- c(genes_of_interest, genes_split)
      }
    }
  }
  
  # Remove duplicates and empty strings
  genes_of_interest <- unique(genes_of_interest[genes_of_interest != ""])
  
  if (length(genes_of_interest) == 0) {
    warning("No genes found for the specified orthogroups")
    return(NULL)
  }
  
  # Perform GO enrichment analysis
  tryCatch({
    ego_result <- enrichGO(
      gene          = genes_of_interest,
      OrgDb         = org_db,
      keyType       = key_type,
      ont           = ont,
      pAdjustMethod = "BH",
      pvalueCutoff  = p_cutoff,
      qvalueCutoff  = q_cutoff
    )
    
    # Create output directory
    if (!dir.exists(plot_dir)) {
      dir.create(plot_dir, recursive = TRUE)
    }
    
    # Save results table
    write.csv(as.data.frame(ego_result), output_path, row.names=FALSE)
    
    # Generate and save visualizations if results exist
    if (nrow(ego_result) > 0) {
      
      # Define publication-quality theme
      publication_theme <- function() {
        theme_minimal() +
        theme(
          # Text elements
          text = element_text(size = font_size, family = "Arial"),
          plot.title = element_text(size = title_size, face = "bold", hjust = 0.5, margin = margin(b = 20)),
          axis.title.x = element_text(size = axis_title_size, face = "bold", margin = margin(t = 15)),
          axis.title.y = element_text(size = axis_title_size, face = "bold", margin = margin(r = 15)),
          axis.text.x = element_text(size = font_size, color = "black"),
          axis.text.y = element_text(size = font_size, color = "black", hjust = 1),
          
          # Legend
          legend.title = element_text(size = font_size + 2, face = "bold"),
          legend.text = element_text(size = font_size),
          legend.position = "right",
          legend.margin = margin(l = 20),
          legend.key.size = unit(0.8, "cm"),
          
          # Panel and grid
          panel.grid.major = element_line(color = "grey90", size = 0.5),
          panel.grid.minor = element_line(color = "grey95", size = 0.3),
          panel.border = element_rect(color = "black", fill = NA, size = 0.8),
          
          # Plot margins
          plot.margin = margin(20, 20, 20, 20),
          
          # Strip text for facets
          strip.text = element_text(size = font_size, face = "bold", margin = margin(5, 5, 5, 5))
        )
      }
      
      # Define color palette (robust version with error handling)
      get_color_palette <- function(palette_name, n_colors) {
        if (palette_name == "viridis") {
          return(viridis::viridis(n_colors))
        } else if (palette_name == "plasma") {
          return(viridis::plasma(n_colors))
        } else if (palette_name == "inferno") {
          return(viridis::inferno(n_colors))
        } else if (palette_name == "magma") {
          return(viridis::magma(n_colors))
        } else if (palette_name == "turbo") {
          return(viridis::turbo(n_colors))
        } else if (palette_name %in% rownames(RColorBrewer::brewer.pal.info)) {
          max_colors <- RColorBrewer::brewer.pal.info[palette_name, "maxcolors"]
          if (n_colors <= max_colors) {
            # Ensure minimum of 3 colors for RColorBrewer to prevent crashes
            return(RColorBrewer::brewer.pal(max(3, n_colors), palette_name))
          } else {
            return(colorRampPalette(RColorBrewer::brewer.pal(max_colors, palette_name))(n_colors))
          }
        } else {
          # Robust default fallback with error handling
          return(RColorBrewer::brewer.pal(max(3, min(n_colors, 8)), "Set2"))
        }
      }
      
      # Enhanced Bar plot
      ont_full_name <- switch(ont,
                             "BP" = "Biological Process",
                             "MF" = "Molecular Function", 
                             "CC" = "Cellular Component",
                             ont)
      
      barplot_plot <- barplot(ego_result, showCategory = show_category, x = "Count") +
        scale_fill_gradientn(colors = get_color_palette(color_palette, 100),
                           name = "-log10(p.adjust)",
                           guide = guide_colorbar(title.position = "top")) +
        labs(
          title = paste("GO", ont_full_name, "Enrichment Analysis"),
          x = "Gene Count",
          y = "GO Terms",
          subtitle = paste("Top", min(show_category, nrow(ego_result)), "enriched terms")
        ) +
        publication_theme() +
        theme(
          axis.text.y = element_text(angle = 0, hjust = 1, size = font_size),
          plot.subtitle = element_text(size = font_size + 2, hjust = 0.5, margin = margin(b = 15))
        )
      
      # Enhanced Dot plot  
      dotplot_plot <- dotplot(ego_result, showCategory = show_category, x = "GeneRatio") +
        scale_color_gradientn(colors = get_color_palette(color_palette, 100),
                            name = "-log10(p.adjust)",
                            guide = guide_colorbar(title.position = "top")) +
        scale_size_continuous(name = "Gene Count", range = c(3, 8)) +
        labs(
          title = paste("GO", ont_full_name, "Enrichment Analysis"),
          x = "Gene Ratio",
          y = "GO Terms",
          subtitle = paste("Top", min(show_category, nrow(ego_result)), "enriched terms")
        ) +
        publication_theme() +
        theme(
          axis.text.y = element_text(angle = 0, hjust = 1, size = font_size),
          plot.subtitle = element_text(size = font_size + 2, hjust = 0.5, margin = margin(b = 15)),
          legend.box = "horizontal",
          legend.margin = margin(t = 10)
        )
      
      # Save plots in multiple formats
      for (format in output_formats) {
        # Bar plot
        barplot_file <- file.path(plot_dir, paste0("GO_enrichment_barplot.", format))
        if (format == "pdf") {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height, 
                 device = "pdf", useDingbats = FALSE)
        } else if (format == "svg") {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height, 
                 device = "svg")
        } else {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height, 
                 dpi = dpi, device = format)
        }
        
        # Dot plot
        dotplot_file <- file.path(plot_dir, paste0("GO_enrichment_dotplot.", format))
        if (format == "pdf") {
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height, 
                 device = "pdf", useDingbats = FALSE)
        } else if (format == "svg") {
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height, 
                 device = "svg")
        } else {
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height, 
                 dpi = dpi, device = format)
        }
      }
      
      print(paste("GO enrichment completed. Results saved to:", output_path))
      print(paste("Publication-quality plots saved to:", plot_dir))
      print(paste("Available formats:", paste(output_formats, collapse = ", ")))
    } else {
      print("No significant GO terms found for the given cutoffs")
    }
    
    return(ego_result)
    
  }, error = function(e) {
    print(paste("Error in GO enrichment:", e$message))
    return(NULL)
  })
}
```

### GO Enrichment Analysis: Increased Gene Families

#### Biological Process (BP)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "BP",
  output_path = "./Notebooks/Tables/GO_enrichment_increased_BP_results.csv",
  target_ogs = increased_genes_OG,
  org_db = org.Alaticeps.eg.db,
  color_palette="Set2",
  plot_height = 18,
  plot_width= 18,
  plot_dir = "./Notebooks/Figures/GO_Enrichment/Increased_BP_Plots"
)
```

#### Molecular Function (MF)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "MF",
  output_path = "./Notebooks/Tables/GO_enrichment_increased_MF_results.csv",
  target_ogs = increased_genes_OG,
  org_db = org.Alaticeps.eg.db,
  color_palette="Set2",
  plot_height = 28,
  plot_width= 28,
  plot_dir = "./Notebooks/Figures/GO_Enrichment/Increased_MF_Plots"
)
```

#### Cellular Component (CC)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "CC",
  output_path = "./Notebooks/Tables/GO_enrichment_increased_CC_results.csv",
  target_ogs = increased_genes_OG,
  org_db = org.Alaticeps.eg.db,
  plot_height = 12,
  plot_width = 12,
  color_palette="Set2",
  plot_dir = "./Notebooks/Figures/GO_Enrichment/Increased_CC_Plots"
)
```

### GO Enrichment Analysis: Decreased Gene Families

#### Biological Process (BP)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "BP",
  output_path = "./Notebooks/Tables/GO_enrichment_decreased_BP_results.csv",
  target_ogs = decreased_genes_OG,
  org_db = org.Alaticeps.eg.db,
  color_palette="Set2",
  plot_dir = "./Notebooks/Figures/GO_Enrichment/decreased_BP_Plots"
)
```

#### Molecular Function (MF)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "MF",
  output_path = "./Notebooks/Tables/GO_enrichment_decreased_MF_results.csv",
  target_ogs = decreased_genes_OG,
  org_db = org.Alaticeps.eg.db,
  color_palette="Set2",
  plot_dir = "./Notebooks/Figures/GO_Enrichment/decreased_MF_Plots"
)
```

#### Cellular Component (CC)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "CC",
  output_path = "./Notebooks/Tables/GO_enrichment_decreased_CC_results.csv",
  target_ogs = decreased_genes_OG,
  org_db = org.Alaticeps.eg.db,
  color_palette="Set2",
  plot_dir = "./Notebooks/Figures/GO_Enrichment/decreased_CC_Plots"
)
```

### GO Enrichment Analysis: Species-Specific Gene Families

#### Biological Process (BP)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "BP",
  output_path = "./Notebooks/Tables/GO_enrichment_unique_nonzero_BP_results.csv",
  target_ogs = unique_nonzero_genes_OG,
  org_db = org.Alaticeps.eg.db,
  color_palette="Set2",
  plot_height = 24,
  plot_width = 24,
  plot_dir = "./Notebooks/Figures/GO_Enrichment/unique_nonzero_BP_Plots"
)
```

#### Molecular Function (MF)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "MF",
  output_path = "./Notebooks/Tables/GO_enrichment_unique_nonzero_MF_results.csv",
  target_ogs = unique_nonzero_genes_OG,
  org_db = org.Alaticeps.eg.db,
  plot_height = 16,
  plot_width = 16,
  color_palette="Set2",
  plot_dir = "./Notebooks/Figures/GO_Enrichment/unique_nonzero_MF_Plots"
)
```

#### Cellular Component (CC)

```{r}
#| vscode: {languageId: r}
enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "CC",
  output_path = "./Notebooks/Tables/GO_enrichment_unique_nonzero_CC_results.csv",
  target_ogs = unique_nonzero_genes_OG,
  org_db = org.Alaticeps.eg.db,
  color_palette="Set2",
  plot_dir = "./Notebooks/Figures/GO_Enrichment/unique_nonzero_CC_Plots"
)
```

## KEGG Pathway (KO) Enrichment Analysis

### KO Enrichment Function
This function performs KEGG pathway enrichment analysis using the enricher function from clusterProfiler.

```{r}
#| vscode: {languageId: r}
enrich_orthogroups_with_enricher <- function(input_path, data_path, KEGG_database, target_ogs, 
                                            output_path, plot_dir="./Outputs/KO_Enrichment/Plots",
                                            p_cutoff=0.05, q_cutoff=0.05, min_gs_size=1,
                                            # Publication-quality plot parameters
                                            show_category=20, plot_width=14, plot_height=10, 
                                            font_size=18, title_size=22, axis_title_size=20,
                                            color_palette="Set2", output_formats=c("png", "pdf"),
                                            dpi=300, theme_style="publication") {
  
  # Load additional libraries for enhanced plotting
  if (!requireNamespace("viridis", quietly = TRUE)) {
    install.packages("viridis")
  }
  if (!requireNamespace("RColorBrewer", quietly = TRUE)) {
    install.packages("RColorBrewer")
  }
  library(viridis)
  library(RColorBrewer)
  
  tryCatch({
    # Load gene-to-pathway mappings
    gene2pathway_name <- read.table(data_path, sep="\t", header=TRUE, stringsAsFactors=FALSE)
    term2gene <- gene2pathway_name[, c("Pathway", "GID")]
    
    # Extract genes from target orthogroups
    orthogroups_df <- read.delim(input_path, header = TRUE)
    genes_of_interest <- c()
    filtered_rows <- orthogroups_df[orthogroups_df$Orthogroup %in% target_ogs, ]
    
    for (row in 1:nrow(filtered_rows)) {
      selected_row <- filtered_rows[row, ]
      
      for (colname in names(selected_row)) {
        if (colname != "Orthogroup") {
          genes_split <- unlist(strsplit(as.character(selected_row[[colname]]), ", "))
          genes_of_interest <- c(genes_of_interest, genes_split)
        }
      }
    }
    
    # Remove duplicates and empty strings
    genes_of_interest <- unique(genes_of_interest[genes_of_interest != ""])
    
    if (length(genes_of_interest) == 0) {
      warning("No genes found for KO enrichment analysis")
      return(NULL)
    }
    
    # Load KEGG database
    load(file=KEGG_database)
    pathway2name$Name <- gsub("\\s*\\[BR:ko[0-9]+\\]", "", pathway2name$Name)
    
    # Perform KEGG enrichment analysis
    KEGG_result <- enricher(
      gene          = genes_of_interest,
      TERM2GENE     = term2gene,
      TERM2NAME     = pathway2name,
      pvalueCutoff  = p_cutoff,
      qvalueCutoff  = q_cutoff,
      pAdjustMethod = "BH",
      minGSSize     = min_gs_size
    )
    
    # Save results
    write.table(as.data.frame(KEGG_result), output_path, sep="\t", row.names=FALSE, quote=FALSE)
    
    # Create plot directory
    if (!dir.exists(plot_dir)) {
      dir.create(plot_dir, recursive=TRUE)
    }
    
    # Generate visualizations if results exist
    if (nrow(KEGG_result) > 0) {
      
      # Define publication-quality theme for KEGG plots
      kegg_publication_theme <- function() {
        theme_minimal() +
        theme(
          # Text elements
          text = element_text(size = font_size, family = "Arial"),
          plot.title = element_text(size = title_size, face = "bold", hjust = 0.5, margin = margin(b = 20)),
          axis.title.x = element_text(size = axis_title_size, face = "bold", margin = margin(t = 15)),
          axis.title.y = element_text(size = axis_title_size, face = "bold", margin = margin(r = 15)),
          axis.text.x = element_text(size = font_size - 1, color = "black", angle = 0),
          axis.text.y = element_text(size = font_size - 2, color = "black", hjust = 1),
          
          # Legend
          legend.title = element_text(size = font_size + 2, face = "bold"),
          legend.text = element_text(size = font_size),
          legend.position = "right",
          legend.margin = margin(l = 20),
          legend.key.size = unit(0.8, "cm"),
          legend.box = "vertical",
          
          # Panel and grid
          panel.grid.major.x = element_line(color = "grey90", size = 0.5),
          panel.grid.minor.x = element_line(color = "grey95", size = 0.3),
          panel.grid.major.y = element_line(color = "grey90", size = 0.3),
          panel.grid.minor.y = element_blank(),
          panel.border = element_rect(color = "black", fill = NA, size = 0.8),
          
          # Plot margins
          plot.margin = margin(20, 25, 20, 20),
          
          # Strip text for facets
          strip.text = element_text(size = font_size, face = "bold", margin = margin(5, 5, 5, 5))
        )
      }
      
      # Define color palette function
      get_kegg_color_palette <- function(palette_name, n_colors) {
        if (palette_name == "viridis") {
          return(viridis::viridis(n_colors))
        } else if (palette_name == "plasma") {
          return(viridis::plasma(n_colors))
        } else if (palette_name == "inferno") {
          return(viridis::inferno(n_colors))
        } else if (palette_name == "magma") {
          return(viridis::magma(n_colors))
        } else if (palette_name == "turbo") {
          return(viridis::turbo(n_colors))
        } else if (palette_name %in% rownames(RColorBrewer::brewer.pal.info)) {
          max_colors <- RColorBrewer::brewer.pal.info[palette_name, "maxcolors"]
          if (n_colors <= max_colors) {
            return(RColorBrewer::brewer.pal(max(3, n_colors), palette_name))
          } else {
            return(colorRampPalette(RColorBrewer::brewer.pal(max_colors, palette_name))(n_colors))
          }
        } else {
          return(RColorBrewer::brewer.pal(max(3, min(n_colors, 8)), "Set2"))  # Default fallback
        }
      }
      
      # Function to truncate long pathway names for better readability
      truncate_names <- function(names, max_length = 50) {
        ifelse(nchar(names) > max_length, 
               paste0(substr(names, 1, max_length), "..."), 
               names)
      }
      
      # Prepare data for enhanced plotting
      kegg_data <- as.data.frame(KEGG_result)
      kegg_data$Description_short <- truncate_names(kegg_data$Description, 45)
      
      # Enhanced Bar plot
      barplot_plot <- barplot(KEGG_result, showCategory = show_category, x = "Count") +
        scale_fill_gradientn(
          colors = get_kegg_color_palette(color_palette, 100),
          name = "-log10(p.adjust)",
          guide = guide_colorbar(
            title.position = "top",
            title.hjust = 0.5,
            barwidth = unit(0.8, "cm"),
            barheight = unit(8, "cm")
          )
        ) +
        labs(
          title = "KEGG Pathway Enrichment Analysis",
          x = "Gene Count",
          y = "KEGG Pathways",
          subtitle = paste("Top", min(show_category, nrow(KEGG_result)), "enriched pathways"),
          caption = "Pathways ranked by statistical significance (p.adjust)"
        ) +
        kegg_publication_theme() +
        theme(
          plot.subtitle = element_text(size = font_size + 2, hjust = 0.5, margin = margin(b = 15)),
          plot.caption = element_text(size = font_size, hjust = 1, margin = margin(t = 10))
        )
      
      # Enhanced Dot plot with GeneRatio
      dotplot_plot <- dotplot(KEGG_result, showCategory = show_category, x = "GeneRatio") +
        scale_color_gradientn(
          colors = get_kegg_color_palette(color_palette, 100),
          name = "-log10(p.adjust)",
          guide = guide_colorbar(
            title.position = "top",
            title.hjust = 0.5,
            barwidth = unit(0.8, "cm"),
            barheight = unit(6, "cm")
          )
        ) +
        scale_size_continuous(
          name = "Gene Count", 
          range = c(3, 10),
          guide = guide_legend(
            title.position = "top",
            title.hjust = 0.5,
            override.aes = list(color = "grey50")
          )
        ) +
        labs(
          title = "KEGG Pathway Enrichment Analysis",
          x = "Gene Ratio",
          y = "KEGG Pathways",
          subtitle = paste("Top", min(show_category, nrow(KEGG_result)), "enriched pathways"),
          caption = "Dot size represents gene count; Color intensity represents significance"
        ) +
        kegg_publication_theme() +
        theme(
          plot.subtitle = element_text(size = font_size + 2, hjust = 0.5, margin = margin(b = 15)),
          plot.caption = element_text(size = font_size, hjust = 1, margin = margin(t = 10)),
          legend.box = "horizontal",
          legend.margin = margin(t = 15)
        )
      
      # Save plots in multiple formats
      for (format in output_formats) {
        # Bar plot
        barplot_file <- file.path(plot_dir, paste0("KO_enrichment_barplot.", format))
        if (format == "pdf") {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height, 
                 device = "pdf", useDingbats = FALSE)
        } else if (format == "svg") {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height, 
                 device = "svg")
        } else {
          ggsave(barplot_file, plot = barplot_plot, width = plot_width, height = plot_height, 
                 dpi = dpi, device = format)
        }
        
        # Dot plot
        dotplot_file <- file.path(plot_dir, paste0("KO_enrichment_dotplot.", format))
        if (format == "pdf") {
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height, 
                 device = "pdf", useDingbats = FALSE)
        } else if (format == "svg") {
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height, 
                 device = "svg")
        } else {
          ggsave(dotplot_file, plot = dotplot_plot, width = plot_width, height = plot_height, 
                 dpi = dpi, device = format)
        }
      }
      
      print(paste("KO enrichment completed. Results saved to:", output_path))
      print(paste("Publication-quality plots saved to:", plot_dir))
      print(paste("Available formats:", paste(output_formats, collapse = ", ")))
    } else {
      print("No significant KEGG pathways found for the given cutoffs")
    }
    
    return(KEGG_result)
    
  }, error = function(e) {
    print(paste("Error in KO enrichment analysis:", e$message))
    print("This may be due to insufficient gene overlap or data availability")
    return(NULL)
  })
}
```

### KO Enrichment Analysis: Increased Gene Families

```{r}
#| vscode: {languageId: r}
enrich_orthogroups_with_enricher(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  data_path = "./Outputs/Eggnog/gene2pathway_name.txt",
  KEGG_database = "./Outputs/Eggnog/kegg_info.RData",
  output_path = "./Notebooks/Tables/KO_enrichment_increased_results.tsv",
  target_ogs = increased_genes_OG,
  plot_width = 16,
  plot_height = 16,
  plot_dir = "./Notebooks/Figures/KO_Enrichment/increased_KO_Plots"
)
```

### KO Enrichment Analysis: Decreased Gene Families

```{r}
#| vscode: {languageId: r}
enrich_orthogroups_with_enricher(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  data_path = "./Outputs/Eggnog/gene2pathway_name.txt",
  KEGG_database = "./Outputs/Eggnog/kegg_info.RData",
  output_path = "./Notebooks/Tables/KO_enrichment_decreased_results.tsv",
  target_ogs = decreased_genes_OG,
  plot_dir = "./Notebooks/Figures/KO_Enrichment/decreased_KO_Plots"
)
```

#### NOTEABLE Output
```
Out[1]: ✓ Done 2m 48.3s
[1] "No significant KEGG pathways found for the given cutoffs"

#
# over-representation test
#
#...@organism 	 UNKNOWN 
#...@ontology 	 UNKNOWN 
#...@gene 	 chr [1:175377] "ROI15509.1" "ROI15537.1" "ROI15538.1" "ROI15539.1" ...
#...pvalues adjusted by 'BH' with cutoff <0.05 
#...0 enriched terms found
#...Citation
 T Wu, E Hu, S Xu, M Chen, P Guo, Z Dai, T Feng, L Zhou, W Tang, L Zhan, X Fu, S Liu, X Bo, and G Yu.
 clusterProfiler 4.0: A universal enrichment tool for interpreting omics data.
 The Innovation. 2021, 2(3):100141 
```

### KO Enrichment Analysis: Species-Specific Gene Families

```{r}
#| vscode: {languageId: r}
enrich_orthogroups_with_enricher(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  data_path = "./Outputs/Eggnog/gene2pathway_name.txt",
  KEGG_database = "./Outputs/Eggnog/kegg_info.RData",
  output_path = "./Notebooks/Tables/KO_enrichment_unique_nonzero_results.tsv",
  target_ogs = unique_nonzero_genes_OG,
  plot_dir = "./Notebooks/Figures/KO_Enrichment/unique_nonzero_KO_Plots"
)
```

## Quick Test Example
This section provides a simple test example for GO enrichment analysis using a small set of orthogroups.

```{r}
#| vscode: {languageId: r}
# Test with a small set of orthogroups
test_ogs <- c("OG0001234", "OG0005678", "OG0009101")  # Example orthogroups

# Test GO enrichment function
test_result <- enrich_orthogroups(
  input_path = "./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv",
  ont = "BP",
  output_path = "./Notebooks/Tables/GO_enrichment_test_results.csv",
  target_ogs = test_ogs,
  org_db = org.Alaticeps.eg.db,
  plot_dir = "./Notebooks/Figures/GO_Enrichment/Test_Plots",
  p_cutoff = 1.0,  # Relaxed cutoffs for testing
  q_cutoff = 1.0
)

if (!is.null(test_result)) {
  print("Test enrichment analysis completed successfully")
  print(head(test_result))
} else {
  print("Test analysis returned no results - check orthogroup IDs and data paths")
}
```

## Summary and Results Organization

The enrichment analysis generates the following outputs:

### GO Enrichment Results
- **Tables**: `./Notebooks/Tables/GO_enrichment_*_results.csv`
- **Figures**: `./Notebooks/Figures/GO_Enrichment/*/`

### KO Enrichment Results  
- **Tables**: `./Notebooks/Tables/KO_enrichment_*_results.tsv`
- **Figures**: `./Notebooks/Figures/KO_Enrichment/*/`

### Analysis Categories
1. **Increased gene families**: Gene families expanded in *Aspiorhynchus_laticeps*
2. **Decreased gene families**: Gene families contracted in *Aspiorhynchus_laticeps*  
3. **Species-specific families**: Gene families unique to *Aspiorhynchus_laticeps*

## Future Improvements and ToDo Items

- **Dual format output**: Save figures in both PNG and PDF formats for publication flexibility
- **Statistical summary**: Generate summary statistics for each enrichment category
- **Interactive plots**: Consider using plotly for interactive visualization
- **Batch processing**: Implement batch processing for multiple species comparisons
- **Advanced visualizations**: Add network plots and heatmaps for pathway relationships

## Data Sources and References

- [OrthoFinder Results](./Outputs/OrthoFinder/Results_Apr25/)
- [Gene Count Data](./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.GeneCount.tsv)
- [Orthogroups Information](./Outputs/OrthoFinder/Results_Apr25/Orthogroups/Orthogroups.tsv)
- [EggNOG Annotation Data](./Outputs/Eggnog/)

---
*This analysis uses the latest OrthoFinder results (Apr 25) and improved error handling for robust enrichment analysis.*
