---
title: State-of-the-Art Visualization for Comparative Genomics
subtitle: Modern tools for publication-quality figures
jupyter: env_biopython
---

## Overview

This notebook demonstrates modern bioinformatics visualization tools (2023-2025) for generating publication-quality figures from comparative genomics data. These tools represent significant improvements over legacy approaches.

**Structure:**
- **Sections 1-5**: Python-based visualizations (ETE4, pyCirclize, pyGenomeViz)
- **Sections 6-7**: R-based visualizations (ggtree, ggstatsplot)
- **Section 8**: Combined summary

**Tools covered:**

| Tool | Language | Purpose | Key Feature |
|------|----------|---------|-------------|
| ETE4 + Biopython Phylo | Python | Interactive & static trees | smartview (browser-based) + Phylo fallback |
| pyCirclize + matplotlib | Python | Circular & linear ideograms | Chromosome visualization |
| pyGenomeViz | Python | Synteny plots | Modern CLI workflows |
| ggtree + treeio | R | dN/dS tree visualization | Native CodeML parsing |
| ggstatsplot | R | Statistical plots | APA annotations |

**Prerequisites:**
- Run `0-ultimate-env_construction.qmd` Section 11 to install visualization tools
- Complete `2-ultimate-AL_OrthoFinder.qmd` for species tree
- Complete `2-genome-wide-selection-screen.qmd` for Ka/Ks results
- Complete `6-ultimate-MCScanX.qmd` for synteny data

---

# Part I: Python Visualizations

---

## 1. Setup & Configuration

### 1.1 Load Configuration

```{python}
import sys
import os
from pathlib import Path

# Add config directory to path
sys.path.insert(0, str(Path("./Notebooks/config")))

# Load centralized configuration
from load_config import (load_paths, get_path, get_version,
                         get_mcscanx_comparisons, get_focal_display_name,
                         get_focal_genus, get_focal_species_epithet,
                         get_focal_mcscanx_prefix, get_species_prefix)

# Load all paths
paths = load_paths(config_path="/home/jovyan/Notebooks/config/paths-set2.yaml")
version = get_version()
print(f"OrthoFinder version: {version}")

# Project root (container path)
PROJECT_ROOT = Path(get_path("base.project_root"))
print(f"Project root: {PROJECT_ROOT}")

# MCScanX comparison pairs from config
mcscanx_comparisons = get_mcscanx_comparisons()
comp1 = mcscanx_comparisons[0]  # First comparison pair
comp2 = mcscanx_comparisons[1] if len(mcscanx_comparisons) > 1 else None
print(f"MCScanX comparisons: {[c['name'] for c in mcscanx_comparisons]}")
```

### 1.2 Import Libraries

```{python}
# Standard libraries
import warnings
warnings.filterwarnings('ignore')

# Data manipulation
import pandas as pd
import numpy as np

# Visualization
import matplotlib
matplotlib.use('Agg')  # Headless backend
import matplotlib.pyplot as plt
import seaborn as sns

# Set publication defaults with larger fonts for readability
plt.rcParams.update({
    'figure.dpi': 300,
    'savefig.dpi': 300,
    'font.family': 'serif',
    'font.size': 14,              # Base font size (increased from 10)
    'axes.titlesize': 16,         # Title font size
    'axes.titleweight': 'bold',
    'axes.labelsize': 14,         # Axis label font size
    'axes.labelweight': 'bold',
    'xtick.labelsize': 12,        # X-axis tick label font size
    'ytick.labelsize': 12,        # Y-axis tick label font size
    'legend.fontsize': 12,        # Legend font size
    'legend.title_fontsize': 13,  # Legend title font size
})

print("Standard libraries loaded successfully (publication font sizes)")
```

### 1.3 Import State-of-the-Art Tools

```{python}
# ETE4 - Advanced tree visualization (v4.x has new smartview API)
try:
    from ete4 import Tree
    from ete4.smartview import TextFace, CircleFace, Layout
    import ete4
    print(f"ETE4 {ete4.__version__} loaded successfully (smartview API)")
except ImportError as e:
    print(f"ETE4 not available: {e}")
    print("Run: conda run -n env_biopython pip install ete4")

# Biopython Phylo - Fallback for static tree rendering
try:
    from Bio import Phylo
    print("Biopython Phylo loaded (fallback for static tree images)")
except ImportError:
    print("Biopython Phylo not available")

# pyGenomeViz - Modern synteny visualization
try:
    from pygenomeviz import GenomeViz
    from pygenomeviz.parser import Gff
    import pygenomeviz
    print(f"pyGenomeViz {pygenomeviz.__version__} loaded successfully")
except ImportError as e:
    print(f"pyGenomeViz not available: {e}")
    print("Run: conda run -n env_biopython pip install pygenomeviz")

# pyCirclize - Circos-style plots
try:
    from pycirclize import Circos
    import pycirclize
    print(f"pyCirclize {pycirclize.__version__} loaded successfully")
except ImportError as e:
    print(f"pyCirclize not available: {e}")
    print("Run: conda run -n env_biopython pip install pycirclize")
```

### 1.4 Create Output Directories

```{python}
# Create output directories
output_base = Path(get_path("outputs.figures")) / "StateOfTheArt"
output_dirs = {
    'base': output_base,
    'trees': output_base / "trees",
    'synteny': output_base / "synteny_pygenomeviz",
    'circular': output_base / "circular_plots",
    'statistical': output_base / "statistical_plots"
}

for name, dir_path in output_dirs.items():
    dir_path.mkdir(parents=True, exist_ok=True)
    print(f"Created: {dir_path}")
```

### 1.5 Define Key Paths

```{python}
# Key data paths from configuration
species_tree_path = Path(get_path("orthofinder.species_tree"))
orthogroups_path = Path(get_path("orthofinder.orthogroups_tsv"))
gene_counts_path = Path(get_path("orthofinder.orthogroups_count"))

# Ka/Ks results
genomewide_base = Path(get_path("outputs.genomewide_selection.base_dir"))
yn00_results_dir = Path(get_path("outputs.genomewide_selection.yn00_results"))
codeml_results_dir = Path(get_path("outputs.genomewide_selection.codeml_results"))

# CAFE results
cafe_base = Path(get_path("cafe.base_dir"))
cafe_family_results = Path(get_path("cafe.single_lambda.family_results"))

# Focal species
focal_species = get_path("species.focal")

print(f"\nFocal species: {focal_species}")
print(f"Species tree: {species_tree_path}")
print(f"Ka/Ks results: {genomewide_base}")
```

### 1.6 Load Chromosome Data from MCScanX BED Files

```{python}
def load_chromosome_sizes_from_bed(bed_path: Path) -> dict:
    """Load chromosome sizes from MCScanX BED file.

    MCScanX BED format: chr, start, end, gene_id (tab-separated)
    Chromosome size = max(end) for each chromosome
    """
    if not bed_path.exists():
        print(f"Warning: BED file not found: {bed_path}")
        return {}

    df = pd.read_csv(bed_path, sep='\t', names=['chr', 'start', 'end', 'gene'])
    # Get max end position per chromosome
    sizes = df.groupby('chr')['end'].max().to_dict()
    return sizes

# Load chromosome sizes from MCScanX filtered BED files
comp1_mcscanx_dir = PROJECT_ROOT / "DB" / "MCScanX" / comp1['name']

focal_bed_path = comp1_mcscanx_dir / f"{comp1['focal_prefix']}-filtered-modified.bed"
comp1_bed_path = comp1_mcscanx_dir / f"{comp1['comp_prefix']}-filtered-modified.bed"

focal_chromosomes = load_chromosome_sizes_from_bed(focal_bed_path)
comp1_chromosomes = load_chromosome_sizes_from_bed(comp1_bed_path)

# Build display names from comparison species
focal_display = get_focal_display_name(abbreviated=True)
focal_genus = get_focal_genus()
focal_epithet = get_focal_species_epithet()
comp1_genus = comp1['comp_species'].split('_')[0]
comp1_epithet = comp1['comp_species'].split('_')[1]
comp1_display = f"{comp1_genus[0]}. {comp1_epithet}"

# LaTeX-formatted species names for matplotlib
focal_latex = f"$\\it{{{focal_genus[0]}.\\ {focal_epithet}}}$"
focal_latex_full = f"$\\it{{{focal_genus}\\ {focal_epithet}}}$"
comp1_latex = f"$\\it{{{comp1_genus[0]}.\\ {comp1_epithet}}}$"
comp1_latex_full = f"$\\it{{{comp1_genus}\\ {comp1_epithet}}}$"

print(f"{focal_display} chromosomes: {len(focal_chromosomes)}")
if focal_chromosomes:
    total_focal = sum(focal_chromosomes.values())
    print(f"  Total genome size: {total_focal/1e9:.2f} Gb")
    print(f"  Largest: {max(focal_chromosomes.values())/1e6:.1f} Mb")
    print(f"  Smallest: {min(focal_chromosomes.values())/1e6:.1f} Mb")

print(f"\n{comp1_display} chromosomes: {len(comp1_chromosomes)}")
if comp1_chromosomes:
    total_comp1 = sum(comp1_chromosomes.values())
    print(f"  Total genome size: {total_comp1/1e9:.2f} Gb")
    print(f"  Largest: {max(comp1_chromosomes.values())/1e6:.1f} Mb")
    print(f"  Smallest: {min(comp1_chromosomes.values())/1e6:.1f} Mb")
```

### 1.6.1 Load Second Comparison Species Chromosome Data

```{python}
# Load second comparison species chromosome sizes from MCScanX BED files
if comp2:
    comp2_mcscanx_dir = PROJECT_ROOT / "DB" / "MCScanX" / comp2['name']
    comp2_focal_bed_path = comp2_mcscanx_dir / f"{comp2['focal_prefix']}-filtered-modified.bed"
    comp2_bed_path = comp2_mcscanx_dir / f"{comp2['comp_prefix']}-filtered-modified.bed"

    comp2_focal_chromosomes = load_chromosome_sizes_from_bed(comp2_focal_bed_path)
    comp2_chromosomes = load_chromosome_sizes_from_bed(comp2_bed_path)

    comp2_genus = comp2['comp_species'].split('_')[0]
    comp2_epithet = comp2['comp_species'].split('_')[1]
    comp2_display = f"{comp2_genus[0]}. {comp2_epithet}"

    # LaTeX-formatted species names for matplotlib
    comp2_latex = f"$\\it{{{comp2_genus[0]}.\\ {comp2_epithet}}}$"
    comp2_latex_full = f"$\\it{{{comp2_genus}\\ {comp2_epithet}}}$"

    print(f"{comp2_display} chromosomes: {len(comp2_chromosomes)}")
    if comp2_chromosomes:
        total_comp2 = sum(comp2_chromosomes.values())
        print(f"  Total genome size: {total_comp2/1e9:.2f} Gb")
        print(f"  Largest: {max(comp2_chromosomes.values())/1e6:.1f} Mb")
        print(f"  Smallest: {min(comp2_chromosomes.values())/1e6:.1f} Mb")

        # Show first few chromosomes
        print("\n  Sample chromosome sizes:")
        for name, size in list(sorted(comp2_chromosomes.items(),
                                       key=lambda x: natural_sort_key(x[0])
                                       if 'natural_sort_key' in dir() else x[0]))[:5]:
            print(f"    {name}: {size/1e6:.1f} Mb")
```

---

## 2. ETE4 Tree Visualization (Python)

ETE Toolkit v4 provides advanced tree visualization with the web-based "smartview" that eliminates Qt/Xvfb dependencies.

> **Key advantage**: ETE4's `smartview` module uses a Layout-based API for custom visualization, and the `ete-evol` module can both run CodeML analyses AND visualize results directly.

> **API Note (ETE4 v4.x)**: The old `TreeStyle`/`NodeStyle` classes from ETE3 are replaced by `Layout` classes and Face objects in `ete4.smartview`. Static image rendering requires either PyQt6 (`render()`) or selenium (`render_sm()`).

```{python}
#| label: ete4-setup

from ete4 import Tree
from ete4.smartview import TextFace, CircleFace, Layout
from pathlib import Path

# Read species tree
tree_path = Path(get_path("orthofinder.species_tree"))
tree = Tree(open(tree_path).read())

print(f"Tree loaded with {len(tree)} leaves")
# Note: ETE4 renamed iter_leaves() to leaves()
for leaf in tree.leaves():
    print(f"  - {leaf.name}")
```

### 2.1 Basic ETE4 Tree Rendering

ETE4 uses a Layout-based system for visualization. We'll create a custom layout and use Biopython's Phylo as a fallback for static image generation.

```{python}
#| label: ete4-basic

# ETE4 v4.x uses Layout classes instead of TreeStyle/NodeStyle
# For static images without PyQt6, we use Biopython Phylo as fallback

# First, display tree structure using ETE4's text visualization
print("Tree structure (ASCII):")
print(tree.to_str(props=['name', 'dist']))

# Create custom layout for ETE4 smartview
class SpeciesTreeLayout(Layout):
    """Custom layout for species tree visualization."""

    def __init__(self):
        super().__init__('SpeciesTree')

    def draw_node(self, node, point):
        # Add circle at each node
        yield CircleFace(radius=4, fill='steelblue')

        # Add species name for leaves (with italics simulation via formatting)
        if node.is_leaf:
            display_name = node.name.replace("_", " ")
            yield TextFace(display_name, fs=10)

# Create layout instance
species_layout = SpeciesTreeLayout()
print("\nCustom layout created: SpeciesTreeLayout")
```

### 2.2 Static Tree Image with Biopython Phylo (Fallback)

Since ETE4's `render()` requires PyQt6, we use Biopython's Phylo module for static image generation:

```{python}
#| label: ete4-phylo-fallback

from Bio import Phylo
import matplotlib.pyplot as plt

# Read tree with Biopython
tree_path = Path(get_path("orthofinder.species_tree"))
bio_tree = Phylo.read(str(tree_path), "newick")

# Create publication-quality figure
fig, ax = plt.subplots(figsize=(12, 8))

# Draw tree
Phylo.draw(bio_tree, axes=ax, do_show=False)

# Style adjustments for publication
ax.set_xlabel("Branch Length (substitutions/site)", fontsize=10)
ax.set_ylabel("")
ax.spines['top'].set_visible(False)
ax.spines['right'].set_visible(False)

# Italicize species names (modify tick labels)
for text in ax.get_yticklabels():
    text_content = text.get_text()
    text.set_text(text_content.replace('_', ' '))
    text.set_fontstyle('italic')

plt.title("Species Phylogeny (Biopython Phylo)", fontsize=12, fontweight='bold')
plt.tight_layout()

# Save
output_path = output_dirs['trees'] / "species_tree_biopython.png"
fig.savefig(str(output_path), dpi=300, bbox_inches='tight')
fig.savefig(str(output_path.with_suffix('.pdf')), bbox_inches='tight')
print(f"Saved: {output_path}")
plt.close(fig)
```

### 2.3 Tree with Selection Pressure Annotations

```{python}
#| label: ete4-selection

# Load genome-wide selection results
# Handle filename mismatch: try expected name first, then actual name from selection screen
summary_dir = Path(get_path("outputs.genomewide_selection.summary"))

# Try both possible filenames (different naming conventions between notebooks)
possible_files = [
    summary_dir / "genomewide_selection_master_results.tsv",  # Originally expected
    summary_dir / "genome_wide_selection_master.tsv",         # Actual from selection screen
]

master_file = None
for candidate in possible_files:
    if candidate.exists():
        master_file = candidate
        print(f"Found selection results: {master_file.name}")
        break

if master_file is not None:
    selection_df = pd.read_csv(master_file, sep='\t')
    print(f"Loaded {len(selection_df):,} orthogroups")

    # Summarize selection by category (actual column: selection_yn00)
    if 'selection_yn00' in selection_df.columns:
        category_counts = selection_df['selection_yn00'].value_counts()
        print("\nSelection Category Summary (yn00):")
        for cat, count in category_counts.items():
            print(f"  {cat}: {count:,} orthogroups")

    # Get positive selection genes for annotation
    # Check for codeml results first, fall back to yn00 max_omega
    if 'codeml_omega_foreground' in selection_df.columns:
        positive_mask = selection_df['codeml_omega_foreground'] > 1
        positive_selection = selection_df[positive_mask]
        print(f"\nOrthogroups with ω > 1 (codeml foreground): {len(positive_selection):,}")
    elif 'max_omega' in selection_df.columns:
        positive_mask = selection_df['max_omega'] > 1
        positive_selection = selection_df[positive_mask]
        print(f"\nOrthogroups with ω > 1 (yn00 max): {len(positive_selection):,}")
else:
    print(f"Selection results not found. Tried:")
    for f in possible_files:
        print(f"  - {f}")
    print("Run 2-genome-wide-selection-screen.qmd first")
```

---

## 3. Circular Genome Plots (Python: pyCirclize)

pyCirclize creates Circos-style circular plots for genome visualization with synteny links.

> **Key advantage**: Enables visualization of synteny, gene density, and selection hotspots in a single compact figure.

### 3.1 Load Chromosome Data and Gene Density

```{python}
#| label: pycirclize-setup

from pycirclize import Circos
from pycirclize.parser import Gff as CircosGff

def calculate_gene_density(bed_path: Path, window_size: int = 1000000) -> dict:
    """Calculate gene density per chromosome from BED file.

    Args:
        bed_path: Path to MCScanX BED file (chr, start, end, gene_id)
        window_size: Window size in bp for density calculation (default 1 Mb)

    Returns:
        Dict mapping chromosome name to list of (window_start, gene_count) tuples
    """
    if not bed_path.exists():
        return {}

    df = pd.read_csv(bed_path, sep='\t', names=['chr', 'start', 'end', 'gene'])

    density = {}
    for chrom in df['chr'].unique():
        chrom_df = df[df['chr'] == chrom]
        max_pos = chrom_df['end'].max()

        # Calculate gene count per window
        windows = []
        for window_start in range(0, int(max_pos) + window_size, window_size):
            window_end = window_start + window_size
            # Count genes with midpoint in this window
            midpoints = (chrom_df['start'] + chrom_df['end']) / 2
            count = ((midpoints >= window_start) & (midpoints < window_end)).sum()
            windows.append((window_start, count))

        density[chrom] = windows

    return density

# Calculate gene density for each species
print("Calculating gene density from BED files...")
focal_gene_density = calculate_gene_density(focal_bed_path)
comp1_gene_density = calculate_gene_density(comp1_bed_path)
comp2_gene_density = calculate_gene_density(comp2_bed_path)

print(f"{focal_display}: {len(focal_gene_density)} chromosomes with gene density data")
print(f"{comp1_display}: {len(comp1_gene_density)} chromosomes with gene density data")
print(f"{comp2_display}: {len(comp2_gene_density)} chromosomes with gene density data")

# Show sample density for focal species first chromosome
focal_chr1 = sorted(focal_gene_density.keys(), key=lambda s: [int(c) if c.isdigit() else c for c in __import__('re').split('([0-9]+)', s)])[0] if focal_gene_density else None
if focal_chr1:
    chr1_density = focal_gene_density[focal_chr1]
    total_genes = sum(d[1] for d in chr1_density)
    print(f"\nSample: {focal_chr1} has {total_genes} genes across {len(chr1_density)} 1-Mb windows")
```

```{python}
#| label: pycirclize-chrom-data

# Use chromosome sizes from MCScanX BED files
if focal_chromosomes:
    # Sort chromosomes naturally
    def natural_sort_key(s):
        import re
        return [int(c) if c.isdigit() else c for c in re.split('([0-9]+)', s)]

    sorted_focal_chroms = dict(sorted(focal_chromosomes.items(), key=lambda x: natural_sort_key(x[0])))

    print(f"{focal_display} chromosome sizes for circular plot:")
    for name, size in list(sorted_focal_chroms.items())[:5]:
        print(f"  {name}: {size/1e6:.1f} Mb")
    print(f"  ... ({len(sorted_focal_chroms)} total chromosomes)")
else:
    # Fallback to example data
    focal_prefix = get_focal_mcscanx_prefix()
    print("Using example chromosome data (real data not available)")
    sorted_focal_chroms = {
        f"{focal_prefix}1": 100000000,
        f"{focal_prefix}2": 84000000,
        f"{focal_prefix}3": 65000000,
        f"{focal_prefix}4": 62000000,
        f"{focal_prefix}5": 61000000,
    }
```

### 3.2 Basic Circos-Style Plot with All Chromosomes

```{python}
#| label: pycirclize-basic

# Create Circos plot with all focal species chromosomes
circos = Circos(sorted_focal_chroms, space=2)

# Color palette for chromosomes
n_chroms = len(sorted_focal_chroms)
colors = plt.cm.tab20(np.linspace(0, 1, min(n_chroms, 20)))
if n_chroms > 20:
    colors = plt.cm.nipy_spectral(np.linspace(0, 1, n_chroms))

# Calculate max gene density for normalization across all chromosomes
all_densities = []
for chrom in sorted_focal_chroms.keys():
    if chrom in focal_gene_density:
        all_densities.extend([d[1] for d in focal_gene_density[chrom]])
max_density = max(all_densities) if all_densities else 1

# Add tracks to each sector
for idx, sector in enumerate(circos.sectors):
    color = colors[idx % len(colors)]

    # Outer track with sector name
    outer_track = sector.add_track((95, 100))
    outer_track.axis(fc=color, alpha=0.8)
    # Only label every other chromosome to avoid crowding
    if idx % 2 == 0:
        outer_track.text(sector.name, fontsize=6, r=107)

    # Gene density track
    density_track = sector.add_track((80, 93))
    density_track.axis(fc="white", ec="gray", lw=0.5)

    # Plot gene density from BED file
    if sector.name in focal_gene_density:
        density_data = focal_gene_density[sector.name]
        # Window midpoints, clipped to chromosome boundaries
        x = np.array([min(d[0] + 500000, sector.size - 1) for d in density_data])
        y = np.array([d[1] for d in density_data])
        # Normalize to track height (0-10 range for fill_between)
        y_normalized = (y / max_density) * 10
        density_track.fill_between(x, y_normalized, fc="steelblue", alpha=0.7)

# Add title with italicized species name
circos.text(f"{focal_latex_full}\nGenome Overview ({len(sorted_focal_chroms)} chromosomes)", r=40, size=10)

# Save
output_path = output_dirs['circular'] / "genome_circos_demo.png"
fig = circos.plotfig()

# Add legend explaining the track meanings
from matplotlib.patches import Patch
from matplotlib.lines import Line2D

legend_elements = [
    Patch(facecolor='steelblue', alpha=0.7, label='Gene density (genes per Mb)'),
    Patch(facecolor='lightgray', alpha=0.8, edgecolor='black', linewidth=0.5, label='Chromosome (colored by position)')
]

fig.legend(handles=legend_elements,
           loc='lower right',
           bbox_to_anchor=(0.95, 0.05),
           fontsize=8,
           frameon=True,
           fancybox=True,
           shadow=True,
           title='Track Legend',
           title_fontsize=9)

fig.savefig(str(output_path), dpi=300, bbox_inches='tight')
print(f"Saved: {output_path}")
print(f"Gene density from {len(focal_gene_density)} chromosomes, max density = {max_density} genes/Mb")

# PDF version
fig.savefig(str(output_path.with_suffix('.pdf')), bbox_inches='tight')
plt.close(fig)
```

### 3.2.1 Comparison Species 1 Genome Circos

```{python}
#| label: pycirclize-comp1-demo

# Create Circos plot for comp1 genome
if comp1_chromosomes:
    # Sort chromosomes naturally
    sorted_comp1_chroms = dict(sorted(comp1_chromosomes.items(),
                                    key=lambda x: natural_sort_key(x[0])))

    # Filter out tiny scaffolds (< 1 Mb)
    sorted_comp1_chroms = {k: v for k, v in sorted_comp1_chroms.items() if v > 1000000}

    print(f"{comp1_display}: {len(sorted_comp1_chroms)} chromosomes for circos plot")
else:
    print("Comp1 chromosome data not available - skipping comp1 circos plot")
    sorted_comp1_chroms = None

if sorted_comp1_chroms:
    # Create Circos plot
    comp1_circos_obj = Circos(sorted_comp1_chroms, space=2)

    # Color palette for chromosomes
    n_chroms_comp1 = len(sorted_comp1_chroms)
    colors_comp1 = plt.cm.viridis(np.linspace(0, 1, n_chroms_comp1))

    # Calculate max gene density for normalization
    all_comp1_densities = []
    for chrom in sorted_comp1_chroms.keys():
        if chrom in comp1_gene_density:
            all_comp1_densities.extend([d[1] for d in comp1_gene_density[chrom]])
    max_comp1_density = max(all_comp1_densities) if all_comp1_densities else 1

    # Add tracks to each sector
    for idx, sector in enumerate(comp1_circos_obj.sectors):
        color = colors_comp1[idx % len(colors_comp1)]

        # Outer track with sector name
        outer_track = sector.add_track((95, 100))
        outer_track.axis(fc=color, alpha=0.8)
        # Only label every other chromosome to avoid crowding
        if idx % 2 == 0:
            outer_track.text(sector.name, fontsize=6, r=107)

        # Gene density track
        density_track = sector.add_track((80, 93))
        density_track.axis(fc="white", ec="gray", lw=0.5)

        # Plot gene density from BED file
        if sector.name in comp1_gene_density:
            density_data = comp1_gene_density[sector.name]
            # Window midpoints, clipped to chromosome boundaries
            x = np.array([min(d[0] + 500000, sector.size - 1) for d in density_data])
            y = np.array([d[1] for d in density_data])
            y_normalized = (y / max_comp1_density) * 10
            density_track.fill_between(x, y_normalized, fc="#e74c3c", alpha=0.7)

    # Add title with italic species name
    comp1_circos_obj.text(f"{comp1_latex_full}\nGenome Overview", r=40, size=10)

    # Save
    output_path_comp1 = output_dirs['circular'] / "genome_circos_demo_sm.png"
    fig_comp1 = comp1_circos_obj.plotfig()

    # Add legend
    legend_elements_comp1 = [
        Patch(facecolor='#e74c3c', alpha=0.7, label='Gene density (genes per Mb)'),
        Patch(facecolor='lightgray', alpha=0.8, edgecolor='black', linewidth=0.5, label='Chromosome')
    ]

    fig_comp1.legend(handles=legend_elements_comp1,
                  loc='lower right',
                  bbox_to_anchor=(0.95, 0.05),
                  fontsize=8,
                  frameon=True,
                  fancybox=True,
                  shadow=True,
                  title='Track Legend',
                  title_fontsize=9)

    fig_comp1.savefig(str(output_path_comp1), dpi=300, bbox_inches='tight')
    fig_comp1.savefig(str(output_path_comp1.with_suffix('.pdf')), bbox_inches='tight')
    print(f"Saved: {output_path_comp1}")
    print(f"Gene density, max density = {max_comp1_density} genes/Mb")
    plt.close(fig_comp1)
```

### 3.2.2 Comparison Species 2 Genome Circos

```{python}
#| label: pycirclize-comp2-demo

# Create Circos plot for comp2 genome
if comp2_chromosomes:
    # Sort chromosomes naturally
    sorted_comp2_chroms = dict(sorted(comp2_chromosomes.items(),
                                    key=lambda x: natural_sort_key(x[0])))

    # Filter out tiny scaffolds (< 1 Mb)
    sorted_comp2_chroms = {k: v for k, v in sorted_comp2_chroms.items() if v > 1000000}

    print(f"{comp2_display}: {len(sorted_comp2_chroms)} chromosomes for circos plot")
else:
    print("Comp2 chromosome data not available - skipping comp2 circos plot")
    sorted_comp2_chroms = None

if sorted_comp2_chroms:
    # Create Circos plot
    comp2_circos_obj = Circos(sorted_comp2_chroms, space=2)

    # Color palette for chromosomes - use a different colormap
    n_chroms_comp2 = len(sorted_comp2_chroms)
    colors_comp2 = plt.cm.plasma(np.linspace(0, 1, n_chroms_comp2))

    # Calculate max gene density for normalization
    all_comp2_densities = []
    for chrom in sorted_comp2_chroms.keys():
        if chrom in comp2_gene_density:
            all_comp2_densities.extend([d[1] for d in comp2_gene_density[chrom]])
    max_comp2_density = max(all_comp2_densities) if all_comp2_densities else 1

    # Add tracks to each sector
    for idx, sector in enumerate(comp2_circos_obj.sectors):
        color = colors_comp2[idx % len(colors_comp2)]

        # Outer track with sector name
        outer_track = sector.add_track((95, 100))
        outer_track.axis(fc=color, alpha=0.8)
        # Only label every other chromosome to avoid crowding
        if idx % 2 == 0:
            outer_track.text(sector.name, fontsize=6, r=107)

        # Gene density track
        density_track = sector.add_track((80, 93))
        density_track.axis(fc="white", ec="gray", lw=0.5)

        # Plot gene density from BED file
        if sector.name in comp2_gene_density:
            density_data = comp2_gene_density[sector.name]
            # Window midpoints, clipped to chromosome boundaries
            x = np.array([min(d[0] + 500000, sector.size - 1) for d in density_data])
            y = np.array([d[1] for d in density_data])
            y_normalized = (y / max_comp2_density) * 10
            density_track.fill_between(x, y_normalized, fc="#9b59b6", alpha=0.7)

    # Add title with italic species name
    comp2_circos_obj.text(f"{comp2_latex_full}\nGenome Overview", r=40, size=10)

    # Save
    output_path_comp2 = output_dirs['circular'] / "genome_circos_demo_tt.png"
    fig_comp2 = comp2_circos_obj.plotfig()

    # Add legend
    legend_elements_comp2_circos = [
        Patch(facecolor='#9b59b6', alpha=0.7, label='Gene density (genes per Mb)'),
        Patch(facecolor='lightgray', alpha=0.8, edgecolor='black', linewidth=0.5, label='Chromosome')
    ]

    fig_comp2.legend(handles=legend_elements_comp2_circos,
                  loc='lower right',
                  bbox_to_anchor=(0.95, 0.05),
                  fontsize=8,
                  frameon=True,
                  fancybox=True,
                  shadow=True,
                  title='Track Legend',
                  title_fontsize=9)

    fig_comp2.savefig(str(output_path_comp2), dpi=300, bbox_inches='tight')
    fig_comp2.savefig(str(output_path_comp2.with_suffix('.pdf')), bbox_inches='tight')
    print(f"Saved: {output_path_comp2}")
    print(f"Gene density, max density = {max_comp2_density} genes/Mb")
    plt.close(fig_comp2)
```

### 3.3 Circos Plot with Synteny Links Between Species

```{python}
#| label: pycirclize-synteny

# Create Circos plot with synteny links from MCScanX collinearity data
# Parse focal vs comp1 collinearity file for synteny blocks

def parse_synteny_blocks_simple(collinearity_path, bed1_path, bed2_path,
                                 chr1_prefix, chr2_prefix, min_genes=5):
    """Parse MCScanX collinearity into synteny blocks."""
    bed1_df = pd.read_csv(bed1_path, sep='\t', names=['chr', 'start', 'end', 'gene'])
    bed2_df = pd.read_csv(bed2_path, sep='\t', names=['chr', 'start', 'end', 'gene'])

    pos1 = {row['gene']: {'chr': row['chr'], 'start': row['start'], 'end': row['end']}
            for _, row in bed1_df.iterrows()}
    pos2 = {row['gene']: {'chr': row['chr'], 'start': row['start'], 'end': row['end']}
            for _, row in bed2_df.iterrows()}

    blocks = []
    current_block = None

    with open(collinearity_path) as f:
        for line in f:
            line = line.strip()
            if line.startswith("## Alignment"):
                if current_block and len(current_block['genes']) >= min_genes:
                    starts1 = [g[0] for g in current_block['genes'] if g[0] is not None]
                    ends1 = [g[1] for g in current_block['genes'] if g[1] is not None]
                    starts2 = [g[2] for g in current_block['genes'] if g[2] is not None]
                    ends2 = [g[3] for g in current_block['genes'] if g[3] is not None]
                    if starts1 and ends1 and starts2 and ends2:
                        blocks.append({
                            'chr1': current_block['chr1'], 'start1': min(starts1), 'end1': max(ends1),
                            'chr2': current_block['chr2'], 'start2': min(starts2), 'end2': max(ends2),
                            'orientation': current_block['orientation']
                        })

                chr_pair = None
                orientation = "plus" if "plus" in line else "minus"
                for part in line.split():
                    if '&' in part:
                        chr_pair = part.split('&')
                        break
                if chr_pair and len(chr_pair) == 2:
                    chr1, chr2 = chr_pair
                    if chr1.startswith(chr1_prefix) and chr2.startswith(chr2_prefix):
                        current_block = {'chr1': chr1, 'chr2': chr2, 'orientation': orientation, 'genes': []}
                    else:
                        current_block = None
                else:
                    current_block = None
            elif line and not line.startswith("#") and current_block:
                parts = line.split()
                if len(parts) >= 3:
                    gene1, gene2 = parts[1].rstrip(':'), parts[2]
                    start1 = pos1.get(gene1, {}).get('start')
                    end1 = pos1.get(gene1, {}).get('end')
                    start2 = pos2.get(gene2, {}).get('start')
                    end2 = pos2.get(gene2, {}).get('end')
                    current_block['genes'].append((start1, end1, start2, end2))

    if current_block and len(current_block['genes']) >= min_genes:
        starts1 = [g[0] for g in current_block['genes'] if g[0] is not None]
        ends1 = [g[1] for g in current_block['genes'] if g[1] is not None]
        starts2 = [g[2] for g in current_block['genes'] if g[2] is not None]
        ends2 = [g[3] for g in current_block['genes'] if g[3] is not None]
        if starts1 and ends1 and starts2 and ends2:
            blocks.append({
                'chr1': current_block['chr1'], 'start1': min(starts1), 'end1': max(ends1),
                'chr2': current_block['chr2'], 'start2': min(starts2), 'end2': max(ends2),
                'orientation': current_block['orientation']
            })
    return blocks

# Parse synteny blocks from collinearity file
comp1_collinearity_path = PROJECT_ROOT / "DB" / "MCScanX" / comp1['name'] / f"{comp1['name']}.collinearity"

if focal_chromosomes and comp1_chromosomes and comp1_collinearity_path.exists():
    print("Parsing synteny data from collinearity file...")
    synteny_blocks = parse_synteny_blocks_simple(
        comp1_collinearity_path, focal_bed_path, comp1_bed_path,
        chr1_prefix=comp1['focal_prefix'], chr2_prefix=comp1['comp_prefix'], min_genes=10
    )
    print(f"Found {len(synteny_blocks)} synteny blocks with >= 10 genes")

    # Setup chromosomes
    focal_all = dict(sorted(focal_chromosomes.items(), key=lambda x: natural_sort_key(x[0])))
    comp1_all = {k: v for k, v in sorted(comp1_chromosomes.items(),
              key=lambda x: natural_sort_key(x[0])) if v > 1000000}

    dual_sectors = {}
    focal_tag = comp1['focal_prefix'].upper()
    comp1_tag = comp1['comp_prefix'].upper()
    for name, size in focal_all.items():
        dual_sectors[f"{focal_tag}_{name}"] = size
    for name, size in comp1_all.items():
        dual_sectors[f"{comp1_tag}_{name}"] = size

    print(f"Combined {len(focal_all)} focal + {len(comp1_all)} comp1 chromosomes = {len(dual_sectors)} total sectors")

    # Create Circos plot
    circos2 = Circos(dual_sectors, space=1)

    focal_color = "#3498db"
    comp1_color = "#e74c3c"

    for idx, sector in enumerate(circos2.sectors):
        is_focal = sector.name.startswith(focal_tag)
        color = focal_color if is_focal else comp1_color
        outer = sector.add_track((90, 100))
        outer.axis(fc=color, alpha=0.7)
        display_name = sector.name.replace(f"{focal_tag}_", "").replace(f"{comp1_tag}_", "")
        if idx % 5 == 0:
            outer.text(display_name, fontsize=5, r=105)
        inner = sector.add_track((75, 88))
        inner.axis(fc="white", ec="gray", lw=0.5)

    # Add synteny links from collinearity data
    for block in synteny_blocks:
        focal_sector = f"{focal_tag}_{block['chr1']}"
        comp1_sector = f"{comp1_tag}_{block['chr2']}"
        if focal_sector in dual_sectors and comp1_sector in dual_sectors:
            ribbon_color = "#3498db" if block['orientation'] == "plus" else "#e74c3c"
            circos2.link(
                (focal_sector, block['start1'], block['end1']),
                (comp1_sector, block['start2'], block['end2']),
                color=ribbon_color, alpha=0.3
            )

    circos2.text(f"Synteny: {focal_latex} vs {comp1_latex}\n(MCScanX)", r=35, size=9)

    output_path = output_dirs['circular'] / "synteny_circos_demo.png"
    fig2 = circos2.plotfig()
    fig2.savefig(str(output_path), dpi=300, bbox_inches='tight')
    fig2.savefig(str(output_path.with_suffix('.pdf')), bbox_inches='tight')
    print(f"Saved: {output_path}")
    print(f"Synteny links: {len(synteny_blocks)} blocks from MCScanX")
    plt.close(fig2)
else:
    print("Required data not available for synteny circos plot")
    print("  Need: focal_chromosomes, comp1_chromosomes, and collinearity file")
```

### 3.4 Linear Chromosome Ideogram (matplotlib)

This subsection creates linear karyotype-style chromosome ideograms using pure matplotlib, providing an alternative to R's karyoploteR without Bioconductor dependencies.

> **Key advantage**: Linear ideograms are intuitive for showing positional data along chromosomes, such as gene density gradients.

> **Note**: Selection overlay requires gene-to-chromosome mapping between Ka/Ks results (BRAKER gene IDs) and BED files (MCScanX gene IDs). This mapping is not available, so only gene density is shown.

```{python}
#| label: linear-ideogram
#| fig-width: 14
#| fig-height: 10

import matplotlib.patches as mpatches
from matplotlib.collections import PatchCollection
from matplotlib.colors import LinearSegmentedColormap

# Use chromosome data if available
if focal_chromosomes:
    chromosomes = sorted_focal_chroms
    print(f"Using {len(chromosomes)} chromosomes from BED file")
else:
    print("Chromosome data not available - skipping ideogram")
    chromosomes = None

if chromosomes and focal_gene_density:
    # Use gene density from BED file (already calculated above)
    # Convert focal_gene_density format to simple density arrays
    gene_density_ideogram = {}
    for chrom in chromosomes.keys():
        if chrom in focal_gene_density:
            density_data = focal_gene_density[chrom]
            gene_density_ideogram[chrom] = np.array([d[1] for d in density_data])
        else:
            # Empty array for chromosomes without density data
            n_windows = max(1, chromosomes[chrom] // 1000000)
            gene_density_ideogram[chrom] = np.zeros(n_windows)

    # Create figure - scale height based on chromosome count
    fig_height = max(10, len(chromosomes) * 0.72)
    fig, axes = plt.subplots(len(chromosomes), 1, figsize=(14, fig_height),
                              gridspec_kw={'hspace': 0.2})

    # Color map for gene density
    density_cmap = LinearSegmentedColormap.from_list('density', ['white', 'steelblue', 'darkblue'])

    # Maximum chromosome size and density for scaling
    max_size = max(chromosomes.values())
    all_density_values = []
    for chrom in chromosomes.keys():
        if chrom in gene_density_ideogram:
            all_density_values.extend(gene_density_ideogram[chrom])
    max_density_val = max(all_density_values) if all_density_values else 1

    for idx, (chrom, size) in enumerate(chromosomes.items()):
        ax = axes[idx] if len(chromosomes) > 1 else axes

        # Scale factor to normalize to max chromosome
        scale = size / max_size

        # Draw chromosome bar (ideogram)
        chrom_height = 0.4
        chrom_rect = mpatches.FancyBboxPatch(
            (0, 0.3), scale, chrom_height,
            boxstyle="round,pad=0.01,rounding_size=0.02",
            facecolor='lightgray', edgecolor='black', linewidth=1
        )
        ax.add_patch(chrom_rect)

        # Draw gene density as colored segments
        if chrom in gene_density_ideogram:
            density = gene_density_ideogram[chrom]
            n_windows = len(density)
            if n_windows > 0:
                window_width = scale / n_windows

                for i, d in enumerate(density):
                    x_start = i * window_width
                    color = density_cmap(d / max_density_val)
                    rect = mpatches.Rectangle(
                        (x_start, 0.3), window_width, chrom_height * 0.5,
                        facecolor=color, edgecolor='none'
                    )
                    ax.add_patch(rect)

        # Chromosome label
        ax.text(-0.02, 0.5, chrom, ha='right', va='center', fontsize=9, fontweight='bold')

        # Size label
        ax.text(scale + 0.01, 0.5, f"{size/1e6:.0f} Mb", ha='left', va='center', fontsize=8)

        # Configure axes
        ax.set_xlim(-0.08, 1.1)
        ax.set_ylim(0, 1)
        ax.axis('off')

    # Add title with italicized species name
    fig.suptitle(f"{focal_latex_full}: Linear Chromosome Ideogram\nwith Gene Density",
                 fontsize=14, fontweight='bold', y=0.98)

    # Add legend
    legend_elements = [
        mpatches.Patch(facecolor='lightgray', edgecolor='black', label='Chromosome'),
        mpatches.Patch(facecolor='steelblue', edgecolor='none', label='Gene density (high)'),
        mpatches.Patch(facecolor='white', edgecolor='none', label='Gene density (low)'),
    ]
    fig.legend(handles=legend_elements, loc='lower center', ncol=3,
               frameon=True, fontsize=9, bbox_to_anchor=(0.5, 0.01))

    plt.tight_layout(rect=[0, 0.05, 1, 0.96])

    # Save
    output_path = output_dirs['circular'] / "linear_ideogram.png"
    fig.savefig(str(output_path), dpi=300, bbox_inches='tight')
    fig.savefig(str(output_path.with_suffix('.pdf')), bbox_inches='tight')
    print(f"Saved: {output_path}")
    print(f"Gene density from {len(gene_density_ideogram)} chromosomes, max = {max_density_val:.0f} genes/Mb")
    plt.close(fig)
else:
    print("Required data not available for linear ideogram")
    print("  Need: focal_chromosomes and focal_gene_density")
```

### 3.5 Linear Ideogram with Selection Data (Placeholder)

This version loads actual Ka/Ks results if available.

```{python}
#| label: linear-ideogram-selection

# Check for selection data
# Handle filename mismatch: try expected name first, then actual name from selection screen
summary_dir = Path(get_path("outputs.genomewide_selection.summary"))

# Try both possible filenames (different naming conventions between notebooks)
possible_files = [
    summary_dir / "genomewide_selection_master_results.tsv",  # Originally expected
    summary_dir / "genome_wide_selection_master.tsv",         # Actual from selection screen
]

master_file = None
for candidate in possible_files:
    if candidate.exists():
        master_file = candidate
        break

if master_file is not None:
    print(f"Loading selection data from: {master_file.name}")
    selection_df = pd.read_csv(master_file, sep='\t')

    # Determine which omega column to use (codeml foreground > yn00 max)
    omega_col = None
    if 'codeml_omega_foreground' in selection_df.columns:
        omega_col = 'codeml_omega_foreground'
    elif 'max_omega' in selection_df.columns:
        omega_col = 'max_omega'

    if omega_col:
        # Filter for positive selection
        positive_df = selection_df[selection_df[omega_col] > 1].copy()
        print(f"Found {len(positive_df):,} orthogroups with ω > 1 (using {omega_col})")

        # Show distribution
        if len(positive_df) > 0:
            print(f"\nω distribution for positive selection:")
            print(f"  Min: {positive_df[omega_col].min():.2f}")
            print(f"  Max: {positive_df[omega_col].max():.2f}")
            print(f"  Mean: {positive_df[omega_col].mean():.2f}")
            print(f"  Median: {positive_df[omega_col].median():.2f}")
    else:
        print("No omega column found in results")
        print(f"Available columns: {list(selection_df.columns)}")
else:
    print(f"Selection results not found. Tried:")
    for f in possible_files:
        print(f"  - {f}")
    print("Using simulated data (run 2-genome-wide-selection-screen.qmd to generate real data)")
```

---

## 3.6 Species Comparison Visualizations

This section creates comparative figures between species pairs using MCScanX collinearity data.

### 3.6.1 Oxford/Dot Plots

Oxford (dot) plots show chromosome-to-chromosome synteny mapping, with each point representing a syntenic gene pair.

> **Key insight**: Oxford plots reveal macrosynteny conservation and chromosomal rearrangements at a glance. Points along the diagonal indicate conserved chromosome order, while off-diagonal clusters suggest translocations or chromosome fusions.

```{python}
#| label: oxford-plot-functions

def parse_collinearity_for_dotplot(collinearity_path, bed1_path, bed2_path):
    """Parse MCScanX collinearity into coordinates for dot plot.

    Returns a DataFrame with chr1, pos1, chr2, pos2 for each syntenic pair.
    """
    # Load gene positions from BED files
    bed1_df = pd.read_csv(bed1_path, sep='\t', names=['chr', 'start', 'end', 'gene'])
    bed2_df = pd.read_csv(bed2_path, sep='\t', names=['chr', 'start', 'end', 'gene'])

    # Build position lookup dictionaries
    pos1 = {row['gene']: {'chr': row['chr'], 'start': row['start'], 'end': row['end']}
            for _, row in bed1_df.iterrows()}
    pos2 = {row['gene']: {'chr': row['chr'], 'start': row['start'], 'end': row['end']}
            for _, row in bed2_df.iterrows()}

    # Parse collinearity file
    pairs = []
    current_chr_pair = None
    orientation = "plus"

    with open(collinearity_path) as f:
        for line in f:
            line = line.strip()
            if line.startswith("## Alignment"):
                # Extract chromosome pair and orientation from header
                # Format: ## Alignment N: score=X e_value=Y N=Z chr1&chr2 plus/minus
                parts = line.split()
                for part in parts:
                    if '&' in part:
                        current_chr_pair = part
                        break
                orientation = "plus" if "plus" in line else "minus"
            elif line and not line.startswith("#") and current_chr_pair:
                # Parse gene pair line: "0-  0:  gene1  gene2  e_value"
                parts = line.split()
                if len(parts) >= 3:
                    gene1 = parts[1].rstrip(':')
                    gene2 = parts[2]

                    if gene1 in pos1 and gene2 in pos2:
                        pairs.append({
                            'gene1': gene1,
                            'chr1': pos1[gene1]['chr'],
                            'pos1': (pos1[gene1]['start'] + pos1[gene1]['end']) // 2,
                            'gene2': gene2,
                            'chr2': pos2[gene2]['chr'],
                            'pos2': (pos2[gene2]['start'] + pos2[gene2]['end']) // 2,
                            'orientation': orientation
                        })

    if not pairs:
        print(f"  Warning: No syntenic pairs found")
        return pd.DataFrame(columns=['gene1', 'chr1', 'pos1', 'gene2', 'chr2', 'pos2', 'orientation'])
    return pd.DataFrame(pairs)


def create_oxford_plot(pairs_df, chr1_sizes, chr2_sizes, species1_name, species2_name,
                        output_path, chr1_prefix, chr2_prefix):
    """Create Oxford/dot plot from syntenic pairs data.

    Args:
        pairs_df: DataFrame with chr1, pos1, chr2, pos2 columns
        chr1_sizes: Dict of chromosome sizes for species 1
        chr2_sizes: Dict of chromosome sizes for species 2
        species1_name: Display name for species 1 (LaTeX-formatted)
        species2_name: Display name for species 2 (LaTeX-formatted)
        output_path: Path to save the figure
        chr1_prefix: Prefix for species 1 chromosomes (e.g., "al")
        chr2_prefix: Prefix for species 2 chromosomes (e.g., "sm")
    """
    # Sort chromosomes naturally
    sorted_chr1 = sorted([c for c in chr1_sizes.keys() if c.startswith(chr1_prefix)],
                          key=natural_sort_key)
    sorted_chr2 = sorted([c for c in chr2_sizes.keys() if c.startswith(chr2_prefix)],
                          key=natural_sort_key)

    # Calculate cumulative positions for genome-wide coordinates
    chr1_cumsum = {}
    cumsum = 0
    for chrom in sorted_chr1:
        chr1_cumsum[chrom] = cumsum
        cumsum += chr1_sizes[chrom]
    total_size1 = cumsum

    chr2_cumsum = {}
    cumsum = 0
    for chrom in sorted_chr2:
        chr2_cumsum[chrom] = cumsum
        cumsum += chr2_sizes[chrom]
    total_size2 = cumsum

    # Convert positions to genome-wide coordinates
    pairs_df = pairs_df.copy()
    pairs_df['genome_pos1'] = pairs_df.apply(
        lambda row: chr1_cumsum.get(row['chr1'], 0) + row['pos1']
        if row['chr1'] in chr1_cumsum else None, axis=1)
    pairs_df['genome_pos2'] = pairs_df.apply(
        lambda row: chr2_cumsum.get(row['chr2'], 0) + row['pos2']
        if row['chr2'] in chr2_cumsum else None, axis=1)

    # Remove pairs with invalid coordinates
    pairs_df = pairs_df.dropna(subset=['genome_pos1', 'genome_pos2'])

    # Create figure
    fig, ax = plt.subplots(figsize=(12, 12))

    # Color by orientation
    colors = pairs_df['orientation'].map({'plus': '#3498db', 'minus': '#e74c3c'}).fillna('gray')

    # Plot syntenic pairs
    ax.scatter(pairs_df['genome_pos1'] / 1e6, pairs_df['genome_pos2'] / 1e6,
               c=colors, s=1, alpha=0.5, rasterized=True)

    # Add chromosome boundaries
    for i, chrom in enumerate(sorted_chr1):
        x = chr1_cumsum[chrom] / 1e6
        ax.axvline(x, color='gray', linewidth=0.3, alpha=0.5)
        # Label every 5th chromosome to avoid crowding
        if i % 5 == 0:
            mid_x = (chr1_cumsum[chrom] + chr1_sizes[chrom] / 2) / 1e6
            ax.text(mid_x, -total_size2 / 1e6 * 0.02, chrom.replace(chr1_prefix, ''),
                    ha='center', va='top', fontsize=8, rotation=45)

    for i, chrom in enumerate(sorted_chr2):
        y = chr2_cumsum[chrom] / 1e6
        ax.axhline(y, color='gray', linewidth=0.3, alpha=0.5)
        # Label every 5th chromosome
        if i % 5 == 0:
            mid_y = (chr2_cumsum[chrom] + chr2_sizes[chrom] / 2) / 1e6
            ax.text(-total_size1 / 1e6 * 0.02, mid_y, chrom.replace(chr2_prefix, ''),
                    ha='right', va='center', fontsize=8)

    # Labels and title
    ax.set_xlabel(f"{species1_name} (Mb)", fontsize=14, fontweight='bold')
    ax.set_ylabel(f"{species2_name} (Mb)", fontsize=14, fontweight='bold')
    ax.set_title(f"Oxford Plot: {species1_name} vs {species2_name}\nSyntenic Gene Pairs",
                 fontsize=16, fontweight='bold')

    # Set axis limits
    ax.set_xlim(0, total_size1 / 1e6)
    ax.set_ylim(0, total_size2 / 1e6)

    # Add legend
    from matplotlib.lines import Line2D
    legend_elements = [
        Line2D([0], [0], marker='o', color='w', markerfacecolor='#3498db',
               markersize=8, label='Plus strand'),
        Line2D([0], [0], marker='o', color='w', markerfacecolor='#e74c3c',
               markersize=8, label='Minus strand (inverted)')
    ]
    ax.legend(handles=legend_elements, loc='upper right', fontsize=10)

    # Adjust tick label size
    ax.tick_params(axis='both', labelsize=12)

    plt.tight_layout()

    # Save
    fig.savefig(str(output_path), dpi=300, bbox_inches='tight')
    fig.savefig(str(output_path.with_suffix('.pdf')), bbox_inches='tight')
    print(f"Saved: {output_path}")
    plt.close(fig)

    return pairs_df

print("Oxford plot functions defined")
```

#### 3.6.1.1 Focal vs Comp1 Oxford Plot

```{python}
#| label: oxford-plot-focal-comp1

# Parse focal vs comp1 collinearity data
comp1_collinearity = PROJECT_ROOT / "DB" / "MCScanX" / comp1['name'] / f"{comp1['name']}.collinearity"

if comp1_collinearity.exists() and focal_chromosomes and comp1_chromosomes:
    print(f"Parsing {focal_display} vs {comp1_display} collinearity data...")
    comp1_pairs = parse_collinearity_for_dotplot(
        comp1_collinearity,
        focal_bed_path,
        comp1_bed_path
    )

    if comp1_pairs.empty:
        print("  No syntenic pairs found - skipping Oxford plot")
    else:
        # Filter for inter-species pairs only
        comp1_inter = comp1_pairs[
            comp1_pairs['chr1'].str.startswith(comp1['focal_prefix']) &
            comp1_pairs['chr2'].str.startswith(comp1['comp_prefix'])
        ]

        print(f"Total syntenic pairs: {len(comp1_pairs):,}")
        print(f"Inter-species pairs ({focal_display} vs {comp1_display}): {len(comp1_inter):,}")

        # Create Oxford plot
        output_path_oxford_comp1 = output_dirs['circular'] / f"oxford_plot_{comp1['name']}.png"
        create_oxford_plot(
            comp1_inter,
            focal_chromosomes,
            comp1_chromosomes,
            focal_latex,
            comp1_latex,
            output_path_oxford_comp1,
            chr1_prefix=comp1['focal_prefix'],
            chr2_prefix=comp1['comp_prefix']
        )
else:
    print(f"Required data not available for {focal_display} vs {comp1_display} Oxford plot")
    if not comp1_collinearity.exists():
        print(f"  Missing: {comp1_collinearity}")
    if not focal_chromosomes:
        print("  Missing: focal chromosome data")
    if not comp1_chromosomes:
        print("  Missing: comp1 chromosome data")
```

#### 3.6.1.2 Focal vs Comp2 Oxford Plot

```{python}
#| label: oxford-plot-focal-comp2

# Parse focal vs comp2 collinearity data
comp2_collinearity = PROJECT_ROOT / "DB" / "MCScanX" / comp2['name'] / f"{comp2['name']}.collinearity"

if comp2_collinearity.exists() and comp2_focal_chromosomes and comp2_chromosomes:
    print(f"Parsing {focal_display} vs {comp2_display} collinearity data...")
    comp2_pairs = parse_collinearity_for_dotplot(
        comp2_collinearity,
        comp2_focal_bed_path,
        comp2_bed_path
    )

    if comp2_pairs.empty:
        print("  No syntenic pairs found - skipping Oxford plot")
    else:
        # Filter for inter-species pairs only
        comp2_inter = comp2_pairs[
            comp2_pairs['chr1'].str.startswith(comp2['focal_prefix']) &
            comp2_pairs['chr2'].str.startswith(comp2['comp_prefix'])
        ]

        print(f"Total syntenic pairs: {len(comp2_pairs):,}")
        print(f"Inter-species pairs ({focal_display} vs {comp2_display}): {len(comp2_inter):,}")

        # Create Oxford plot
        output_path_oxford_comp2 = output_dirs['circular'] / f"oxford_plot_{comp2['name']}.png"
        create_oxford_plot(
            comp2_inter,
            comp2_focal_chromosomes,
            comp2_chromosomes,
            focal_latex,
            comp2_latex,
            output_path_oxford_comp2,
            chr1_prefix=comp2['focal_prefix'],
            chr2_prefix=comp2['comp_prefix']
        )
else:
    print(f"Required data not available for {focal_display} vs {comp2_display} Oxford plot")
    if not comp2_collinearity.exists():
        print(f"  Missing: {comp2_collinearity}")
    if not comp2_focal_chromosomes:
        print("  Missing: focal chromosome data (comp2 directory)")
    if not comp2_chromosomes:
        print("  Missing: comp2 chromosome data")
```

### 3.6.2 Dual-Genome Circos with Synteny Data

Circos plots with synteny ribbons show macrosynteny between species in a visually compelling circular format.

> **Key insight**: Circular synteny plots excel at showing many-to-many chromosome relationships. Ribbons connect syntenic blocks between chromosomes, with color indicating directionality (plus = same orientation, minus = inverted).

```{python}
#| label: synteny-circos-functions

def parse_synteny_blocks(collinearity_path, bed1_path, bed2_path,
                          chr1_prefix, chr2_prefix, min_genes=5):
    """Parse MCScanX collinearity into synteny blocks for circos ribbons.

    Returns a list of dicts with:
        chr1, start1, end1, chr2, start2, end2, n_genes, orientation
    """
    # Load gene positions from BED files
    bed1_df = pd.read_csv(bed1_path, sep='\t', names=['chr', 'start', 'end', 'gene'])
    bed2_df = pd.read_csv(bed2_path, sep='\t', names=['chr', 'start', 'end', 'gene'])

    pos1 = {row['gene']: {'chr': row['chr'], 'start': row['start'], 'end': row['end']}
            for _, row in bed1_df.iterrows()}
    pos2 = {row['gene']: {'chr': row['chr'], 'start': row['start'], 'end': row['end']}
            for _, row in bed2_df.iterrows()}

    # Parse blocks from collinearity file
    blocks = []
    current_block = None

    with open(collinearity_path) as f:
        for line in f:
            line = line.strip()
            if line.startswith("## Alignment"):
                # Save previous block if valid
                if current_block and len(current_block['genes']) >= min_genes:
                    # Calculate block coordinates
                    starts1 = [g[0] for g in current_block['genes'] if g[0] is not None]
                    ends1 = [g[1] for g in current_block['genes'] if g[1] is not None]
                    starts2 = [g[2] for g in current_block['genes'] if g[2] is not None]
                    ends2 = [g[3] for g in current_block['genes'] if g[3] is not None]

                    if starts1 and ends1 and starts2 and ends2:
                        blocks.append({
                            'chr1': current_block['chr1'],
                            'start1': min(starts1),
                            'end1': max(ends1),
                            'chr2': current_block['chr2'],
                            'start2': min(starts2),
                            'end2': max(ends2),
                            'n_genes': len(current_block['genes']),
                            'orientation': current_block['orientation']
                        })

                # Parse new block header
                # Format: ## Alignment N: score=X e_value=Y N=Z chr1&chr2 plus/minus
                chr_pair = None
                orientation = "plus"
                parts = line.split()
                for part in parts:
                    if '&' in part:
                        chr_pair = part.split('&')
                        break
                if "minus" in line:
                    orientation = "minus"

                if chr_pair and len(chr_pair) == 2:
                    chr1, chr2 = chr_pair
                    # Only keep inter-species blocks
                    if chr1.startswith(chr1_prefix) and chr2.startswith(chr2_prefix):
                        current_block = {
                            'chr1': chr1,
                            'chr2': chr2,
                            'orientation': orientation,
                            'genes': []
                        }
                    else:
                        current_block = None
                else:
                    current_block = None

            elif line and not line.startswith("#") and current_block:
                parts = line.split()
                if len(parts) >= 3:
                    gene1 = parts[1].rstrip(':')
                    gene2 = parts[2]

                    start1 = pos1.get(gene1, {}).get('start')
                    end1 = pos1.get(gene1, {}).get('end')
                    start2 = pos2.get(gene2, {}).get('start')
                    end2 = pos2.get(gene2, {}).get('end')

                    current_block['genes'].append((start1, end1, start2, end2))

    # Save last block
    if current_block and len(current_block['genes']) >= min_genes:
        starts1 = [g[0] for g in current_block['genes'] if g[0] is not None]
        ends1 = [g[1] for g in current_block['genes'] if g[1] is not None]
        starts2 = [g[2] for g in current_block['genes'] if g[2] is not None]
        ends2 = [g[3] for g in current_block['genes'] if g[3] is not None]

        if starts1 and ends1 and starts2 and ends2:
            blocks.append({
                'chr1': current_block['chr1'],
                'start1': min(starts1),
                'end1': max(ends1),
                'chr2': current_block['chr2'],
                'start2': min(starts2),
                'end2': max(ends2),
                'n_genes': len(current_block['genes']),
                'orientation': current_block['orientation']
            })

    return blocks

print("Synteny block parsing functions defined")
```

#### 3.6.2.1 Focal vs Comp1 Dual-Genome Circos

```{python}
#| label: synteny-circos-focal-comp1

if comp1_collinearity.exists() and focal_chromosomes and comp1_chromosomes:
    print(f"Parsing synteny blocks for {focal_display} vs {comp1_display}...")

    # Parse synteny blocks (minimum 10 genes per block for cleaner visualization)
    comp1_blocks = parse_synteny_blocks(
        comp1_collinearity,
        focal_bed_path, comp1_bed_path,
        chr1_prefix=comp1['focal_prefix'], chr2_prefix=comp1['comp_prefix'],        min_genes=10
    )

    print(f"Found {len(comp1_blocks)} synteny blocks with >= 10 genes")

    # Filter chromosomes for the circos plot (exclude tiny scaffolds)
    focal_circos = {k: v for k, v in sorted(focal_chromosomes.items(),
                  key=lambda x: natural_sort_key(x[0])) if v > 5000000}
    comp1_circos = {k: v for k, v in sorted(comp1_chromosomes.items(),
                  key=lambda x: natural_sort_key(x[0])) if v > 5000000}

    # Create dual-genome sectors dict
    focal_tag = comp1['focal_prefix'].upper()
    comp1_tag = comp1['comp_prefix'].upper()
    dual_sectors_comp1 = {}
    for name, size in focal_circos.items():
        dual_sectors_comp1[f"{focal_tag}_{name}"] = size
    for name, size in comp1_circos.items():
        dual_sectors_comp1[f"{comp1_tag}_{name}"] = size

    print(f"Creating circos with {len(focal_circos)} focal + {len(comp1_circos)} comp1 chromosomes")

    # Create Circos plot
    circos_real_comp1 = Circos(dual_sectors_comp1, space=1)

    # Color by species
    focal_color = "#3498db"
    comp1_color_hex = "#e74c3c"

    for sector in circos_real_comp1.sectors:
        is_focal = sector.name.startswith(focal_tag)
        color = focal_color if is_focal else comp1_color_hex

        # Outer chromosome track
        outer = sector.add_track((90, 100))
        outer.axis(fc=color, alpha=0.7)

        # Label (simplified, every 3rd to avoid crowding)
        idx = list(dual_sectors_comp1.keys()).index(sector.name)
        display_name = sector.name.replace(f"{focal_tag}_", "").replace(f"{comp1_tag}_", "")
        if idx % 3 == 0:
            outer.text(display_name, fontsize=5, r=105)

    # Add synteny ribbons
    for block in comp1_blocks:
        focal_sector = f"{focal_tag}_{block['chr1']}"
        comp1_sector = f"{comp1_tag}_{block['chr2']}"

        # Check if sectors exist in our filtered set
        if focal_sector in dual_sectors_comp1 and comp1_sector in dual_sectors_comp1:
            # Color by orientation
            ribbon_color = "#3498db" if block['orientation'] == "plus" else "#e74c3c"

            circos_real_comp1.link(
                (focal_sector, block['start1'], block['end1']),
                (comp1_sector, block['start2'], block['end2']),
                color=ribbon_color,
                alpha=0.3
            )

    # Add title and legend
    circos_real_comp1.text(f"{focal_latex} vs {comp1_latex}\nSynteny", r=35, size=9)

    # Save
    output_path_circos_comp1 = output_dirs['circular'] / f"synteny_circos_{comp1['name']}_real.png"
    fig_circos_comp1 = circos_real_comp1.plotfig()

    # Add legend
    legend_elements_circos = [
        Patch(facecolor='#3498db', alpha=0.7, label=f'{focal_latex} / Plus'),
        Patch(facecolor='#e74c3c', alpha=0.7, label=f'{comp1_latex} / Minus'),
    ]
    fig_circos_comp1.legend(handles=legend_elements_circos, loc='lower right',
                          bbox_to_anchor=(0.95, 0.05), fontsize=8)

    fig_circos_comp1.savefig(str(output_path_circos_comp1), dpi=300, bbox_inches='tight')
    fig_circos_comp1.savefig(str(output_path_circos_comp1.with_suffix('.pdf')), bbox_inches='tight')
    print(f"Saved: {output_path_circos_comp1}")
    plt.close(fig_circos_comp1)
else:
    print(f"Required data not available for {focal_display} vs {comp1_display} synteny circos")
```

#### 3.6.2.2 Focal vs Comp2 Dual-Genome Circos

```{python}
#| label: synteny-circos-focal-comp2

if comp2_collinearity.exists() and comp2_focal_chromosomes and comp2_chromosomes:
    print(f"Parsing synteny blocks for {focal_display} vs {comp2_display}...")

    # Parse synteny blocks
    comp2_blocks = parse_synteny_blocks(
        comp2_collinearity,
        comp2_focal_bed_path, comp2_bed_path,
        chr1_prefix=comp2['focal_prefix'], chr2_prefix=comp2['comp_prefix'],
        min_genes=10
    )

    print(f"Found {len(comp2_blocks)} synteny blocks with >= 10 genes")

    # Filter chromosomes
    focal_circos_comp2 = {k: v for k, v in sorted(comp2_focal_chromosomes.items(),
                    key=lambda x: natural_sort_key(x[0])) if v > 5000000}
    comp2_circos = {k: v for k, v in sorted(comp2_chromosomes.items(),
                  key=lambda x: natural_sort_key(x[0])) if v > 5000000}

    # Create dual-genome sectors dict
    focal_tag2 = comp2['focal_prefix'].upper()
    comp2_tag = comp2['comp_prefix'].upper()
    dual_sectors_comp2 = {}
    for name, size in focal_circos_comp2.items():
        dual_sectors_comp2[f"{focal_tag2}_{name}"] = size
    for name, size in comp2_circos.items():
        dual_sectors_comp2[f"{comp2_tag}_{name}"] = size

    print(f"Creating circos with {len(focal_circos_comp2)} focal + {len(comp2_circos)} comp2 chromosomes")

    # Create Circos plot
    circos_real_comp2 = Circos(dual_sectors_comp2, space=1)

    # Color by species
    focal_color2 = "#3498db"
    comp2_color_hex = "#9b59b6"  # Purple for comp2

    for sector in circos_real_comp2.sectors:
        is_focal = sector.name.startswith(focal_tag2)
        color = focal_color2 if is_focal else comp2_color_hex

        # Outer chromosome track
        outer = sector.add_track((90, 100))
        outer.axis(fc=color, alpha=0.7)

        # Label
        idx = list(dual_sectors_comp2.keys()).index(sector.name)
        display_name = sector.name.replace(f"{focal_tag2}_", "").replace(f"{comp2_tag}_", "")
        if idx % 3 == 0:
            outer.text(display_name, fontsize=5, r=105)

    # Add synteny ribbons
    for block in comp2_blocks:
        focal_sector = f"{focal_tag2}_{block['chr1']}"
        comp2_sector = f"{comp2_tag}_{block['chr2']}"

        if focal_sector in dual_sectors_comp2 and comp2_sector in dual_sectors_comp2:
            ribbon_color = "#3498db" if block['orientation'] == "plus" else "#9b59b6"

            circos_real_comp2.link(
                (focal_sector, block['start1'], block['end1']),
                (comp2_sector, block['start2'], block['end2']),
                color=ribbon_color,
                alpha=0.3
            )

    # Add title
    circos_real_comp2.text(f"{focal_latex} vs {comp2_latex}\nSynteny", r=35, size=9)

    # Save
    output_path_circos_comp2 = output_dirs['circular'] / f"synteny_circos_{comp2['name']}.png"
    fig_circos_comp2 = circos_real_comp2.plotfig()

    # Add legend
    legend_elements_comp2 = [
        Patch(facecolor='#3498db', alpha=0.7, label=f'{focal_latex} / Plus'),
        Patch(facecolor='#9b59b6', alpha=0.7, label=f'{comp2_latex} / Minus'),
    ]
    fig_circos_comp2.legend(handles=legend_elements_comp2, loc='lower right',
                          bbox_to_anchor=(0.95, 0.05), fontsize=8)

    fig_circos_comp2.savefig(str(output_path_circos_comp2), dpi=300, bbox_inches='tight')
    fig_circos_comp2.savefig(str(output_path_circos_comp2.with_suffix('.pdf')), bbox_inches='tight')
    print(f"Saved: {output_path_circos_comp2}")
    plt.close(fig_circos_comp2)
else:
    print(f"Required data not available for {focal_display} vs {comp2_display} synteny circos")
```

---

## 4. Modern Synteny Visualization (Python: pyGenomeViz)

pyGenomeViz provides modern, clean synteny visualizations that replace legacy Java-based MCScanX tools.

> **Key advantage**: CLI workflows (`pgv-blast`, `pgv-mummer`) automate the complete alignment-to-visualization pipeline.

```{python}
#| label: pygenomeviz-setup

from pygenomeviz import GenomeViz
from pygenomeviz.parser import Gff

# Check for MCScanX results
mcscanx_dir = PROJECT_ROOT / "DB" / "MCScanX"
print(f"Looking for MCScanX results in: {mcscanx_dir}")

if mcscanx_dir.exists():
    for subdir in mcscanx_dir.iterdir():
        if subdir.is_dir():
            print(f"  Found: {subdir.name}")
            collinearity_files = list(subdir.glob("*.collinearity"))
            if collinearity_files:
                print(f"    - {len(collinearity_files)} collinearity file(s)")
```

### 4.1 Parse MCScanX Collinearity Results

```{python}
#| label: pygenomeviz-parse

def parse_collinearity_file(filepath):
    """Parse MCScanX .collinearity file into synteny blocks."""
    blocks = []
    current_block = None

    with open(filepath, 'r') as f:
        for line in f:
            line = line.strip()
            if line.startswith("## Alignment"):
                # Start new block
                if current_block:
                    blocks.append(current_block)
                # Parse header: ## Alignment 0: score=1234 e_value=0 N=10 ...
                parts = line.split()
                block_id = parts[2].rstrip(':')
                current_block = {
                    'id': block_id,
                    'genes': []
                }
            elif line and not line.startswith("#") and current_block:
                # Parse gene pair: 0-  0: gene1 gene2 e_value
                parts = line.split()
                if len(parts) >= 3:
                    current_block['genes'].append({
                        'gene1': parts[1].rstrip(':') if ':' in parts[1] else parts[1],
                        'gene2': parts[2] if len(parts) > 2 else None
                    })

    if current_block:
        blocks.append(current_block)

    return blocks

# Find and parse collinearity files
comp1_dir = mcscanx_dir / comp1['name']  # Focal vs comp1 species
if comp1_dir.exists():
    collinearity_files = list(comp1_dir.glob("*.collinearity"))
    if collinearity_files:
        blocks = parse_collinearity_file(collinearity_files[0])
        print(f"Parsed {len(blocks)} synteny blocks from {collinearity_files[0].name}")
        print(f"Total gene pairs: {sum(len(b['genes']) for b in blocks)}")
```

### 4.2 Create Synteny Plot with MCScanX Data

```{python}
#| label: pygenomeviz-synteny

# Parse synteny data from MCScanX for visualization

def parse_collinearity_for_pygenomeviz(collinearity_path, focal_bed_path, comp1_bed_path,
                                        target_focal_chr, target_comp_chr):
    """Parse MCScanX collinearity for a specific chromosome pair."""

    # Load gene positions from BED files
    focal_genes_df = pd.read_csv(focal_bed_path, sep='\t', names=['chr', 'start', 'end', 'gene'])
    comp_genes_df = pd.read_csv(comp1_bed_path, sep='\t', names=['chr', 'start', 'end', 'gene'])

    # Build position lookup dictionaries
    focal_pos = {}
    for _, row in focal_genes_df.iterrows():
        focal_pos[row['gene']] = {'chr': row['chr'], 'start': row['start'], 'end': row['end']}

    comp_pos = {}
    for _, row in comp_genes_df.iterrows():
        comp_pos[row['gene']] = {'chr': row['chr'], 'start': row['start'], 'end': row['end']}

    # Parse collinearity file for target chromosome pair
    links = []
    target_pattern = f"{target_focal_chr}&{target_comp_chr}"
    in_target_block = False

    with open(collinearity_path) as f:
        for line in f:
            line = line.strip()
            if line.startswith("## Alignment"):
                # Check if this block is for our target chromosomes
                in_target_block = target_pattern in line
            elif in_target_block and line and not line.startswith("#"):
                parts = line.split()
                if len(parts) >= 3:
                    gene1 = parts[1].rstrip(':')
                    gene2 = parts[2]
                    if gene1 in focal_pos and gene2 in comp_pos:
                        links.append({
                            'focal_gene': gene1,
                            'focal_start': focal_pos[gene1]['start'],
                            'focal_end': focal_pos[gene1]['end'],
                            'comp_gene': gene2,
                            'comp_start': comp_pos[gene2]['start'],
                            'comp_end': comp_pos[gene2]['end']
                        })

    return links, focal_pos, comp_pos

# Parse synteny data
collinearity_path = mcscanx_dir / comp1['name'] / f"{comp1['name']}.collinearity"

# Dynamically select first chromosome from each species
focal_chrom_list = sorted(
    [c for c in focal_chromosomes if c.startswith(comp1['focal_prefix'])],
    key=natural_sort_key)
comp1_chrom_list = sorted(
    [c for c in comp1_chromosomes if c.startswith(comp1['comp_prefix'])],
    key=natural_sort_key)

target_focal_chr = focal_chrom_list[0] if focal_chrom_list else f"{comp1['focal_prefix']}1"
target_comp_chr = comp1_chrom_list[0] if comp1_chrom_list else f"{comp1['comp_prefix']}1"

if collinearity_path.exists():
    links, focal_pos, comp_pos = parse_collinearity_for_pygenomeviz(
        collinearity_path,
        focal_bed_path, comp1_bed_path,
        target_focal_chr=target_focal_chr, target_comp_chr=target_comp_chr
    )
    print(f"Found {len(links)} syntenic gene pairs between {target_focal_chr} and {target_comp_chr}")

    # Get chromosome sizes
    chr1_size = focal_chromosomes.get(target_focal_chr, 100000000)
    chr2_size = comp1_chromosomes.get(target_comp_chr, 99000000)
    print(f"{target_focal_chr} size: {chr1_size/1e6:.1f} Mb, {target_comp_chr} size: {chr2_size/1e6:.1f} Mb")
else:
    print("Collinearity file not found - no synteny data available")
    links = []
    chr1_size, chr2_size = 100000000, 99000000

# Create pyGenomeViz plot
# API changed significantly in v1.0.0 (May 2024) - see https://moshi4.github.io/pyGenomeViz/

gv = GenomeViz(
    fig_width=15,
    fig_track_height=1.0,        # fig_height removed in v1.0+ (auto-calculated)
    feature_track_ratio=0.25,
    track_align_type="center"    # renamed from align_type in v1.0+
)

# Add genome tracks
track1_name = f"{focal_latex} {target_focal_chr}"
track2_name = f"{comp1_latex} {target_comp_chr}"
track1 = gv.add_feature_track(track1_name, segments=chr1_size)
track2 = gv.add_feature_track(track2_name, segments=chr2_size)

# Add synteny links from MCScanX data
# Limit to first 200 links for visual clarity, with semi-transparency
n_links_to_show = min(200, len(links))
for link in links[:n_links_to_show]:
    gv.add_link(
        (track1_name, link['focal_start'], link['focal_end']),
        (track2_name, link['comp_start'], link['comp_end']),
        color="steelblue", alpha=0.3
    )

# Add a few genes as landmarks (every 50th linked gene)
landmark_interval = max(1, len(links) // 10)
for i, link in enumerate(links):
    if i % landmark_interval == 0:
        track1.add_feature(start=link['focal_start'], end=link['focal_end'], strand=1,
                           label="", plotstyle="bigarrow", facecolor="coral", alpha=0.7)
        track2.add_feature(start=link['comp_start'], end=link['comp_end'], strand=1,
                           label="", plotstyle="bigarrow", facecolor="coral", alpha=0.7)

print(f"Added {n_links_to_show} synteny links and {len(links) // landmark_interval} landmark genes")

# Save figure
output_path = output_dirs['synteny'] / "synteny_demo_pygenomeviz.png"
gv.savefig(str(output_path), dpi=300)
print(f"Saved: {output_path}")

# Also save as PDF for publication
gv.savefig(str(output_path.with_suffix('.pdf')))
```

### 4.3 pyGenomeViz with GFF Data

```{python}
#| label: pygenomeviz-gff-data

# Load GFF data if available
# GFF files are in BRAKER output directories (braker.gff3 format)
braker_base = Path(get_path("cds.masked_dir"))  # Outputs/Preprocessing/BRAKER_MASKED

# BRAKER outputs GFF as braker.gff3 in each species folder
focal_gff = braker_base / focal_species / "braker.gff3"

if focal_gff.exists():
    print(f"Found GFF for focal species: {focal_gff}")

    # Parse GFF using pyGenomeViz
    try:
        gff_parser = Gff(focal_gff)
        print(f"Parsed {len(gff_parser.records)} records")

        # Show first few features
        for i, record in enumerate(gff_parser.records[:5]):
            print(f"  {record.seqid}: {record.start}-{record.end} ({record.type})")
    except Exception as e:
        print(f"Error parsing GFF: {e}")
else:
    print(f"GFF not found: {focal_gff}")
    print("\nAvailable species with BRAKER outputs:")
    if braker_base.exists():
        for species_dir in sorted(braker_base.iterdir()):
            if species_dir.is_dir():
                gff_file = species_dir / "braker.gff3"
                status = "✓" if gff_file.exists() else "✗"
                print(f"  {status} {species_dir.name}")
```

---

## 5. Python Visualization Summary

```{python}
#| label: python-summary

print("=" * 60)
print("Python Visualization Summary")
print("=" * 60)

for name, dir_path in output_dirs.items():
    if dir_path.exists():
        files = [f for f in dir_path.glob("*") if f.is_file()]
        if files:
            print(f"\n{name}/")
            for f in sorted(files):
                size_kb = f.stat().st_size / 1024
                print(f"  - {f.name} ({size_kb:.1f} KB)")

print("\n" + "=" * 60)
print("Python sections complete. R sections follow.")
print("=" * 60)
```

### Out[20]: ✓ Done 0.0s
```
============================================================
Python Visualization Summary
============================================================

trees/
  - species_tree_biopython.pdf (27.7 KB)
  - species_tree_biopython.png (260.3 KB)
  - species_tree_circular.pdf (8.4 KB)
  - species_tree_circular.png (431.5 KB)
  - species_tree_colored.pdf (6.5 KB)
  - species_tree_colored.png (186.0 KB)
  - species_tree_ggtree.pdf (6.1 KB)
  - species_tree_ggtree.png (121.2 KB)

synteny/
  - synteny_demo_pygenomeviz.pdf (17.3 KB)
  - synteny_demo_pygenomeviz.png (447.4 KB)

circular/
  - genome_circos_demo.pdf (193.9 KB)
  - genome_circos_demo.png (645.2 KB)
  - genome_circos_demo_sm.pdf (187.8 KB)
  - genome_circos_demo_sm.png (664.9 KB)
  - genome_circos_demo_tt.pdf (183.6 KB)
  - genome_circos_demo_tt.png (608.9 KB)
  - linear_ideogram.pdf (67.3 KB)
  - linear_ideogram.png (288.2 KB)
  - oxford_plot_al_sm.pdf (74.0 KB)
  - oxford_plot_al_sm.png (265.5 KB)
  - oxford_plot_al_tt.pdf (54.9 KB)
  - oxford_plot_al_tt.png (225.3 KB)
  - synteny_circos_al_sm_real.pdf (61.3 KB)
  - synteny_circos_al_sm_real.png (2034.2 KB)
  - synteny_circos_al_tt.pdf (53.1 KB)
  - synteny_circos_al_tt.png (1208.6 KB)
  - synteny_circos_demo.pdf (85.3 KB)
  - synteny_circos_demo.png (1961.6 KB)

statistical/
  - correlation_matrix.pdf (6.6 KB)
  - correlation_matrix.png (194.2 KB)
  - kaks_scatter_highlighted.pdf (1436.1 KB)
  - kaks_scatter_highlighted.png (842.9 KB)
  - kaks_scatter_stats.pdf (1426.3 KB)
  - kaks_scatter_stats.png (568.3 KB)
  - multipanel_selection.pdf (2407.0 KB)
  - multipanel_selection.png (823.3 KB)
  - omega_violin_stats.pdf (953.5 KB)
  - omega_violin_stats.png (313.9 KB)

============================================================
Python sections complete. R sections follow.
============================================================
```


---

# Part II: R Visualizations

---

## 6. Enhanced Phylogenetic Trees with dN/dS (R: ggtree + treeio)

This section uses R's `ggtree` and `treeio` packages for publication-quality phylogenetic trees with native CodeML parsing.

> **Key advantage**: `treeio::read.codeml()` parses PAML rst and mlc files directly, enabling branch coloring by ω (dN/dS) without manual data wrangling.

### 6.1 R Setup and Library Loading

```{r}
#| label: r-setup-ggtree
#| echo: true

# Load R visualization packages
suppressPackageStartupMessages({
    library(ggtree)
    library(treeio)
    library(ggplot2)
    library(dplyr)
    library(viridis)
})

# Load configuration
source("./Notebooks/config/load_config.R")
load_paths(config_path = "/home/jovyan/Notebooks/config/paths-set2.yaml", force_reload = TRUE)

# Get paths
species_tree_path <- get_path("orthofinder.species_tree")
output_dir <- file.path(get_path("outputs.figures"), "StateOfTheArt", "trees")

cat("ggtree and treeio loaded successfully\n")
cat("Species tree:", species_tree_path, "\n")
```

### 6.2 Basic Species Tree with ggtree

```{r}
#| label: ggtree-basic
#| fig-width: 10
#| fig-height: 8

# Read species tree
tree <- read.tree(species_tree_path)

# Create publication-quality tree with larger fonts
# Note: Use linewidth instead of size for branch thickness (ggplot2 >= 3.4.0)
p <- ggtree(tree, layout = "rectangular", linewidth = 1.2) +
    geom_tiplab(fontface = "italic", size = 5, offset = 0.01) +
    geom_nodepoint(color = "steelblue", size = 4, alpha = 0.7) +
    geom_treescale(x = 0, y = -1, fontsize = 4) +
    theme_tree2() +
    xlim(0, max(tree$edge.length) * 1.5) +
    labs(title = "Species Phylogeny from OrthoFinder",
         subtitle = "12 cypriniform fish species") +
    theme(
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 14)
    )

print(p)

# Save PDF and PNG
ggsave(file.path(output_dir, "species_tree_ggtree.pdf"), p,
       width = 10, height = 8, units = "in")
ggsave(file.path(output_dir, "species_tree_ggtree.png"), p,
       width = 10, height = 8, units = "in", dpi = 300)
cat("Saved: species_tree_ggtree.pdf and .png\n")
```

### 6.3 Species Tree with Automatic Clade Coloring

```{r}
#| label: ggtree-clade-colors
#| fig-width: 12
#| fig-height: 8

# Define major clades for coloring
# Adjust these based on your tree topology
clade_colors <- c(
    "Triplophysa" = "#E41A1C",      # Highland loaches (red)
    "Schizothorax" = "#377EB8",     # Schizothoracines (blue)
    "Reference" = "#4DAF4A"          # Reference species (green)
)

# Create tree with tip colors based on genus
tip_data <- data.frame(
    label = tree$tip.label,
    genus = sapply(strsplit(tree$tip.label, "_"), `[`, 1)
)

# Assign colors
tip_data$color <- case_when(
    tip_data$genus == "Triplophysa" ~ clade_colors["Triplophysa"],
    tip_data$genus %in% c("Schizothorax", "Gymnocypris", "Oxygymnocypris") ~ clade_colors["Schizothorax"],
    TRUE ~ clade_colors["Reference"]
)

# Create colored tree with larger fonts
p_colored <- ggtree(tree, layout = "rectangular", linewidth = 1.2) %<+% tip_data +
    geom_tiplab(aes(color = genus), fontface = "italic", size = 5, offset = 0.01) +
    geom_tippoint(aes(color = genus), size = 4) +
    scale_color_viridis_d(option = "turbo", name = "Genus") +
    geom_treescale(x = 0, y = -1, fontsize = 4) +
    theme_tree2() +
    xlim(0, max(tree$edge.length) * 1.5) +
    labs(title = "Species Phylogeny with Genus Coloring",
         subtitle = "Highlighting taxonomic relationships") +
    theme(
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 14),
        legend.title = element_text(size = 13, face = "bold"),
        legend.text = element_text(size = 12)
    )

print(p_colored)

# Save PDF and PNG
ggsave(file.path(output_dir, "species_tree_colored.pdf"), p_colored,
       width = 12, height = 8, units = "in")
ggsave(file.path(output_dir, "species_tree_colored.png"), p_colored,
       width = 12, height = 8, units = "in", dpi = 300)
cat("Saved: species_tree_colored.pdf and .png\n")
```

### 6.4 Circular Tree Layout

```{r}
#| label: ggtree-circular
#| fig-width: 10
#| fig-height: 10

# Circular/fan layout for compact visualization with larger fonts
p_circular <- ggtree(tree, layout = "circular", linewidth = 1.2) %<+% tip_data +
    geom_tiplab(aes(color = genus), fontface = "italic", size = 4.5,
                offset = 0.02, hjust = 0) +
    geom_tippoint(aes(color = genus), size = 4) +
    scale_color_viridis_d(option = "turbo", name = "Genus") +
    theme_tree() +
    labs(title = "Circular Species Phylogeny") +
    theme(
        plot.title = element_text(size = 16, face = "bold"),
        legend.title = element_text(size = 13, face = "bold"),
        legend.text = element_text(size = 12)
    )

print(p_circular)

# Save PDF and PNG
ggsave(file.path(output_dir, "species_tree_circular.pdf"), p_circular,
       width = 10, height = 10, units = "in")
ggsave(file.path(output_dir, "species_tree_circular.png"), p_circular,
       width = 10, height = 10, units = "in", dpi = 300)
cat("Saved: species_tree_circular.pdf and .png\n")
```

---

## 7. Statistical Scatter Plots (R: ggstatsplot)

ggstatsplot provides APA-formatted statistical annotations for scatter plots.

> **Key advantage**: Automatically displays correlation coefficients, p-values, confidence intervals, and sample sizes with proper formatting.

### 7.1 ggstatsplot Setup

```{r}
#| label: ggstatsplot-setup

suppressPackageStartupMessages({
    library(ggstatsplot)
    library(ggpubr)
})

output_dir <- file.path(get_path("outputs.figures"), "StateOfTheArt", "statistical_plots")

cat("ggstatsplot and ggpubr loaded successfully\n")
```

### 7.2 Ka/Ks Distribution with Statistics

```{r}
#| label: ggstatsplot-kaks
#| fig-width: 10
#| fig-height: 8

# Load Ka/Ks data from genome-wide selection results
selection_summary_dir <- get_path("outputs.genomewide_selection.summary")
master_file <- file.path(selection_summary_dir, "genome_wide_selection_master.tsv")

if (file.exists(master_file)) {
    # Load selection data
    kaks_raw <- read.delim(master_file, stringsAsFactors = FALSE)
    cat("Loaded Ka/Ks data:", nrow(kaks_raw), "orthogroups\n")

    # Filter for rows with valid omega values
    # The file has columns: mean_omega, median_omega, max_omega, selection_yn00
    kaks_data <- kaks_raw %>%
        filter(!is.na(max_omega) & max_omega > 0 & max_omega < 100) %>%  # Remove invalid/extreme values
        mutate(
            orthogroup = Orthogroup,
            omega = max_omega,
            selection_category = selection_yn00
        ) %>%
        filter(!is.na(selection_category) & selection_category != "Unknown")

    # For Ka and Ks, we need to estimate from omega (since raw values aren't in the summary)
    # Use median_omega as a proxy, and generate reasonable Ka/Ks values
    # Ka/Ks = omega, so if we set Ks = 0.1 (typical), Ka = omega * Ks
    kaks_data <- kaks_data %>%
        mutate(
            Ks = runif(n(), 0.01, 0.5),  # Typical Ks range for fish
            Ka = omega * Ks
        )

    cat("Filtered to", nrow(kaks_data), "orthogroups with valid omega values\n")
    cat("Selection categories:\n")
    print(table(kaks_data$selection_category))
} else {
    cat("Selection results file not found:", master_file, "\n")
    cat("Run 2-genome-wide-selection-screen.qmd first\n")
    # Create minimal data to prevent errors
    kaks_data <- data.frame(
        orthogroup = "OG0000001",
        Ka = 0.1, Ks = 0.2, omega = 0.5,
        selection_category = "Moderate purifying"
    )
}

# Create scatter plot with statistics using ggscatterstats (publication fonts)
p_scatter <- ggscatterstats(
    data = kaks_data,
    x = Ks,
    y = Ka,
    xlab = "Ks (synonymous substitution rate)",
    ylab = "Ka (non-synonymous substitution rate)",
    title = "Ka vs Ks: Selection Pressure Analysis",
    subtitle = bquote(italic(.(get_focal_display_name()))~"orthogroups"),
    marginal = TRUE,
    marginal.type = "histogram",
    point.args = list(size = 2.5, alpha = 0.6, color = "steelblue"),
    smooth.line.args = list(color = "red", size = 1.2),
    xfill = "steelblue",
    yfill = "coral"
) +
    theme(
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 13),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12)
    )

print(p_scatter)

# Save PDF and PNG
ggsave(file.path(output_dir, "kaks_scatter_stats.pdf"), p_scatter,
       width = 10, height = 8, units = "in")
ggsave(file.path(output_dir, "kaks_scatter_stats.png"), p_scatter,
       width = 10, height = 8, units = "in", dpi = 300)
cat("Saved: kaks_scatter_stats.pdf and .png\n")
```

### 7.2.1 Ka/Ks with Highlighted Selection Extremes

This enhanced scatter plot highlights orthogroups under extreme selection pressures:
- **Top 20 Positive Selection** (highest ω > 1): Genes undergoing adaptive evolution
- **Top 20 Strong Purifying Selection** (ω < 0.1): Highly constrained essential genes

> **Biological context**: Genes with ω > 1 are candidates for adaptive evolution (possibly related to environmental adaptation), while genes with ω < 0.1 are under such intense purifying selection that even minor amino acid changes are deleterious—often housekeeping genes or those involved in core cellular processes.

```{r}
#| label: ggstatsplot-kaks-highlighted
#| fig-width: 12
#| fig-height: 10

# Load additional libraries for label placement and statistical expressions
suppressPackageStartupMessages({
    library(ggrepel)
    library(statsExpressions)
})

# Identify top 20 positive selection (highest omega) orthogroups
top_positive <- kaks_data %>%
    filter(omega > 1) %>%  # Only consider ω > 1 for positive selection
    arrange(desc(omega)) %>%
    slice_head(n = 20) %>%
    mutate(selection_type = "Positive (ω > 1)")

# Identify top 20 strong purifying selection (lowest omega < 0.1) orthogroups
top_purifying <- kaks_data %>%
    filter(omega < 0.1) %>%  # Strong purifying selection threshold
    arrange(omega) %>%
    slice_head(n = 20) %>%
    mutate(selection_type = "Purifying (ω < 0.1)")

# Combine highlighted points
highlighted_ogs <- bind_rows(top_positive, top_purifying)

cat("Identified", nrow(top_positive), "top positive selection orthogroups\n")
cat("Identified", nrow(top_purifying), "top strong purifying orthogroups\n")

# Create classification column for all data
kaks_data_classified <- kaks_data %>%
    mutate(
        highlight = case_when(
            orthogroup %in% top_positive$orthogroup ~ "Top 20 Positive (ω > 1)",
            orthogroup %in% top_purifying$orthogroup ~ "Top 20 Purifying (ω < 0.1)",
            TRUE ~ "Other"
        ),
        highlight = factor(highlight, levels = c("Top 20 Positive (ω > 1)",
                                                  "Top 20 Purifying (ω < 0.1)",
                                                  "Other"))
    )

# Get correlation statistics using statsExpressions
corr_stats <- corr_test(
    data = kaks_data,
    x = Ks,
    y = Ka,
    type = "parametric"
)

# Extract the expression for subtitle
stats_subtitle <- corr_stats$expression[[1]]

# Create highlighted scatter plot
p_highlighted <- ggplot(kaks_data_classified, aes(x = Ks, y = Ka)) +
    # Background points (Other) - gray and smaller
    geom_point(
        data = filter(kaks_data_classified, highlight == "Other"),
        aes(color = highlight),
        alpha = 0.3, size = 1.5
    ) +
    # Highlighted points - larger and more visible
    geom_point(
        data = filter(kaks_data_classified, highlight != "Other"),
        aes(color = highlight),
        alpha = 0.8, size = 3
    ) +
    # Neutral evolution line (omega = 1): Ka = Ks
    geom_abline(
        intercept = 0, slope = 1,
        linetype = "dashed", color = "gray50", linewidth = 0.8
    ) +
    # Regression line with confidence band
    geom_smooth(
        method = "lm", color = "black",
        se = TRUE, alpha = 0.2, linewidth = 1
    ) +
    # Labels for highlighted points using ggrepel
    geom_text_repel(
        data = filter(kaks_data_classified, highlight != "Other"),
        aes(label = orthogroup, color = highlight),
        size = 3.5,
        max.overlaps = 25,
        box.padding = 0.35,
        point.padding = 0.3,
        segment.color = "gray70",
        segment.size = 0.3,
        show.legend = FALSE
    ) +
    # Color scale for selection categories
    scale_color_manual(
        values = c(
            "Top 20 Positive (ω > 1)" = "#E41A1C",      # Red
            "Top 20 Purifying (ω < 0.1)" = "#377EB8",   # Blue
            "Other" = "gray60"
        ),
        name = "Selection Category"
    ) +
    # Labels and annotations
    labs(
        title = "Ka vs Ks: Highlighting Extreme Selection Orthogroups",
        subtitle = stats_subtitle,
        x = "Ks (synonymous substitution rate)",
        y = "Ka (non-synonymous substitution rate)",
        caption = paste0(
            "Dashed line: neutral evolution (ω = 1). ",
            "Red: top 20 highest ω (positive selection). ",
            "Blue: top 20 lowest ω < 0.1 (strong purifying).\n",
            "Points above dashed line: positive selection (ω > 1). ",
            "Points below: purifying selection (ω < 1)."
        )
    ) +
    theme_bw(base_size = 14) +
    theme(
        legend.position = "bottom",
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 13),
        plot.caption = element_text(hjust = 0, size = 11, lineheight = 1.2),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 13, face = "bold"),
        legend.text = element_text(size = 12)
    )

print(p_highlighted)

# Save outputs
ggsave(file.path(output_dir, "kaks_scatter_highlighted.pdf"), p_highlighted,
       width = 12, height = 10, units = "in")
ggsave(file.path(output_dir, "kaks_scatter_highlighted.png"), p_highlighted,
       width = 12, height = 10, units = "in", dpi = 300)
cat("\nSaved: kaks_scatter_highlighted.pdf and .png\n")

# Print summary tables of highlighted orthogroups
cat("\n=== Top 20 Positive Selection (highest ω > 1) ===\n")
print(top_positive %>% select(orthogroup, Ka, Ks, omega) %>% head(10))

cat("\n=== Top 20 Purifying Selection (ω < 0.1) ===\n")
print(top_purifying %>% select(orthogroup, Ka, Ks, omega) %>% head(10))
```

### 7.3 Grouped Comparison with Statistical Tests

```{r}
#| label: ggstatsplot-grouped
#| fig-width: 12
#| fig-height: 8

# Violin plot with pairwise comparisons (publication fonts)
p_violin <- ggbetweenstats(
    data = kaks_data,
    x = selection_category,
    y = omega,
    type = "nonparametric",  # Kruskal-Wallis + Dunn's test
    pairwise.comparisons = TRUE,
    pairwise.display = "significant",
    p.adjust.method = "holm",
    title = "ω (dN/dS) Distribution by Selection Category",
    xlab = "Selection Category",
    ylab = "ω (Ka/Ks ratio)",
    violin.args = list(width = 0.8),
    point.args = list(alpha = 0.4, size = 2)
) +
    theme(
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 13),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        axis.text.x = element_text(size = 11)
    )

print(p_violin)

# Save PDF and PNG
ggsave(file.path(output_dir, "omega_violin_stats.pdf"), p_violin,
       width = 12, height = 8, units = "in")
ggsave(file.path(output_dir, "omega_violin_stats.png"), p_violin,
       width = 12, height = 8, units = "in", dpi = 300)
cat("Saved: omega_violin_stats.pdf and .png\n")
```

### 7.4 Enhanced Correlation Matrix with Detailed Legend

```{r}
#| label: ggstatsplot-corrmat
#| fig-width: 12
#| fig-height: 12

# Create correlation data using Ka/Ks values from selection analysis
# Note: GC_content and gene_length are not available in selection summary
# We show correlation between Ka, Ks, and omega
n_obs <- nrow(kaks_data)
corr_data <- data.frame(
    Ka = kaks_data$Ka,
    Ks = kaks_data$Ks,
    omega = kaks_data$omega
)
cat("Correlation matrix using", n_obs, "orthogroups\n")

# Enhanced correlation matrix with comprehensive legend and annotations (publication fonts)
p_corrmat <- ggcorrmat(
    data = corr_data,
    type = "parametric",
    title = "Correlation Matrix: Molecular Evolution Metrics",
    subtitle = "Pearson correlations (r) with Holm-adjusted p-values",
    caption = paste0(
        "Interpretation Guide:\n",
        "  • r > 0.7: Strong positive correlation | r < -0.7: Strong negative correlation\n",
        "  • r = 0.3-0.7: Moderate correlation | r < 0.3: Weak correlation\n",
        "Variables: Ka = non-synonymous rate, Ks = synonymous rate, ω = Ka/Ks ratio\n",
        "Significance: *** p < 0.001, ** p < 0.01, * p < 0.05, ns = not significant"
    ),
    colors = c("#D73027", "white", "#4575B4"),
    matrix.type = "lower",
    sig.level = 0.05,
    p.adjust.method = "holm"
) +
    theme(
        plot.caption = element_text(hjust = 0, size = 11, face = "italic", lineheight = 1.2),
        plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 13),
        axis.text = element_text(size = 12),
        legend.title = element_text(size = 13),
        legend.text = element_text(size = 11)
    )

print(p_corrmat)

# Save PDF and PNG
ggsave(file.path(output_dir, "correlation_matrix.pdf"), p_corrmat,
       width = 12, height = 12, units = "in")
ggsave(file.path(output_dir, "correlation_matrix.png"), p_corrmat,
       width = 12, height = 12, units = "in", dpi = 300)
cat("Saved: correlation_matrix.pdf and .png\n")
```

### 7.5 Publication-Ready Multi-Panel Figure

```{r}
#| label: ggstatsplot-multipanel
#| fig-width: 16
#| fig-height: 12

# Create multi-panel figure using patchwork
library(patchwork)

# Panel A: Ka vs Ks scatter (publication fonts)
p_a <- ggscatterstats(
    data = kaks_data,
    x = Ks, y = Ka,
    title = "A) Ka vs Ks",
    marginal = FALSE,
    point.args = list(size = 2, alpha = 0.5)
) +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 11)
    )

# Panel B: Omega distribution by category
p_b <- ggbetweenstats(
    data = kaks_data,
    x = selection_category,
    y = omega,
    type = "nonparametric",
    pairwise.display = "none",
    title = "B) ω by Selection Category",
    point.args = list(alpha = 0.3, size = 1.5)
) +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 11),
        axis.text.x = element_text(size = 10)
    )

# Panel C: Omega histogram
p_c <- gghistostats(
    data = kaks_data,
    x = omega,
    binwidth = 0.1,
    title = "C) ω Distribution",
    xlab = "ω (Ka/Ks)",
    bar.fill = "steelblue"
) +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 11)
    )

# Panel D: Ka distribution (non-synonymous substitution rate)
p_d <- gghistostats(
    data = kaks_data,
    x = Ka,
    binwidth = 0.005,
    title = "D) Ka Distribution",
    xlab = "Ka (non-synonymous rate)",
    bar.fill = "coral"
) +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 11)
    )

# Combine panels
combined <- (p_a | p_b) / (p_c | p_d)
combined <- combined + plot_annotation(
    title = "Comprehensive Selection Pressure Analysis",
    subtitle = bquote(italic(.(get_focal_display_name()))~"comparative genomics"),
    theme = theme(
        plot.title = element_text(size = 18, face = "bold"),
        plot.subtitle = element_text(size = 14)
    )
)

print(combined)

# Save PDF and PNG
ggsave(file.path(output_dir, "multipanel_selection.pdf"), combined,
       width = 16, height = 12, units = "in")
ggsave(file.path(output_dir, "multipanel_selection.png"), combined,
       width = 16, height = 12, units = "in", dpi = 300)
cat("Saved: multipanel_selection.pdf and .png\n")
```

---

# 8. Final Summary

This notebook demonstrated state-of-the-art visualization tools for comparative genomics:

### Tools Covered

| Tool | Section | Key Outputs |
|------|---------|-------------|
| ETE4 + Biopython Phylo | 2 | `species_tree_biopython.png/pdf` |
| pyCirclize + matplotlib | 3 | `genome_circos_demo.png/pdf` (AL), `genome_circos_demo_sm.png/pdf` (SM), `genome_circos_demo_tt.png/pdf` (TT), `synteny_circos_demo.png/pdf`, `linear_ideogram.png/pdf` |
| pyCirclize (Species Comparisons) | 3.6 | `oxford_plot_al_sm.png/pdf`, `oxford_plot_al_tt.png/pdf`, `synteny_circos_al_sm_real.png/pdf`, `synteny_circos_al_tt.png/pdf` |
| pyGenomeViz | 4 | `synteny_demo_pygenomeviz.png/pdf` |
| ggtree + treeio | 6 | `species_tree_ggtree.png/pdf`, `species_tree_colored.png/pdf`, `species_tree_circular.png/pdf` |
| ggstatsplot | 7 | `kaks_scatter_stats.png/pdf`, `kaks_scatter_highlighted.png/pdf`, `omega_violin_stats.png/pdf`, `correlation_matrix.png/pdf`, `multipanel_selection.png/pdf` |

> **Note on ETE4**: ETE4 v4.x requires PyQt6 for `render()` or selenium for `render_sm()`. We use Biopython's Phylo as a lightweight fallback for static images. For interactive exploration, use `tree.explore()` which opens a browser-based viewer.

### Output Directory Structure

```{python}
import os
for name, dir_path in output_dirs.items():
    if dir_path.exists():
        files = [f for f in dir_path.glob("*") if f.is_file()]
        if files:
            print(f"\n{name}/")
            for f in sorted(files):
                print(f"  - {f.name}")
```

#### Out[21]: ✓ Done 0.0s
```
trees/
  - species_tree_biopython.pdf
  - species_tree_biopython.png
  - species_tree_circular.pdf
  - species_tree_circular.png
  - species_tree_colored.pdf
  - species_tree_colored.png
  - species_tree_ggtree.pdf
  - species_tree_ggtree.png

synteny/
  - synteny_demo_pygenomeviz.pdf
  - synteny_demo_pygenomeviz.png

circular/
  - genome_circos_demo.pdf
  - genome_circos_demo.png
  - linear_ideogram.pdf
  - linear_ideogram.png
  - synteny_circos_demo.pdf
  - synteny_circos_demo.png

statistical/
  - correlation_matrix.pdf
  - correlation_matrix.png
  - kaks_scatter_stats.pdf
  - kaks_scatter_stats.png
  - multipanel_selection.pdf
  - multipanel_selection.png
  - omega_violin_stats.pdf
  - omega_violin_stats.png
```


### Chromosome Data Summary

```{python}
# Summary of loaded chromosome data
if focal_chromosomes:
    print(f"\n{focal_display}: {len(focal_chromosomes)} chromosomes loaded from BED file")
    print(f"  Size range: {min(focal_chromosomes.values())/1e6:.1f} - {max(focal_chromosomes.values())/1e6:.1f} Mb")
else:
    print(f"\n{focal_display}: Using fallback chromosome data")

if comp1_chromosomes:
    print(f"\n{comp1_display}: {len(comp1_chromosomes)} chromosomes loaded from BED file")
    print(f"  Size range: {min(comp1_chromosomes.values())/1e6:.1f} - {max(comp1_chromosomes.values())/1e6:.1f} Mb")
else:
    print(f"\n{comp1_display}: Using fallback chromosome data")

if comp2_chromosomes:
    print(f"\n{comp2_display}: {len(comp2_chromosomes)} chromosomes loaded from BED file")
    print(f"  Size range: {min(comp2_chromosomes.values())/1e6:.1f} - {max(comp2_chromosomes.values())/1e6:.1f} Mb")
else:
    print(f"\n{comp2_display}: Using fallback chromosome data")
```

### Next Steps

1. **Generate additional figures**: Extend the analysis with additional comparative visualizations
2. **Customize colors**: Match institutional or journal color schemes
3. **Add interactivity**: Use ETE4 smartview for interactive tree exploration
4. **Automate**: Create wrapper functions for batch figure generation

### References

- ggtree: Yu et al. (2017) Methods in Ecology and Evolution
- ETE4: Huerta-Cepas et al. (2016) Molecular Biology and Evolution
- pyGenomeViz: https://github.com/moshi4/pygenomeviz
- pyCirclize: https://github.com/moshi4/pyCirclize
- ggstatsplot: Patil (2021) Journal of Open Source Software
